---
title: "Body size - growth rates analysis"
author: "Amelia Ritger"
date: "2024-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlme) #lme(), nlme()
library(knitr) #kable()
```

# Load Tidied Data
```{r}
source("mergeAllData.R", local = knitr::knit_global())

body_sizeALL <- read_csv(here("experiment", "data", "body_size_all.csv"))
#Load temperature data AFTER you have run source(ReadRPiData)
weekly_temp <- read_csv(here("experiment", "data", "weekly_temperature.csv")) %>%
  select(-treatment)
```

## Create tidied df
```{r}
all_size <- all %>%
  filter(!is.na(avg_size)) %>%
  mutate(avg_size_log = log(avg_size)) %>%
  group_by(tank, genet) %>%
  mutate(avg_size_diff = avg_size_log - lag(avg_size_log),
         avg_size_diff2 = avg_size_diff - lag(avg_size_diff),
         genet_tank = paste(tank, genet, sep = ""),
         tank = as.numeric(tank),
         date_num = as.numeric(as.Date(date)-min(as.Date(date)))) %>%
  ungroup() %>%
  left_join(weekly_temp, by = c("date" = "friday", "tank")) %>%
  select(date, date_num, tank, treatment, mhw, genet, genet_tank, avg_size, sd_size, avg_size_log, avg_biomass_g, sd_biomass_g, avg_size_diff, avg_size_diff2, avg_temp)
```

```{r}
ggplot(all_size, aes(x=date)) +
  geom_smooth(aes(y=avg_size, color=genet))
  
  
ggplot(all_size, aes(x=date)) +
  geom_smooth(aes(y=log(avg_size), color=genet))
```


## Define residual diagnostic plot function
```{r}
# Define a function to plot residual diagnostics for a model
residual_diagnostics <- function(model) {
  # Extract residuals and fitted values
  residuals <- resid(model)
  fitted_values <- fitted(model)
  
  # 1. Residuals vs Fitted Plot (using ggplot2)
  ggplot(data.frame(fitted = fitted_values, residuals = residuals), aes(x = fitted, y = residuals)) +
    geom_point() +
    geom_smooth(method = "loess", color = "red") +  # smoothed line to detect trends
    labs(x = "Fitted values", y = "Residuals") +
    theme_minimal() +
    ggtitle("Residuals vs Fitted Plot")
  
  # 2. Set up the plot layout for two plots side by side (Base R plot)
  par(mfrow = c(1, 2)) # Layout for two plots side by side

  # 2a. Residuals vs Fitted Plot (Base R)
  plot(fitted_values, residuals, 
       xlab = "Fitted Values", 
       ylab = "Residuals", 
       main = "Residuals vs Fitted",
       pch = 16, col = "gray")
  abline(h = 0, col = "red", lty = 2) # Add a horizontal line at 0
  
  # 2b. Spread-Level Plot (Scale-Location Plot)
  plot(fitted_values, sqrt(abs(residuals)), 
       xlab = "Fitted Values", 
       ylab = "Sqrt(|Residuals|)", 
       main = "Spread-Level Plot",
       pch = 16, col = "gray")

  # 3. ACF and PACF of Residuals (Base R)
  acf(residuals, main = "ACF of Residuals")
  pacf(residuals, main = "PACF of Residuals")
  
  par(mfrow = c(1, 1)) # Layout for two plots side by side
  # 4. Q-Q plot of Residuals (Base R)
  qqnorm(residuals)
  qqline(residuals, col = "red")  # Add a red line to the Q-Q plot
}
```

## Jump right into linear mixed model with coAR/coARMA - keep it consistent!
```{r}
ggplot(all_size, aes(x = date_num, y = avg_size_diff, color = genet_tank)) +
  geom_point() +
  geom_line() +
  labs(title = "Body size over time", x = "Date", y = "Avg. size (log)") +
  theme_minimal()

size_diff <- all_size %>%
  filter(!is.na(avg_size_diff))

model_ar1 <- lme(avg_size_log ~ date_num + treatment * genet,
                   random = ~ date_num|genet_tank,
                   data = all_size,
                   correlation = corAR1(form = ~ 1 | genet_tank),
                   control = lmeControl(maxIter = 100, msMaxIter = 100))

model_arma1 <- lme(avg_size_diff ~  treatment * genet,
                   random = ~ 1|genet_tank,
                   data = size_diff,
                   correlation = corARMA(p = 1, q = 0, form = ~ 1 | genet_tank),
                   control = lmeControl(maxIter = 100, msMaxIter = 100))

AIC(model_ar1, model_arma1)
BIC(model_ar1, model_arma1)

residual_diagnostics(model_ar1)
residual_diagnostics(model_arma1)

summary(model_ar1)
summary(model_arma1)

plot(residuals(model_ar1))  # Plot residuals over time
acf(residuals(model_ar1))   # ACF of residuals
pacf(residuals(model_ar1))  # PACF of residuals

ranef(model_ar1)

ranefs_ar1 <- ranef(model_ar1)
ranefs_ar2 <- ranef(model_ar2)

# Plot random intercepts for each genet_tank
ggplot(ranefs_ar1, aes(x = `(Intercept)`, y = rownames(ranefs_ar1))) +
  geom_point() +
  labs(title = "Random Intercepts by genet_tank", x = "Random Intercept", y = "genet_tank") +
  theme_minimal()
ggplot(ranefs_ar2, aes(x = `(Intercept)`, y = rownames(ranefs_ar2))) +
  geom_point() +
  labs(title = "Random Intercepts by genet_tank", x = "Random Intercept", y = "genet_tank") +
  theme_minimal()

# Plot intercept vs slope
ggplot(ranefs_ar1, aes(x = `(Intercept)`, y = date_num)) +
  geom_point() +
  labs(title = "Random Intercepts vs Random Slopes", x = "Random Intercept", y = "Random Slope (date_num)") +
  theme_minimal()
```

Non linear
```{r}
initial_size <- all_size %>%
  group_by(tank, genet) %>%
  filter(date == min(date)) %>%
  select(tank, genet, initial_size = avg_size)

all_size_init <- all_size %>%
  left_join(initial_size, by = c("tank", "genet"))

nlme_model_mixed <- nlme(avg_size ~ initial_size * exp(-decay_rate * date_num), 
                        data = all_size_init, 
                        fixed = initial_size + decay_rate ~ 1,
                        random = initial_size ~ 1 | genet_tank,
                        groups = ~ genet,
                        start = list(fixed = c(initial_size = max(all_size_init$avg_size), decay_rate = 0.1)))

nlme_model_mixed_treatment <- nlme(avg_size ~ initial_size * exp(-decay_rate * date_num) + treatment,
                                  data = all_size_init,
                                  fixed = initial_size + decay_rate + treatment ~ 1,
                                  random = initial_size ~ 1 | genet_tank, 
                                  groups = ~ genet,
                                  start = list(fixed = c(initial_size = max(all_size_init$avg_size), decay_rate = 0.1, treatment = 0)))

#anova(nlme_model_mixed_treatment, nlme_model_mixed_treatment)
residual_diagnostics(nlme_model_mixed_treatment)
residual_diagnostics(nlme_model_mixed_treatment)

plot(residuals(nlme_model_mixed_treatment))
plot(residuals(nlme_model_mixed_treatment))
plot(fitted(nlme_model_mixed_treatment), residuals(nlme_model_mixed_treatment))
plot(fitted(nlme_model_mixed_treatment), residuals(nlme_model_mixed_treatment))

# Predict values from the model
all_size_init$predicted_size <- predict(nlme_model_mixed_treatment, level=0)

ggplot(all_size_init, aes(x = date_num, y = avg_size)) +
  geom_point(aes(color = genet), alpha = 0.6) +  # Plot original data points
  geom_smooth(aes(color = genet)) +
  geom_line(aes(y = predicted_size), color = "blue", size = 1) +  # Plot model predictions, grouped by genet
  geom_smooth(method = "loess", color = "red", se = FALSE) +  # Add LOESS curve for comparison
  labs(x = "Date", y = "Average Size (mm)", color = "Model") +
  theme_bw()  # Customize color scale if needed
```
Exponential decay curve is right fit for my data

```{r}
summary(nlme_model_mixed_treatment)
```

Key Results from the Output:
1. Fixed Effects:

    initial_size:
        The estimate is 3.311795, which suggests that the initial size is a significant factor in the model (i.e., how large the object starts off influences the rate of decay).
        The t-value of 11.01537 and p-value = 0 indicate that this parameter is highly significant, meaning that initial size is an important predictor of avg_size.
    decay_rate:
        The estimate for the decay rate is 0.018269, which suggests that the rate at which the size decays is also significant.
        The t-value of 23.95625 and p-value = 0 show that decay rate is highly significant, indicating that decay is a meaningful component in explaining changes in avg_size.
    treatment:
        The estimate for treatment is 3.566029, which suggests that the treatment effect increases average size by about 3.57 units on average.
        The t-value of 65.03473 and p-value = 0 indicate that treatment has a very strong and highly significant effect on avg_size.

2. Random Effects:

    Initial size:
        The standard deviation of the random effects for initial_size is 0.6611108, which suggests variability between groups (e.g., genetic groups or tanks) in terms of initial size.
    Residual:
        The standard deviation of the residuals (the unexplained variation) is 0.4431341, which gives an indication of the error or unexplained variability after accounting for the fixed and random effects.

3. Correlation:

    There is some correlation between decay_rate and treatment (-0.137), but it's relatively weak. This suggests that while treatment and decay rate may have some relationship, the impact of one on the other is not very strong.
    The decay rate and initial size have a high positive correlation (0.923), which indicates that the decay rate is closely tied to the initial size, suggesting that larger initial sizes may be associated with a slower decay.

4. Model Diagnostics (Residuals):

    The standardized residuals show values ranging from -2.72 to 3.56, which is typical for a model like this. Ideally, you want the residuals to be evenly distributed around zero, and in this case, they appear to be reasonably centered, indicating no major issues with the model fit.

    If you plot the residuals vs. fitted values or do a QQ plot of the residuals, you would want to see a roughly uniform spread around zero and a straight line in the QQ plot. This would confirm that the model assumptions are met (i.e., the residuals are randomly distributed).

5. Number of Observations and Groups:

    Number of Observations: 1425 data points, which is a solid sample size for fitting a model.
    Number of Groups: 5 groups (likely representing genet), which is a relatively small number of groups but may still be sufficient, especially since you're including random effects.
    
    
    All fixed effects (initial size, decay rate, and treatment) are highly significant (p-value = 0), suggesting they are all contributing meaningfully to the model.
The decay rate and initial size have a very strong relationship, which is logical given the nature of decay models. Larger initial sizes tend to decay slower.
Treatment has a significant impact, and this aligns with the treatment being a fixed effect in the model.

The random effect for initial_size indicates some variability between genet_tank combinations. This is expected in many biological models where variability exists between groups (in your case, genet_tank). The magnitude of this variance (0.661) suggests that there is notable group-level variation in the initial size.


#####################


Effect of Treatment: If you’re interested in how different treatments affect the decay rate (rather than just adding treatment as a fixed effect), you could look at interactions between treatment and decay_rate.


Interaction Between Treatment and Groups: If the treatment might affect groups differently (e.g., treatments having a stronger or weaker effect depending on the genet), you could explore adding interactions between treatment and genet.

# Create plot for WSN
```{r}
all_size_plot <- all_size %>%
  #filter(mhw == "during") %>%
  group_by(date, treatment, genet) %>%
  summarize(mean = mean(avg_size),
            sd = sd(avg_size),
            se = sd(avg_size)/sqrt(length(avg_size)),
            ci = qt(0.95, length(avg_size)-1)*se,
            lower = mean - ci,
            upper = mean + ci) %>%
  dplyr::select(date, treatment, genet, mean, sd, lower, upper)

ggplot(all_size_plot, aes(x=date,color=genet, y=mean)) +
  geom_vline(xintercept = as.numeric(as.Date("2023-12-12")), linetype = "dashed", color = "black", alpha = 0.8) +
  geom_point(data = all_size, aes(x=date, y=avg_size, color=genet), alpha=0.3) +
  geom_smooth(aes(y=mean), method = "loess", span=1.2, se=FALSE) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Control", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_x_date(date_labels = "%b %d", 
                   breaks= seq(min(all_size_plot$date), max(all_size_plot$date), length=7)) +
  labs(x = "Date",
       y = "Average size (mm)",
       color = "Genet") +
  guides(fill = "none",
         color=guide_legend(override.aes=list(fill=NA))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 20),
        legend.position = "top")

#ggsave(here("experiment", "figures", "WSN", "size_genet.png"), width=12, height=7)

ggplot(all_size_plot, aes(x=date,color=treatment, y=mean)) +
  geom_vline(xintercept = as.numeric(as.Date("2023-12-12")), linetype = "dashed", color = "black", alpha = 0.8) +
  geom_point(data = all_size, aes(x=date, y=avg_size, color=treatment), alpha=0.3) +
  geom_smooth(aes(y=mean), method = "loess", span=1.2, se=FALSE) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +  
  scale_fill_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  #facet_wrap(~genet) +
             #labeller =  labeller(treatment = c("cold" = "Control", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_x_date(date_labels = "%b %d", 
                   breaks= seq(min(all_size_plot$date), max(all_size_plot$date), length=7)) +
  labs(x = "Date",
       y = "Average size (mm)",
       color = "Treatment") +
  guides(fill = "none",
         color=guide_legend(override.aes=list(fill=NA))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 20),
        legend.position = "top")

#ggsave(here("experiment", "figures", "WSN", "size_treatment.png"), width=12, height=7)
```

## Calculate rate of decay (for nice presentation)
```{r}
fit_results <- all_size %>%
  group_by(treatment, genet) %>%
  do({
    # Fit the exponential decay model to each subset of data
    nls_fit <- tryCatch({
      nls(avg_size ~ y0 * exp(-lambda * date_num),
          data = .,
          start = list(y0 = first(.$avg_size), lambda = 0.1))
    }, error = function(e) NULL)  # Handle errors (e.g., non-convergence)

    # Return model fit result (if successful)
    if (!is.null(nls_fit)) {
      tibble(
        treatment = unique(.$treatment),
        genet = unique(.$genet),
        lambda = coef(nls_fit)['lambda'],
        y0 = coef(nls_fit)['y0'],
        model = list(nls_fit)
      )
    }
  })

# View the results
fit_results

fit_results_sum <- fit_results %>%
  select(treatment, genet, lambda) %>%
  ungroup() %>%
  mutate(lamba = round(lambda, 4))
```

