---
title: "Analyzing behavioral data"
author: "Amelia Ritger"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstatix) #shapiro_test()
library(knitr) #kable()
```

# Import data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
daily_temp <- read_csv(here("experiment", "data", "three_day_temperature.csv"))
```

# Create clean df
```{r}
behavior_all <- all %>%
  mutate(genet_tank = as.factor(paste(tank, genet, sep = ""))) %>%
  dplyr::select(date, tank, genet, genet_tank, treatment, mhw, n_true, percent_open, percent_fully_open, percent_closed)

temp_clean <- daily_temp %>%
  mutate(date = if_else(date == ymd("2023-09-29"), ymd("2023-10-01"), date),
         tank = as.factor(tank)) %>%
  select(-treatment)

behavior_clean <- behavior_all %>%
  left_join(temp_clean, by = c("date", "tank")) %>%
  mutate(date_num = as.numeric(difftime(date, ymd("2023-10-01"), units = "days")))
```

# Initial plot to visualize trends
```{r}
ggplot(behavior_clean, aes(x = genet, y = percent_fully_open, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9), alpha = 0.5) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.3, color = alpha("black", 0.75)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5), labels = c("During MHW", "After MHW")) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17), labels = c("During MHW", "After MHW")) + # Define shapes for mhw levels
  # change color of violin 
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  labs(x = "Genet",
       y = "% open polyps",
       color = "Genet",
       shape = "MHW",
       size = "MHW") +
  guides(fill = "none",
         color = "none") +
  theme_bw() +
  theme(text = element_text(size = 20))

#ggsave(here("experiment", "figures", "WSN", "behavior.png"), width=12, height=7)
#ggsave(file = here("experiment", "figures", "WSN","behavior.svg"), plot = image, width=12, height=7)
```

## Check ANOVA assumptions
```{r}
# Fit your ANOVA model
model1 <- aov(percent_fully_open ~ mhw * treatment * genet, data = behavior_clean)
model2 <- lmer(percent_fully_open ~ mhw * treatment * genet + (1 | genet_tank), data = behavior_clean)

# Extract residuals
residuals <- residuals(model)

# Q-Q plot
qqnorm(residuals)
qqline(residuals, col = "red")

# Plot residuals vs. fitted values
plot(model, which = 1)

# Boxplot of residuals to check for outliers
boxplot(residuals ~ fitted(model))
```

# Define residual diagnostic plot function
```{r}
# Define a function to plot residual diagnostics for a model
residual_diagnostics <- function(model) {
  # Extract residuals and fitted values
  residuals <- resid(model)
  fitted_values <- fitted(model)
  
  # 1. Residuals vs Fitted Plot (using ggplot2)
  ggplot(data.frame(fitted = fitted_values, residuals = residuals), aes(x = fitted, y = residuals)) +
    geom_point() +
    geom_smooth(method = "loess", color = "red") +  # smoothed line to detect trends
    labs(x = "Fitted values", y = "Residuals") +
    theme_minimal() +
    ggtitle("Residuals vs Fitted Plot")
  
  # 2. Set up the plot layout for two plots side by side (Base R plot)
  par(mfrow = c(1, 2)) # Layout for two plots side by side

  # 2a. Residuals vs Fitted Plot (Base R)
  plot(fitted_values, residuals, 
       xlab = "Fitted Values", 
       ylab = "Residuals", 
       main = "Residuals vs Fitted",
       pch = 16, col = "gray")
  abline(h = 0, col = "red", lty = 2) # Add a horizontal line at 0
  
  # 2b. Spread-Level Plot (Scale-Location Plot)
  plot(fitted_values, sqrt(abs(residuals)), 
       xlab = "Fitted Values", 
       ylab = "Sqrt(|Residuals|)", 
       main = "Spread-Level Plot",
       pch = 16, col = "gray")

  # 3. ACF and PACF of Residuals (Base R)
  acf(residuals, main = "ACF of Residuals")
  pacf(residuals, main = "PACF of Residuals")
  
  par(mfrow = c(1, 1)) # Layout for two plots side by side
  # 4. Q-Q plot of Residuals (Base R)
  qqnorm(residuals)
  qqline(residuals, col = "red")  # Add a red line to the Q-Q plot
}
```

## Check distribution
```{r}
behavior_sum <- behavior_clean %>%
  group_by(treatment, mhw, genet) %>%
  summarize(n = n(),
            mean = mean(percent_fully_open),
            sd = sd(percent_fully_open))

behavior_box <- behavior_clean %>%
  group_by(treatment, mhw, genet)
            
ggplot(behavior_box, aes(x=genet, y=percent_fully_open)) +
  #geom_point() +
  geom_boxplot(aes(fill = genet)) +
  facet_grid(~ treatment+mhw)

normality<-behavior_clean %>%
  group_by(treatment, mhw, genet) %>%
  shapiro_test(percent_fully_open)

behavior_c <- behavior_clean %>%
  filter(genet == "C", 
         mhw == "after",
         treatment == "cold")

ggplot(behavior_c, aes(x=percent_fully_open)) +
  geom_histogram() +
  facet_wrap(~treatment)

ggplot(behavior_clean, aes(sample = percent_fully_open, group = mhw)) +
  stat_qq() + 
  stat_qq_line() +
  facet_grid(genet ~ treatment) +  # Facet by genet and treatment combinations
  theme_minimal() + 
  labs(title = "QQ Plot Faceted by Genet and Treatment",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")
```

## Try corARMA model
```{r}
autoArima <- auto.arima(behavior_clean$percent_fully_open, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(autoArima)

behavior_arma <- lme(percent_fully_open ~ treatment * genet + mhw,
                   random = ~1|genet_tank,
                   data = behavior_clean,
                   correlation = corARMA(p = 4, q = 1))
```

## Summary corARMA
```{r}
summary(behavior_arma)
residual_diagnostics(behavior_arma)
```

## Check basic GLMM model
```{r}
model <- lmer(percent_fully_open ~ treatment * genet * mhw + (1|genet_tank), data = behavior_clean)
model2 <- lmer(percent_fully_open ~ treatment + genet + mhw + (1|genet_tank), data = behavior_clean)

anova(model, model2)
```
*MHW interaction is a significantly better fit that +MHW

## Check model diagnostics
```{r}
residual_diagnostics(model)
```
Not good. Let's check with avg_temp instead of treatment?
 
## Avg temp instead of treatment
```{r}
model <- lmer(percent_fully_open ~ treatment * genet * mhw + (1|genet_tank), data = behavior_clean)
model_temp <- lmer(percent_fully_open ~ avg_temp * genet * mhw + (1 | genet_tank), data = behavior_clean)

anova(model, model_temp)

summary(model_temp)

residual_diagnostics(model_temp)
```
It's about the same performance. Adding *mhw is better model performance. Also there's a lag so let's try incorporating that...

## Incorporate time lag
```{r}
model_with_ar1 <- lme(percent_fully_open ~ avg_temp * genet * mhw, 
                      random = ~ 1 | genet_tank,
                      correlation = corAR1(form = ~ date_num | genet_tank), 
                      data = behavior_clean)

model_with_arma <- lme(percent_fully_open ~ avg_temp * genet * mhw, 
                      random = ~ 1 | genet_tank,
                      correlation = corARMA(p = 4, q = 1),
                      #correlation = corAR1(form = ~ date_num | genet_tank), 
                      data = behavior_clean)

model_with_random_slopes <- lme(percent_fully_open ~ avg_temp * genet * mhw, 
                                random = ~ date_num | genet_tank, 
                                data = behavior_clean)

anova(model_with_arma, model_with_random_slopes)
```
None of these significantly improve diagnostic plots.

## Let's check polynomial model
```{r}
model_poly <- lmer(percent_fully_open ~ poly(avg_temp, 2) * genet * mhw + (1 | genet_tank), 
                   data = behavior_clean)
anova(model_temp, model_poly)

residual_diagnostics(model_poly)
```
Polynomial model is not better diagnostic. 

Now that I'm looking at all of them and comparing with internet QQ plots, the QQ plots look "fine". 
Let's just go with best fitting model. 

## Let's check different time influences
```{r}
model_temp_time <- lmer(percent_fully_open ~ date_num + avg_temp * genet * mhw + (1 | genet_tank), data = behavior_clean)
#model_with_time_random <- lmer(percent_fully_open ~ avg_temp * genet * mhw + (1 | genet_tank) + (1 | date_num), data = behavior_clean) #WAAAAAY overfitting
#model_with_time_randomSlope <- lmer(percent_fully_open ~ avg_temp * genet * mhw + (date_num | genet_tank), data = behavior_clean)

anova(model_temp_time, model_with_time_random) 

residual_diagnostics(model_poly)

model_with_autocorrelation <- lme(percent_fully_open ~ avg_temp * genet * mhw + date_num, 
                                  random = ~ 1 | genet_tank, 
                                  correlation = corAR1(form = ~ date_num | genet_tank),
                                  data = behavior_clean)
# For model_with_autocorrelation
AIC(model_with_autocorrelation)
BIC(model_with_autocorrelation)
logLik(model_with_autocorrelation)
deviance(model_with_autocorrelation)

# For model_temp_time
AIC(model_temp_time)
BIC(model_temp_time)
logLik(model_temp_time)
deviance(model_temp_time)

residual_diagnostics(model_with_autocorrelation)
Box.test(residuals(model_temp_time), lag = 10, type = "Ljung-Box")
```
Need to do new types of autocorrelation... because temporal autocorrelation is very much present!

## Try different types of autocorrelation
```{r}
modelAR <- lme(percent_fully_open ~ avg_temp * genet * mhw + date_num,
                       random = ~ 1 | genet_tank,
                       correlation = corAR1(form = ~ date_num | genet_tank, fixed = FALSE),
                       #correlation = corARMA(p = 4, q = 1, form = ~ date_num | genet_tank),
                       data = behavior_clean)

modelARMA <- lme(percent_fully_open ~ avg_temp * genet * mhw,
                       random = ~ 1 | genet_tank,
                       #correlation = corAR1(form = ~ date_num | genet_tank, fixed = FALSE),
                       correlation = corARMA(p = 4, q = 1, form = ~ 1 | genet_tank),
                       data = behavior_clean)

modelARMA1 <- lme(percent_fully_open ~ avg_temp * genet * mhw + date_num,
                       random = ~ 1 | genet_tank,
                       correlation = corARMA(p = 4, q = 1, form = ~ 1 | genet_tank),
                       data = behavior_clean)

modelARMA2 <- lme(percent_fully_open ~ treatment * genet + date_num,
                       random = ~ 1 | genet_tank,
                       correlation = corARMA(p = 4, q = 1, form = ~ 1 | genet_tank),
                       data = behavior_clean)


modelARMA3 <- lme(percent_fully_open ~ avg_temp * genet + date_num,
                       random = ~ 1 | genet_tank,
                       correlation = corARMA(p = 4, q = 1, form = ~ 1 | genet_tank),
                       data = behavior_clean)

AIC(modelARMA3, modelARMA2)
BIC(modelARMA3, modelARMA2)

summary(modelARMA2)

residual_diagnostics(modelARMA1)
residual_diagnostics(modelARMA2)
```

Temporal autocorrelation still present but I don't have capacity to deal with it right now without consulting someone.

MHW is not significant when we are already accounting for date_num and avg_temp. All around date_num provides better fit than MHW, even when we're modeling treatment and not avg_temp. 

# Here's the "final" model (pre-consultation) - What I'm presenting at WSN
```{r}
model_final <- lme(percent_fully_open ~ avg_temp * genet + date_num,
                       random = ~ 1 | genet_tank,
                       correlation = corARMA(p = 4, q = 1, form = ~ 1 | genet_tank),
                       data = behavior_clean)
summary(model_final)
model_final_table <- round(summary(model_final)$tTable,4)
kable(model_final_table, digits = 3, caption = "Fixed Effects")

summary(emmeans(model_final, pairwise ~ genet | avg_temp))
```
1. Model Fit and Diagnostics

Start with a summary of the model fit, including relevant metrics (e.g., AIC, BIC, log-likelihood) and random effects structure:

Model Fit:

    The linear mixed-effects model was fitted using restricted maximum likelihood (REML).
    The model includes both fixed effects (e.g., temperature, genetic factors, date) and a random intercept for each genet_tank group, accounting for variability between tanks.
    The model was fitted with an ARMA(4,1) correlation structure to account for autocorrelation in the data over time.

Fit Statistics:

    AIC = -1699.9
    BIC = -1592.3
    logLik = 867.95

These statistics suggest a good model fit, as the AIC and BIC values are relatively low. Lower values of AIC and BIC indicate a better-fitting model.
2. Random Effects

Provide details of the random effects, including the variance and the structure of the correlation:

Random Effects:

    Random Intercept (per genet_tank): The standard deviation of the random intercept is very small (0.000025), indicating minimal variability between tanks.
    Residuals: The residual standard deviation is 0.2085, suggesting variability in the outcome after accounting for the fixed effects and random intercept.

Correlation Structure (ARMA(4,1)):

    ARMA Parameters:
        AR(1) = -0.6328
        AR(2) = 0.3797
        AR(3) = 0.2527
        AR(4) = 0.1922
        MA(1) = 0.7647

These parameters describe how previous values in the time series influence the current observation, accounting for autocorrelation.

3. Fixed Effects

Present the results for each predictor in the model, along with p-values and t-values:

Fixed Effects Estimates:

    Intercept: 0.6481 (t = 8.13, p < 0.001), indicating a significant baseline value for the dependent variable.
    Avg_temp: -0.0114 (t = -2.53, p = 0.0116), showing a small but significant negative effect of average temperature on the percentage of the fully open variable.
    Genet (B, C, D, E):
        GenetE: -0.3690 (t = -3.53, p = 0.0007), indicating a significant negative effect of Genet E compared to the reference group (presumably Genet A).
        Other Genet comparisons (B, C, D): All have p-values greater than 0.05 (e.g., GenetB: p = 0.8918, GenetC: p = 0.6894), suggesting these are not significantly different from the reference group.
    Date_num: 0.0017 (t = 10.07, p < 0.001), showing a significant positive trend over time.
    Interactions (avg_temp × genet):
        avg_temp × genetB: -0.0124 (t = -2.02, p = 0.0435), indicating a significant interaction effect between average temperature and Genet B.
        avg_temp × genetC: -0.0111 (t = -1.81, p = 0.0711), showing a marginally significant interaction effect for Genet C.
        avg_temp × genetD: 0.0012 (t = 0.19, p = 0.8505), indicating no significant interaction for Genet D.
        avg_temp × genetE: 0.0102 (t = 1.66, p = 0.0973), suggesting a marginally significant interaction with Genet E.

4. Collinearity

The correlation matrix for the fixed effects shows some correlations between predictors. For instance, avg_temp is highly correlated with the genet variables (r ~ 0.67). This suggests that the model accounts for multicollinearity, but these correlations should be considered when interpreting the individual fixed effects.

5. Residual Diagnostics

The standardized residuals range from -2.90 to 3.02, which indicates a relatively normal distribution of residuals, as extreme values are within an acceptable range. This suggests that the model does not suffer from large outliers or poor fit.
6. Conclusion and Interpretation

Based on the model results:

    Temperature (avg_temp) has a significant negative effect on the response variable, with a small but meaningful magnitude.
    Genetic factors (genetB, genetC, genetD, genetE) show varying effects, with GenetE exhibiting a significant negative effect relative to the reference group, while the other genetic factors are not significantly different from the reference.
    Date (date_num) has a significant positive effect, indicating a temporal trend in the data.
    There are significant interaction effects between temperature and some genetic factors (e.g., GenetB), but other interactions (e.g., GenetD) were not significant.

Given that GenetE and avg_temp × genetB are the most significant predictors, these variables should be prioritized for further investigation.
Suggested Reporting Style:

In the Results section of your paper, you might write something like this:

We fitted a linear mixed-effects model using REML to predict the percentage of fully open behavior based on average temperature, genetic factors, and date, while accounting for the correlation structure of the data using an ARMA(4,1) model. The model included a random intercept for genet_tank to account for tank-level variation.

The model fit was assessed using AIC (−1699.9) and BIC (−1592.3), indicating a good model fit. The random intercept variance was very small (SD = 0.000025), and the residual standard deviation was 0.2085.

Key findings include:

    Temperature (avg_temp) had a significant negative effect on the outcome (estimate = −0.0114, t = −2.53, p = 0.0116).
    Genetic factor GenetE showed a significant negative effect compared to the reference group (estimate = −0.3690, t = −3.53, p = 0.0007).
    The date variable had a significant positive effect, indicating a temporal trend (estimate = 0.0017, t = 10.07, p < 0.001).
    Significant interactions were observed between avg_temp and genetB (estimate = −0.0124, t = −2.02, p = 0.0435).

These findings suggest that both temperature and genetic factors, particularly GenetE, influence the behavior of interest, with a positive temporal trend across the study period.

# Make plot(s) for WSN
```{r}
genet_b <- behavior_clean %>%
  filter(genet == "B") %>%
  mutate(week = as.integer(difftime(date, min(date), units = "weeks")),
         weeks = cut(week, breaks = seq(1, max(week), by = 2), include.lowest = TRUE, labels = FALSE)) %>%
  group_by(weeks) %>%
  mutate(first_date = min(date)) %>%
         #first_date = if_else(first_date == ymd("2023-10-01"), ymd("2023-10-10"), first_date)) %>%
  ungroup()

# Generate a plot showing the relationship between avg_temp and percent_fully_open for each genetB group
ggplot(genet_b, aes(x = avg_temp, y = percent_fully_open, color = treatment)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment)) +
  labs(title = "Interaction between Avg Temp and GenetB on Percent Fully Open",
       x = "Average Temperature",
       y = "Percent Fully Open",
       color = "GenetB") +
  theme_minimal()
  
ggplot(genet_b, aes(x = first_date, y = percent_open, fill = treatment, color=treatment)) +
  geom_vline(xintercept = as.numeric(as.POSIXct("2023-12-04")), linetype = "dashed", color = "black", alpha = 0.8) +
  geom_point(alpha = 0.5) +
  geom_boxplot(aes(group = first_date), alpha=0.3) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +  
  scale_fill_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  scale_x_datetime(date_labels = "%b %d") +
  labs(x = "Date",
       y = "% open Genet B polyps",
       color = "") +
  guides(fill = "none",
         color = "none") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 20))

#ggsave(here("experiment", "figures", "WSN", "behaviorD.png"), width=12, height=7)
```


