---
title: "Analyzing body size data collected by UCSB undergraduate students"
author: "Amelia Ritger"
date: "2024-08-21"
output: html_document
---

# load libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(here)
library(tidyverse)
library(janitor)
library(ggpubr)
```

# read in .xslx files
```{r, message = FALSE}
rm(list=ls()) #start with a clean environment

pathname <- here("experiment", "data", "bodySize")
file_names <- list.files(path = pathname, 
                         full.names = F,
                         pattern='*.xlsx') #get a list of the names of the files (only .xlsx)

for (temp_name in file_names){
  print(temp_name)
  file.temp <- read_excel(paste0(pathname, "/", temp_name), trim_ws = TRUE) %>%
    select(!matches("[[:digit:]]")) %>% #remove columns that contain numbers (i.e. empty columns)
    clean_names() %>%
    fill(sampler, date_analysis, photo_name, tile_number, .direction = "down") %>% #fill in missing rows
    mutate(tile_number = as.factor(tile_number)) %>%
    #rename column containing some version of the word "flag" as "hello"
    rename(flag = contains("flag")) %>% #rename column containing some version of the word "flag" as "hello"
    select(1:10, flag)
  assign(gsub(".xlsx", "", temp_name, fixed = T), file.temp)
}
```

# merge all dataframes into one
```{r}
all_data <- Reduce(full_join, mget(ls(pattern = "Corynactis"))) %>%
  #mutate_at(vars(contains("cm")), ~ .x * 10) %>% #convert cm to mm
  mutate(basal_disk_diameter_mm_a = coalesce(basal_disk_diameter_mm_a, basal_disk_diameter_cm_a),   #move cm over to mm column
         basal_disk_diameter_mm_b = coalesce(basal_disk_diameter_mm_b, basal_disk_diameter_cm_b), #move cm over to mm column
    basal_disk_diameter_mm_a = ifelse(sampler != "Sanjana",  basal_disk_diameter_mm_a*10, basal_disk_diameter_mm_a),
    basal_disk_diameter_mm_b = ifelse(sampler != "Sanjana", basal_disk_diameter_mm_b*10, basal_disk_diameter_mm_b)) %>%
  select(!contains("cm")) %>% #remove cm columns
  mutate_at(vars(contains("basal_disk_diameter")), ~ round(., 1)) %>% #round to nearest decimal place
  filter(!is.na(basal_disk_diameter_mm_a | basal_disk_diameter_mm_b) & basal_disk_diameter_mm_a | basal_disk_diameter_mm_b != 0) #remove NA/0 measurements

data_clean <- all_data %>%
  filter(date_photo > ymd("2023-10-01"), #remove Oct 1 data - way too much variation in student measurements
         !(sampler == "Julia" & date_photo == ymd("2023-11-03")),#remove Julia's data from 11/3/2023
         !(sampler == "Lili" & grepl("sideways", flag))) %>% #remove sideways polyps from Lili's data 
  mutate(avg_diameter_mm = (basal_disk_diameter_mm_a+basal_disk_diameter_mm_b)/2,
         treatment = factor(case_when(tank_number < 6 ~ "cold", tank_number > 10 ~ "extreme", TRUE ~ "severe"),
                            levels = c("cold", "severe", "extreme")),
         tank=as.factor(tank_number),
         genet=as.factor(toupper(genet)),
         date=date_photo,
         mhw = factor(case_when(date < ymd("2023-12-12") ~ "during", TRUE ~ "after"),
                     levels = c("during", "after")),
         ww_g = 0.0975+3.02*avg_diameter_mm+1.08*avg_diameter_mm^2) %>% #calculate wet weight in grams, equation from 2019/2020 Corynactis body size measurements
  select(sampler, date, tank, genet, treatment, mhw, basal_disk_diameter_mm_a, basal_disk_diameter_mm_b, avg_diameter_mm, ww_g, flag)
```

# start some vis for QA/QC

## check for missing data
```{r}
data_check <- all_data %>%
  group_by(tank_number, date_photo, sampler) %>%
  filter(tank_number %in% c(2,4,6,9,11,14)) %>% #change tank number accordingly 2,4,6,9,11,14
  summarize(n = n()) #%>% #comment out everything after this to check overall # unique samplers
  #group_by(tank_number, date_photo) %>%
  #summarize(n=n())

#visualize with histogram
ggplot(all_data, aes(x = tank_number, color = sampler)) +
  geom_histogram() +
  scale_x_continuous(breaks = unique(all_data$tank_number))
```

## visualize average body size per tank between samplers
```{r}
#Checked+corrected data based on max measurement + 2 s.d. + any measured polyps > +-4 polyps counted
sampler_check <- data_clean %>%
  group_by(tank, date, genet, sampler) %>%
  summarize(max_size = max(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b), na.rm=T),
            avg_size = mean(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b), na.rm=T),
            sd_size = sd(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b), na.rm=T))  %>%
  ungroup() %>%
  arrange(-max_size, tank, genet)

sampler_check_sum <- sampler_check %>%
  group_by(tank, genet) %>%
  summarize(avg_size_all = mean(avg_size, na.rm=T),
            sd_size = sd(avg_size, na.rm=T),
            min_size = min(avg_size, na.rm=T),
            max_size = max(avg_size, na.rm=T)) %>%
  ungroup()

sampler_check_sum_sum <- sampler_check_sum %>%
  summarize(mean_sd = mean(sd_size, na.rm=T),
            sd_sd = sd(sd_size, na.rm=T))

ggplot(sampler_check, aes(x = date, y = avg_size, color = genet)) +
  geom_point(aes(shape=sampler)) +
  labs(y="Average size (mm)")
```

# Compare avg # polyps counted per tank with the master sheet 
```{r}
master <- read_csv(here("experiment", "data", "mortality.csv")) %>% #each file will be read in, specify which columns you need read in to avoid any error
  clean_names() %>%
  mutate(date = if_else(mdy(date) == "2024-01-31", mdy(date)-days(1), mdy(date)),
         tank = as.factor(tank),
         genet = as.factor(genet)) %>%
  select(date, tank, genet, alive, n, removed)

data_to_check <- data_clean %>%
  group_by(genet, tank, date) %>%
  summarize(n_polyps_samplers = n()/n_distinct(sampler),
            n_samplers = n_distinct(sampler)) %>%
  filter(tank == 4)

merged_check <- left_join(master, data_to_check, by = c("date", "tank", "genet")) %>%
  filter(!is.na(n_polyps_samplers)) %>%
  mutate(diff = abs(n-n_polyps_samplers),
         diff2 = abs(alive-n_polyps_samplers),
         #identify minimum value between diff and diff2
         min_diff = pmin(diff, diff2)) %>%
  arrange(tank, date, genet) %>%
  filter(tank == 4)

data_check <- data_clean %>%
  filter(date == ymd("2023-12-08"),
         tank == 1,
         genet == "D")
```


# Now the data has been checked and corrected, time for the proper plotting

## Create dataset - body size per tank for each genet
```{r}
body_size <- data_clean %>%
  group_by(genet, tank, date, treatment, mhw) %>%
  summarize(avg_size = mean(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b)),
            sd_size = sd(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b)),
            avg_biomass_g = mean(ww_g),
            sd_biomass_g = sd(ww_g),
            total_biomass_g = sum(ww_g)/(n()/n_distinct(sampler)),
            n_polyps = n()/n_distinct(sampler))
            #samplers=n_distinct(sampler))
```

## Plot average body size between treatments
```{r}
ggplot(body_size, aes(x=date, y=n_polyps, color=treatment)) +
  geom_point(aes(), size=1, alpha=0.3) +
  geom_smooth(aes(fill=treatment, size=mhw), method="lm", se=T) +
  #stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, hjust = -1, vjust = 10) +  
  scale_size_manual(values = c("during" = 1, "after" = 1)) + # Define sizes for mhw levels
  scale_fill_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Avg. number of polyps",
       color = "Treatment",
       fill = "Treatment") +
  guides(size = "none")

#ggsave(here("experiment", "figures", "n_polyps_MHW.png"), width=12, height=7)
```

## Create dataset - body size per treatment for each genet
```{r}
body_size_group <- data_clean %>%
  group_by(genet, treatment, date) %>%
  summarize(avg_size = mean(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b), na.rm=T),
            sd_size = sd(c(basal_disk_diameter_mm_a, basal_disk_diameter_mm_b), na.rm=T),
            n=n())
```

## Plot average body size between treatments for each genet
```{r}
#plot it
ggplot(body_size_group, aes(x=date, y=avg_size, group=treatment)) +
  geom_point(aes(color=treatment)) +
  geom_ribbon(aes(ymin=avg_size-sd_size, ymax=avg_size+sd_size, fill=treatment), alpha=0.2) +
  geom_line(aes(color=treatment), size=1) +
  scale_fill_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  facet_wrap(~genet) +
  labs(y="Average size (mm)",
       x="Date")
```

## Plot average body size between genets for each treatment
```{r}
ggplot(body_size, aes(x=date, y=avg_size, group=genet)) +
  geom_ribbon(aes(ymin=avg_size-sd_size, ymax=avg_size+sd_size, fill=genet), alpha=0.2) +
  geom_point(aes(color=genet), size=1) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  facet_wrap(~treatment, labeller = labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  labs(y="Average size (mm)",
       x="Date",
       color="Genet",
       fill="Genet")
```

