---
title: "Corynactis MHWsim experiment data analysis"
author: "Amelia Ritger"
date: "2023-11-01"
output: html_document
---

# Load packages 
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(ggpmisc) #statpolyeq ggplot
library(forecast) #ARIMA
library(tseries) #ADF test
library(urca) #KPSS test
library(forecast) #ts()
library(ggrepel) #geom_text_repel
library(broom) #tidy()
library(gridExtra) #grid.arrange()
```

# Run this ONE TIME to load temperature data, then save the file to your local machine
```{r}
source("readRPiData.R", local = knitr::knit_global())
```

# Load Tidied Data
```{r, include = FALSE}
source("mergeAllData.R", local = knitr::knit_global())

#Load temperature data AFTER you have run source(ReadRPiData)
rpi_temp <- read_csv(here("experiment", "data", "rpi_temp.csv"))
```

# Visualize the data

## Data QA/QC plots
```{r}
ggplot(all, aes(x=date, y=diff_n_cont, color=genet)) +
  geom_point() +
  facet_wrap(~treatment) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal()
```

## Plot basic growth rate across entire experiment
```{r, fig.width=10}
ggplot(all, aes(x=date, y=growth_rate_cont, color=genet)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, na.rm=TRUE) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")

#ggsave(here("experiment", "figures", "growth_genet.png"), width=12, height=7)
```

## Plot growth rate of all genets during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~mhw, scales="free", labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Treatment")

#ggsave(here("experiment", "figures", "growth_MHW.png"), width=12, height=7)
```

## Plot growth rate of each genet during and after MHW
```{r, fig.width=10, fig.height=7}
ggplot(all, aes(x=date, y=growth_rate, color = genet)) +
  geom_point(alpha=0.1) +
  facet_grid(mhw~treatment, 
             #switch = 'x',
             scales = "free_x",
             labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"),
                                 treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  #coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")
  theme(strip.placement = "outside")

#ggsave(here("experiment", "figures", "growth_MHW_genet.png"), width=12, height=10)
```

## Plot growth rate of each genet during and after MHW - but not faceted by MHW
```{r}
label_plot <- c("0.13", "0.09", "0.071", "0.0522", "0.0761",
                "0.151", "0.0933", "0.0252", "0.116", "0.0413", #extreme during
                "0.166", "0.142", "0.0522", "0.134", "0.0902",
                "-0.0023", "0.00741", "0.026", "0.00843", "-0.0131",
                "0.0102", "0.00652", "0.00881", "0.00421", "-0.0117", #extreme after
                "-0.0104", "-0.000895", "0.0217", "0.0127", "-0.0131")
color_plot <- rep(c("#9370DB", "#C21B78", "#FF9933", "#FF3333", "#662B45"), times = 6)
treatment_plot <- rep(c(rep("cold", 5), rep("extreme", 5), rep("severe", 5)), times = 2)
date_plot <- c(rep("2023-10-01", 15), rep("2024-01-10", 15))

p <- ggplot(all, aes(x=date, y=growth_rate_cont, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +   geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
    geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed") +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
  #stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  #un-comment this to get updated values to add to label_plot
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  p <- p + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 25-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
p

#ggsave(here("experiment", "figures", "growth_MHW_genet_combined.png"), width=12, height=7)
```

## Plot raw population numbers
```{r}
g <- ggplot(all, aes(x=date, y=n_true, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +
  geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed", linewidth = 0.5) +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
 # stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  #coord_cartesian(ylim = c(5, 40), expand=TRUE) +
  labs(x = "Date",
       y = "Number of polyps",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  g <- g + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 40-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
g

#ggsave(here("experiment", "figures", "n_MHW_genet_combined.png"), width=12, height=7)
```

## Plot behavior data
```{r, fig.width=10}
# %closed facet grid
ggplot(all, aes(x=genet, y=percent_closed, fill=genet, group=genet)) +
  geom_point(aes(color=genet)) +
  geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
  facet_grid(mhw~treatment)

# %closed facet wrap
ggplot(all, aes(x = genet, y = percent_closed, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9)) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.5, outlier.colour = "transparent") +
  facet_wrap(~treatment) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5)) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17))  # Define shapes for mhw levels

# %fullyOpen facet wrap
ggplot(all, aes(x = genet, y = percent_fully_open, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9), alpha = 0.5) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.3, color = alpha("black", 0.75)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5), labels = c("During MHW", "After MHW")) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17), labels = c("During MHW", "After MHW")) + # Define shapes for mhw levels
  # change color of violin 
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Genet",
       y = "% open polyps",
       color = "Genet",
       shape = "MHW",
       size = "MHW") +
  guides(fill = "none",
         color = "none")

#ggsave(here("experiment", "figures", "open_MHW_genet.png"), width=12, height=7)
```

## Plot total biomass over time
```{r}
ggplot(all, aes(x=date, y=total_biomass)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="lm", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Total biomass (g)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "total_biomass_mhw.png"), width=12, height=7)
```

## Plot average body size over time
```{r}
ggplot(all, aes(x=date, y=avg_size)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="loess", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Average body size (mm)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "avg_size_mhw.png"), width=12, height=7)
```

## Plot mortality (lol)
```{r}
ggplot(all, aes(x=date, y=dying_dead, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~treatment, 
             labeller = labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00")) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10), labels = function(x) format(x, nsmall = 0)) +
  labs(x = "Date",
       y = "Mortality (no. polyps)",
       color = "Treatment") +
  guides(color = "none")

#ggsave(here("experiment", "figures", "mortality_MHW.png"), width=12, height=7)
```

############################################################################################################

# Start running stats

## Count data - GLARMA 
```{r}

```

## Body size data - ARIMA forecasting
https://stats.stackexchange.com/questions/281666/how-does-acf-pacf-identify-the-order-of-ma-and-ar-terms
https://stats.stackexchange.com/questions/32634/difference-time-series-before-arima-or-within-arima

### Log transform data
```{r}
all_size <- all %>%
  filter(!is.na(avg_size)) %>%
  mutate(avg_size_log = log(avg_size))

# scatter plot of transformed data
ggplot(all_size, aes(x=date, y=avg_size_log)) +
  geom_point(aes(color=treatment)) +
  labs(x = "Average body size (mm)",
       y = "Frequency")
# histogram of transformed data
ggplot(all_size, aes(x=avg_size_log)) +
  geom_histogram(binwidth=0.1) +
  labs(x = "log(Average body size (mm))",
       y = "Frequency")
# transformed data density plot
ggplot(all_size, aes(x = avg_size_log)) +
  geom_density() +
  ggtitle("Density Plot of Transformed Data")
# Q-Q plot of log-transformed data
qqnorm(all_size$avg_size_log)
qqline(all_size$avg_size_log, col = "red")
```
It's not perfectly normal but definitely better log-transformed. Looks a bit like exponential decay, so let's make the data stationary by differencing.

### Difference to make the data stationary
```{r}
all_size <- all_size %>%
  group_by(tank, genet) %>%
  mutate(avg_size_diff = avg_size_log - lag(avg_size_log),
         avg_size_diff2 = avg_size_diff - lag(avg_size_diff)) #need second order differencing for slope=0

ggplot(all_size, aes(x=date)) +
  geom_point(aes(y=avg_size_diff), color="black", alpha=0.3) +
  geom_point(aes(y=avg_size_diff2), color="blue", alpha=0.3) +
  geom_smooth(aes(y=avg_size_diff), method="lm", color="black") +
  geom_smooth(aes(y=avg_size_diff2), method="lm", color="blue") +
  labs(x = "Average body size (mm)",
       y = "Frequency")
```

### Plot a slightly more complex vis of the first differenced data
```{r, fig.width=10}
# Fit linear models and extract coefficients
models <- all_size %>%
  group_by(treatment, genet, mhw) %>%
  do(tidy(lm(avg_size_diff ~ date, data = .))) %>%
  select(treatment, genet, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  mutate(equation = sprintf("y = %.2f + %.2e * x", `(Intercept)`, date)) %>%
  rename(slope = date) %>%
  ungroup() %>%
  group_by(mhw) %>%
  mutate(x_axis = as.POSIXct(ifelse(mhw == "during", "2023-10-15", "2024-01-15")),
         y_axis = rep(seq(from = 0.2, to = 0.5, length.out = 5), length.out = n())) %>%
  ungroup() %>%
  select(-mhw)

# Join regression equations with the original data
all_size_reg <- all_size %>%
  left_join(models, by = c("treatment", "genet"), relationship = "many-to-many")

ggplot(all_size_reg) +
  geom_point(aes(x=date, y=avg_size_diff, color=genet), alpha=0.3) +
  geom_smooth(aes(x=date, y=avg_size_diff, color=genet ,linetype = mhw), method="lm") +
  facet_wrap(~treatment) +
  geom_label_repel(data = models, aes(x=x_axis, y=y_axis, label = equation, color = genet), size = 3) +
  labs(x = "Date",
       y = "Differenced average body size (mm)") +
  theme_bw()

#ggsave(here("experiment", "figures", "body_size_regression.png"), width=15, height=7)
```

### Run ADF and KPSS tests on first and second differenced data
```{r}
# List to store results
adf_results <- list()

# Loop through each group
for (group in unique(all_size$treatment)) {
  group_data <- all_size %>%
    filter(treatment == group,
           !is.na(avg_size_diff2)) %>%
    pull(avg_size_diff2) #chnage this to avg_size_diff for first differenced data
  # Perform the ADF test
  adf_test <- adf.test(group_data, alternative = "stationary")
  # Store the result
  adf_results[[group]] <- adf_test
}

# Print results
adf_results

# List to store results
kpss_results <- list()
# Loop through each group
for (group in unique(all_size$treatment)) {
  group_data <- all_size %>%
    filter(treatment == group,
           !is.na(avg_size_diff2)) %>%
    pull(avg_size_diff2) #chnage this to avg_size_diff for first differenced data
  # Perform the KPSS test
  kpss_test <- ur.kpss(group_data, type = "mu")  # Use "tau" if you expect trend stationarity
  # Store the result
  kpss_results[[group]] <- summary(kpss_test)
}

# Print results
kpss_results
```
Both p-value is less than 0.01 and test statistic is lower than critical values, so second-order differenced data is now almost certainly stationary. First-order differencing only showed strong evidence for stationarity with the ADF test.

### Plot ACF and PACF for the log-transformed body size data
```{r}
Acf(all_size$avg_size_log, main = "ACF of Log-transformed Series")
Pacf(all_size$avg_size_log, main = "PACF of Log-transformed Series")

Acf(all_size$avg_size_diff, main = "ACF of Differenced Series")
Pacf(all_size$avg_size_diff, main = "PACF of Differenced Series")

Acf(all_size$avg_size_diff2, main = "ACF of 2-Differenced Series")
Pacf(all_size$avg_size_diff2, main = "PACF of 2-Differenced Series")
```
There is definitely a lot of autocorrelation with the data, for both 1-diff and 2-diff. Makes sense because body size should be highly correlated over time.

### Remove NA values and convert factors to numeric
```{r}
all_size_na <- all_size %>%
  filter(!is.na(avg_size_diff)) %>%
  mutate(genet = as.numeric(genet),
         treatment = as.numeric(treatment),
         mhw = as.numeric(mhw)) %>%
  select(date, tank, genet, n, treatment, mhw, avg_size, avg_size_log, avg_size_diff, avg_size_diff2)
```

### First-differenced data

#### Auto ARIMA - basic
```{r}
autoArima <- auto.arima(all_size_na$avg_size_diff, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(autoArima)
```
I ran this all again with second-differenced data, and the model doesn't appear significantly better (in some cases like BIC, not better) and so it doesn't justify the added complexity of a second-differenced model. so let's stick with first-differenced model.

#### Check model residuals
```{r}
residuals_auto <- residuals(autoArima)
plot(residuals_auto, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto, main = "ACF of Residuals")
Box.test(residuals_auto, type = "Ljung-Box")

adf_test_residuals <- adf.test(residuals_auto, alternative = "stationary") #definitely stationary
qqnorm(residuals_auto)
qqline(residuals_auto)
shapiro.test(residuals_auto)

ggplot() +
  geom_histogram(aes(x = residuals_auto), bins = 20, color = "black") +
  labs(x = "Residuals",
       y = "Frequency")

fitted_values <- fitted(autoArima)
plot(fitted_values, residuals_auto, main="Residuals vs Fitted Values")
abline(h=0, col="red")
```
2-order differencing is not significantly better than 1-order differencing so let's stick with 1-order. 

AR(1) model: This is autoregressive data, and the current value of the time series depends linearly on its immediately preceding value - makes sense because body size shouldn't change that drastically week by week. 
I(1): First-order differencing applied to make the series stationary - which makes sense because preliminary analyses showed slope of zero with second-differenced data. 
MA(2): current value of the time series depends on the last two periods’ forecast errors - means that the current value of the series depends on the errors made in the previous two periods, which doesn't make a ton of intuitive sense except maybe if there are big changes to body size (i.e. reproduction?) that just means we won't see those changes finish until 2 weeks later? <<< CHECK THIS LOGIC.

#### Check ts trends/seasonality
```{r}
all_size_ts <- ts(all_size_na$avg_size_diff, frequency = 52) #weekly data as ts object
class(all_size_ts)
summary(all_size_ts)
acf(all_size_ts)
pacf(all_size_ts)

decomposition <- stl(all_size_ts, s.window = "periodic")
plot(decomposition)
```
There's no apparent seasonality and a slight positive trend in the data

#### Fit ARIMA with different I values
```{r}
# Fit ARIMA with different seasonal orders
model1 <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 1))
model2 <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2))
model3 <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 3))

Acf(residuals(model1), main = paste("ACF of Residuals model1"))
Pacf(residuals(model1), main = paste("PACF of Residuals model1"))

Acf(residuals(model2), main = paste("ACF of Residuals model2"))
Pacf(residuals(model2), main = paste("PACF of Residuals model2"))

Acf(residuals(model3), main = paste("ACF of Residuals model3"))
Pacf(residuals(model3), main = paste("PACF of Residuals model3"))

Box.test(residuals(model1), lag = 20, type = "Ljung-Box")
Box.test(residuals(model2), lag = 20, type = "Ljung-Box")
Box.test(residuals(model3), lag = 20, type = "Ljung-Box")
```
I(2) is definitely the best fit. 

#### Consider potential covariates
```{r}
#create matrix of dummy variables incorporating all potential covariates
xreg_All <- model.matrix(~ treatment + genet + mhw - 1, data = all_size_na)
xreg_NoGenet <- model.matrix(~ treatment + mhw - 1, data = all_size_na)
xreg_MHW <- model.matrix(~ mhw - 1, data = all_size_na)
xreg_Treatment <- model.matrix(~ treatment - 1, data = all_size_na)

arima_covariates_all <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_All)
summary(arima_covariates_all)

arima_covariates_noGenet <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_NoGenet)
summary(arima_covariates_noGenet)

arima_covariates_mhw <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_MHW)
summary(arima_covariates_mhw)

arima_covariates_treatment <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_Treatment)
summary(arima_covariates_treatment)
```
Looks like genet is not a significant covariate, but treatment and during/after MHW are slightly impactful. <<< DISCUSS WITH CHRIS.

Auto-ARIMA with Covariates model (check AR, I, MA)
```{r}
xreg_Covariate <- model.matrix(~ genet + mhw - 1, data = all_size_na)

autoArima_Covariate <- auto.arima(all_size_na$avg_size_diff, seasonal = TRUE, stepwise = FALSE, approximation = FALSE, xreg = xreg_Covariate)
summary(autoArima_Covariate)
```
No change to AR, I, or MA with any of the potential covariates.

*So, for first-differenced data, here are the two model options*
#### 
Basic ARIMA(1,1,2): 
Coefficients:
         ar1      ma1     ma2
      0.7682  -1.5296  0.5354
s.e.  0.0518   0.0662  0.0646

sigma^2 = 0.002058:  log likelihood = 2258.51
AIC=-4509.02   AICc=-4508.99   BIC=-4488.19

                      ME       RMSE        MAE MPE MAPE      MASE         ACF1
Training set 0.001281871 0.04529954 0.03479757 Inf  Inf 0.7851294 -0.009148644

####
####
ARIMA(1,1,2) with MHW/Treatment covariates:
Coefficients:
         ar1      ma1     ma2  treatment     mhw
      0.7661  -1.5293  0.5351    -0.0029  0.0012
s.e.  0.0539   0.0682  0.0662     0.0025  0.0155

sigma^2 = 0.002059:  log likelihood = 2259.17
AIC=-4506.35   AICc=-4506.28   BIC=-4475.1

                      ME       RMSE        MAE MPE MAPE      MASE         ACF1
Training set 0.001324029 0.04527702 0.03477904 NaN  Inf 0.7847113 -0.008522662
####

#### Check model residuals for MHW/Treatment ARIMA
```{r}
residuals_auto <- residuals(arima_covariates_noGenet)
plot(residuals_auto, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto, main = "ACF of Residuals")
Box.test(residuals_auto, type = "Ljung-Box")

adf_test_residuals <- adf.test(residuals_auto, alternative = "stationary") #definitely stationary
qqnorm(residuals_auto)
qqline(residuals_auto)
shapiro.test(residuals_auto)

ggplot() +
  geom_histogram(aes(x = residuals_auto), bins = 20, color = "black") +
  labs(x = "Residuals",
       y = "Frequency")

fitted_values <- fitted(arima_covariates_noGenet)
plot(fitted_values, residuals_auto, main="Residuals vs Fitted Values")
abline(h=0, col="red")
```

### Now, let's check out ARIMA models on second-differenced data (since it is technically more stationary)
```{r}
autoArima2 <- auto.arima(all_size_na$avg_size_diff2, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(autoArima2)
```

AR(1) model: This is autoregressive data, and the current value of the time series depends linearly on its immediately preceding value - makes sense because body size shouldn't change that drastically week by week. 
I(0): No differencing applied to make the series stationary - which makes sense because the body size data has been differenced twice which we know makes it stationary.
MA(4): current value of the time series depends on the last FOUR periods’ forecast errors - means that the current value of the series depends on the errors made in the previous four periods. Greater value than first-differenced data model MA(2). Meaning??

#### Check model residuals
```{r}
residuals_auto2 <- residuals(autoArima2)
plot(residuals_auto2, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto2, main = "ACF of Residuals")
Box.test(residuals_auto2, type = "Ljung-Box")

adf_test_residuals2 <- adf.test(residuals_auto2, alternative = "stationary") #definitely stationary
qqnorm(residuals_auto2)
qqline(residuals_auto2)
shapiro.test(residuals_auto2)

ggplot() +
  geom_histogram(aes(x = residuals_auto2), bins = 20, color = "black") +
  labs(x = "Residuals",
       y = "Frequency")

fitted_values2 <- fitted(autoArima2)
plot(fitted_values2, residuals_auto2, main="Residuals vs Fitted Values")
abline(h=0, col="red")
```
Shapiro-Wilks test says residuals are close to normal distribution but not normally distributed... but the plots look fine.

#### Consider potential covariates
```{r}
#create matrix of dummy variables incorporating all potential covariates
xreg_All <- model.matrix(~ treatment + genet + mhw - 1, data = all_size_na)
xreg_NoGenet <- model.matrix(~ treatment + mhw - 1, data = all_size_na)
xreg_NoTreatment <- model.matrix(~ genet + mhw - 1, data = all_size_na)
xreg_NoMHW <- model.matrix(~ treatment + genet - 1, data = all_size_na)
xreg_MHW <- model.matrix(~ mhw - 1, data = all_size_na)
xreg_Treatment <- model.matrix(~ treatment - 1, data = all_size_na)
xreg_Genet <- model.matrix(~ genet - 1, data = all_size_na)

arima_no_covariates2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4))
summary(arima_no_covariates2)

arima_covariates_all2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_All)
summary(arima_covariates_all2)

arima_covariates_noGenet2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_NoGenet)
summary(arima_covariates_noGenet2)

arima_covariates_noTreatment2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_NoTreatment)
summary(arima_covariates_noTreatment2)

arima_covariates_noMHW2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_NoMHW)
summary(arima_covariates_noMHW2)

arima_covariates_Genet2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_Genet)
summary(arima_covariates_Genet2)

arima_covariates_mhw2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_MHW)
summary(arima_covariates_mhw2)

arima_covariates_treatment2 <- Arima(all_size_na$avg_size_diff2, order = c(1, 0, 4), xreg = xreg_Treatment)
summary(arima_covariates_treatment2)
```
TreatmentxGenetxMHW model performs the worst. TreatmentxMHW, GenetxMHW, TreatmentxGenet perform second worst. No covariates performs best, Treatment alone, Genet alone, and MHW alone all perform second best. This probably means we should use no covariates in second-differenced data model?

Auto-ARIMA with Covariates model (check AR, I, MA)
```{r}
xreg_Covariate2 <- model.matrix(~ treatment + genet + mhw - 1, data = all_size_na)

autoArima_Covariate2 <- auto.arima(all_size_na$avg_size_diff2, seasonal = TRUE, stepwise = FALSE, approximation = FALSE, xreg = xreg_Covariate2)
summary(autoArima_Covariate2)
```
Same as first-differenced data, no change to AR, I, or MA with any of the potential covariates.

*For second-differenced data, here is the apparent best model option*
####
Series: all_size_na$avg_size_diff2 
ARIMA(1,0,4) with zero mean 
Coefficients:
         ar1      ma1     ma2      ma3      ma4
      0.9289  -0.7029  0.0217  -0.0357  -0.1054
s.e.  0.0412   0.0506  0.0356   0.0353   0.0303

sigma^2 = 0.005237:  log likelihood = 1541.35
AIC=-3070.71   AICc=-3070.64   BIC=-3039.8

Training set error measures:
                       ME      RMSE        MAE       MPE    MAPE      MASE
Training set 0.0007083893 0.0722271 0.05541204 -20.12162 319.208 0.7668489
                    ACF1
Training set 0.000179781
#### 

### Plot differenced data with O.G. values

#### Genet only
```{r, fig.width=10}
# Create the plots
plot_undifferenced <- ggplot(all_size_na, aes(x = date, y = avg_size_log, group=as.factor(genet))) +
  geom_point(aes(color=as.factor(genet)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(genet)), method="lm") +
  ggtitle("Undifferenced Data") +
  xlab("Date") +
  ylab("Log Body size (mm)") +
  theme_bw()

plot_differenced <- ggplot(all_size_na, aes(x = date, y = avg_size_diff, group=as.factor(genet))) +
  geom_point(aes(color=as.factor(genet)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(genet)), method="lm") +
  ggtitle("Differenced Data") +
  xlab("Date") +
  ylab("Differenced Body Size (mm)") +
  theme_bw()

plot_differenced2 <- ggplot(all_size_na, aes(x = date, y = avg_size_diff2, group=as.factor(genet))) +
  geom_point(aes(color=as.factor(genet)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(genet)), method="lm") +
  ggtitle("2-Differenced Data") +
  xlab("Date") +
  ylab("2-Differenced Body Size (mm)") +
  theme_bw()

combined_plot <- grid.arrange(plot_undifferenced, plot_differenced, plot_differenced2, ncol = 3)

#ggsave(here("experiment", "figures", "differenced_comparison_genet.png"), combined_plot, width=14, height=7)
```
Well damn, that makes sense why genet is not a significant covariate in the differenced data models! 

#### Treatment only
```{r, fig.width=10}
# Create the plots
plot_undifferenced <- ggplot(all_size_na, aes(x = date, y = avg_size_log, group=as.factor(treatment))) +
  geom_point(aes(color=as.factor(treatment)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(treatment)), method="lm") +
  ggtitle("Undifferenced Data") +
  xlab("Date") +
  ylab("Log Body size (mm)") +
  theme_bw()

plot_differenced <- ggplot(all_size_na, aes(x = date, y = avg_size_diff, group=as.factor(treatment))) +
  geom_point(aes(color=as.factor(treatment)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(treatment)), method="lm") +
  ggtitle("Differenced Data") +
  xlab("Date") +
  ylab("Differenced Body Size (mm)") +
  theme_bw()

plot_differenced2 <- ggplot(all_size_na, aes(x = date, y = avg_size_diff2, group=as.factor(treatment))) +
  geom_point(aes(color=as.factor(treatment)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(treatment)), method="lm") +
  ggtitle("2-Differenced Data") +
  xlab("Date") +
  ylab("2-Differenced Body Size (mm)") +
  theme_bw()

combined_plot <- grid.arrange(plot_undifferenced, plot_differenced, plot_differenced2, ncol = 3)

#ggsave(here("experiment", "figures", "differenced_comparison_treatment.png"), combined_plot, width=14, height=7)
```
No difference in linear slope between three treatments

#### MHW only
```{r, fig.width=10}
# Create the plots
plot_undifferenced <- ggplot(all_size_na, aes(x = date, y = avg_size_log, group=as.factor(mhw))) +
  geom_point(aes(color=as.factor(mhw)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(mhw)), method="lm") +
  ggtitle("Undifferenced Data") +
  xlab("Date") +
  ylab("Log Body size (mm)") +
  theme_bw()

plot_differenced <- ggplot(all_size_na, aes(x = date, y = avg_size_diff, group=as.factor(mhw))) +
  geom_point(aes(color=as.factor(mhw)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(mhw)), method="lm") +
  ggtitle("Differenced Data") +
  xlab("Date") +
  ylab("Differenced Body Size (mm)") +
  theme_bw()

plot_differenced2 <- ggplot(all_size_na, aes(x = date, y = avg_size_diff2, group=as.factor(mhw))) +
  geom_point(aes(color=as.factor(mhw)), alpha=0.3) +
  geom_smooth(aes(color=as.factor(mhw)), method="lm") +
  ggtitle("2-Differenced Data") +
  xlab("Date") +
  ylab("2-Differenced Body Size (mm)") +
  theme_bw()

combined_plot <- grid.arrange(plot_undifferenced, plot_differenced, plot_differenced2, ncol = 3)

#ggsave(here("experiment", "figures", "differenced_comparison_mhw.png"), combined_plot, width=14, height=7)
```
MHW slope is definitely different in undifferenced data.

### Use basic first-differenced ARIMA to forecast.
```{r}
future_predictions <- forecast(autoArima, h = 50)  # Forecast h periods ahead

plot(future_predictions)
```

Just for conceptual understanding, create a quick autoARIMA model on undifferenced data and plot forecast
```{r}
autoArima_undiff <- auto.arima(all_size_na$avg_size_log, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(autoArima_undiff)

future_predictions <- forecast(autoArima_undiff, h = 50)  # Forecast h periods ahead
plot(future_predictions)
```

### TL;DR First Differenced AutoARIMA model = ARIMA(1,1,2)

## Body size data - MARSS, what we should have been doing all along...
https://atsa-es.github.io/MARSS/

```{r}
all_size_marss <- all_size
```

