---
title: "Corynactis MHWsim experiment data analysis"
author: "Amelia Ritger"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(ggpmisc) #statpolyeq ggplot
library(forecast) #ARIMA
```

# Load Tidied Data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
```

# Visualize the data

## Data QA/QC plots
```{r}
ggplot(all, aes(x=date, y=diff_n_cont, color=genet)) +
  geom_point() +
  facet_wrap(~treatment) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal()
```

## Plot basic growth rate across entire experiment
```{r}
ggplot(all, aes(x=date, y=growth_rate_cont, color=genet)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, na.rm=TRUE) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")

#ggsave(here("experiment", "figures", "growth_genet.png"), width=12, height=7)
```

## Plot growth rate of all genets during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~mhw, scales="free", labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Treatment")

#ggsave(here("experiment", "figures", "growth_MHW.png"), width=12, height=7)
```

## Plot growth rate of each genet during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = genet)) +
  geom_point(alpha=0.1) +
  facet_grid(mhw~treatment, 
             #switch = 'x',
             scales = "free_x",
             labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"),
                                 treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  #coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")
  theme(strip.placement = "outside")

#ggsave(here("experiment", "figures", "growth_MHW_genet.png"), width=12, height=10)
```

## Plot growth rate of each genet during and after MHW - but not faceted by MHW
```{r}
label_plot <- c("0.13", "0.09", "0.071", "0.0522", "0.0761",
                "0.151", "0.0933", "0.0252", "0.116", "0.0413", #extreme during
                "0.166", "0.142", "0.0522", "0.134", "0.0902",
                "-0.0023", "0.00741", "0.026", "0.00843", "-0.0131",
                "0.0102", "0.00652", "0.00881", "0.00421", "-0.0117", #extreme after
                "-0.0104", "-0.000895", "0.0217", "0.0127", "-0.0131")
color_plot <- rep(c("#9370DB", "#C21B78", "#FF9933", "#FF3333", "#662B45"), times = 6)
treatment_plot <- rep(c(rep("cold", 5), rep("extreme", 5), rep("severe", 5)), times = 2)
date_plot <- c(rep("2023-10-01", 15), rep("2024-01-10", 15))

p <- ggplot(all, aes(x=date, y=growth_rate_cont, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +   geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
    geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed") +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
  #stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  #un-comment this to get updated values to add to label_plot
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  p <- p + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 25-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
p

#ggsave(here("experiment", "figures", "growth_MHW_genet_combined.png"), width=12, height=7)
```

## Plot raw population numbers
```{r}
g <- ggplot(all, aes(x=date, y=n_true, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +
  geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed", linewidth = 0.5) +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
 # stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  #coord_cartesian(ylim = c(5, 40), expand=TRUE) +
  labs(x = "Date",
       y = "Number of polyps",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  g <- g + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 40-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
g

#ggsave(here("experiment", "figures", "n_MHW_genet_combined.png"), width=12, height=7)
```

## Plot behavior data
```{r}
# %closed facet grid
ggplot(all, aes(x=genet, y=percent_closed, fill=genet, group=genet)) +
  geom_point(aes(color=genet)) +
  geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
  facet_grid(mhw~treatment)

# %closed facet wrap
ggplot(all, aes(x = genet, y = percent_closed, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9)) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.5, outlier.colour = "transparent") +
  facet_wrap(~treatment) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5)) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17))  # Define shapes for mhw levels

# %fullyOpen facet wrap
ggplot(all, aes(x = genet, y = percent_fully_open, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9), alpha = 0.5) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.3, color = alpha("black", 0.75)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5), labels = c("During MHW", "After MHW")) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17), labels = c("During MHW", "After MHW")) + # Define shapes for mhw levels
  # change color of violin 
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Genet",
       y = "% open polyps",
       color = "Genet",
       shape = "MHW",
       size = "MHW") +
  guides(fill = "none",
         color = "none")

#ggsave(here("experiment", "figures", "open_MHW_genet.png"), width=12, height=7)
```

## Plot total biomass over time
```{r}
ggplot(all, aes(x=date, y=total_biomass_true)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="lm", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Total biomass (g)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "total_biomass_mhw.png"), width=12, height=7)
```

## Plot average body size over time
```{r}
ggplot(all, aes(x=date, y=avg_size)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="loess", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Average body size (mm)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "avg_size_mhw.png"), width=12, height=7)
```

## Plot mortality (lol)
```{r}
ggplot(all, aes(x=date, y=dying_dead, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~treatment, 
             labeller = labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00")) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10), labels = function(x) format(x, nsmall = 0)) +
  labs(x = "Date",
       y = "Mortality (no. polyps)",
       color = "Treatment") +
  guides(color = "none")

#ggsave(here("experiment", "figures", "mortality_MHW.png"), width=12, height=7)
```

############################################################################################################

# Start running stats

## Count data - GLARMA 
```{r}

```

## Body size data - Autoregression
### Log transform data
```{r}
all_transform <- all %>%
  filter(!is.na(avg_size)) %>%
  mutate(avg_size_log = log(avg_size))

# histogram of raw data
ggplot(all_transform, aes(x=avg_size)) +
  geom_histogram(binwidth=0.1) +
  labs(x = "Average body size (mm)",
       y = "Frequency")
# raw data density plot
ggplot(all_transform, aes(x = avg_size)) +
  geom_density() +
  ggtitle("Density Plot of Original Data")
# Q-Q plot of original data
qqnorm(all_transform$avg_size)
qqline(all_transform$avg_size, col = "red")

# histogram of transformed data
ggplot(all_transform, aes(x=avg_size_log)) +
  geom_histogram(binwidth=0.1) +
  labs(x = "log(Average body size (mm))",
       y = "Frequency")
# transformed data density plot
ggplot(all_transform, aes(x = avg_size_log)) +
  geom_density() +
  ggtitle("Density Plot of Transformed Data")
# Q-Q plot of log-transformed data
qqnorm(all_transform$avg_size_log)
qqline(all_transform$avg_size_log, col = "red")
```
It's not perfectly normal but definitely better log-transformed

```{r}
# Perform ADF test to check if data is stationary
adf.test(all_transform$avg_size_log)
# Plot ACF and PACF for the stationary log-transformed body size
Acf(all_transform$avg_size_log)
Pacf(all_transform$avg_size_log)

diff_series <- diff(all_transform$avg_size_log)
Acf(diff_series, main = "ACF of Differenced Series")
Pacf(diff_series, main = "PACF of Differenced Series")
```
Seasonal cycle has lag of 5 for whatever reason

```{r}
# Seasonal differencing
diff_series_seasonal <- diff(all_transform$avg_size_log, lag = 5)

Acf(diff_series_seasonal, main = "Seasonal ACF of Differenced Series")
Pacf(diff_series_seasonal, main = "Seasonal PACF of Differenced Series")
```
According to ChatGPT:
*Strong Peak at Lag 5 in ACF:*
The significant negative peak at lag 5 in the ACF suggests a strong seasonal effect with a period of 5 time units. This indicates that the data has a seasonal pattern with a period of 5 lags.

*Wave Pattern in ACF:*
The wave-like pattern with small positive and negative values elsewhere suggests that the seasonality is prominent at lag 5, but the data beyond this lag doesn’t exhibit strong autocorrelation. This is typical in many time series where the primary seasonality is evident at specific lags, and other autocorrelations diminish.

*Strong Peak at Lag 5 in PACF:*
A strong peak at lag 5 in the PACF suggests that there is a significant seasonal autoregressive (AR) component. This means that past values separated by a season (5 lags in your case) have a strong influence on the current value.

*Wave Pattern in PACF:*
The wave-like pattern with small values elsewhere in the PACF indicates that the influence of past values diminishes quickly after lag 5, reinforcing the idea that lag 5 is the primary seasonal effect.

**Determining Seasonal AR and MA Orders**
*Seasonal AR (PP):*
Since the PACF shows a strong peak at lag 5 and then cuts off or shows a diminishing pattern, it indicates a seasonal AR component. You might start with P=1 or P=2, but the specific choice will depend on how well it fits the data. In this case, P could be 1 if the PACF cuts off or diminishes significantly after lag 5.

*Seasonal MA (QQ):*
The ACF shows a strong peak at lag 5 and then diminishes. This suggests that a seasonal MA component is appropriate. Typically, Q=1 is a good starting point, but it can be adjusted based on model performance.

*Seasonal Differencing (DD):*
If the seasonal pattern persists despite differencing, you might need to include seasonal differencing (DD). Since your ACF and PACF both show strong patterns at lag 5, D=1 is likely appropriate to handle seasonal trends.

```{r}
fit_auto <- auto.arima(all_transform$avg_size_log, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(fit_auto)
residuals_auto <- residuals(fit_auto)
plot(residuals_auto, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto, main = "ACF of Residuals")
Box.test(residuals_auto, type = "Ljung-Box")
```

```{r}
# Fit ARIMA with different seasonal orders
model1 <- Arima(all_transform$avg_size_log, order = c(5, 1, 1), seasonal = list(order = c(1, 1, 1), period = 5))
model3 <- Arima(all_transform$avg_size_log, order = c(5, 1, 1), seasonal = list(order = c(1, 1, 1), period = 5))

Acf(residuals(model1), main = paste("ACF of Residuals model1"))
Pacf(residuals(model1), main = paste("PACF of Residuals model1"))

Acf(residuals(model2), main = paste("ACF of Residuals model2"))
Pacf(residuals(model2), main = paste("PACF of Residuals model2"))

Acf(residuals(model3), main = paste("ACF of Residuals model3"))
Pacf(residuals(model3), main = paste("PACF of Residuals model3"))

Box.test(residuals(model1), lag = 20, type = "Ljung-Box")
Box.test(residuals(model2), lag = 20, type = "Ljung-Box")
Box.test(residuals(model3), lag = 20, type = "Ljung-Box")
```

```{r}
# Fit the model
fit <- Arima(all_transform$avg_size_log, order = c(5, 1, 1), seasonal = list(order = c(1, 1, 1), period = 5))

# Residuals and Diagnostics
plot(residuals_fit, main = "Residuals")
Acf(residuals_fit, main = "ACF of Residuals")
Pacf(residuals_fit, main = "PACF of Residuals")
qqnorm(residuals_fit, main = "QQ Plot of Residuals")
qqline(residuals_fit)
```



```{r}
fit_auto <- auto.arima(all_transform$avg_size_log, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(fit_auto)
residuals_auto <- residuals(fit_auto)
plot(residuals_auto, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto, main = "ACF of Residuals")
Box.test(residuals_auto, type = "Ljung-Box")
```

```{r}
# Fit SARIMA model
fit_sarima <- arima(all_transform$avg_size_log,
                     order = c(2, 1, 1),            # Non-seasonal orders (p, d, q)
                     seasonal = list(order = c(1, 1, 1), period = 5)) # (P, D, Q)
summary(fit_sarima)
residuals_sarima <- residuals(fit_sarima)
plot(residuals_sarima, main = "Residuals of SARIMA Model") + abline(h = 0, col = "red", lty = 2)
Acf(residuals_sarima, main = "ACF of Residuals")
Pacf(residuals_sarima, main = "PACF of Residuals")
Box.test(residuals_sarima, type = "Ljung-Box")
```


```{r}
# Fit an ARIMA model for each treatment and color
# Create a function to fit ARIMA model for each subset of data
fit_arima <- function(all_transform) {
  # Fit ARIMA model on the log-transformed body size
  arima_model <- Arima(all_transform$avg_size_log, order = c(5, 1, 1), seasonal = list(order = c(1, 1, 1), period = 5))
  return(arima_model)
}

# Apply the function to each combination of treatment and color
models <- all_transform %>%
  group_by(treatment, genet, mhw) %>%
  group_modify(~ {
    model <- fit_arima(.x)
    tibble(model = list(model))
  })

# Check the fitted models
models

# To predict future values, use the fitted model on new data
# Example for one combination of treatment and color
example_model <- models$model[1]
future_predictions <- forecast(example_model, h = 10)  # Forecast 10 periods ahead

# Print the forecasted values
print(future_predictions)
```



