---
title: "Corynactis MHWsim experiment data analysis"
author: "Amelia Ritger"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(ggpmisc) #statpolyeq ggplot
library(forecast) #ARIMA
library(tseries) #ADF test
library(forecast) #ts()
library(ggrepel) #geom_text_repel
```

# Load Tidied Data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
```

# Visualize the data

## Data QA/QC plots
```{r}
ggplot(all, aes(x=date, y=diff_n_cont, color=genet)) +
  geom_point() +
  facet_wrap(~treatment) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal()
```

## Plot basic growth rate across entire experiment
```{r}
ggplot(all, aes(x=date, y=growth_rate_cont, color=genet)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, na.rm=TRUE) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")

#ggsave(here("experiment", "figures", "growth_genet.png"), width=12, height=7)
```

## Plot growth rate of all genets during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~mhw, scales="free", labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Treatment")

#ggsave(here("experiment", "figures", "growth_MHW.png"), width=12, height=7)
```

## Plot growth rate of each genet during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = genet)) +
  geom_point(alpha=0.1) +
  facet_grid(mhw~treatment, 
             #switch = 'x',
             scales = "free_x",
             labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"),
                                 treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  #coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")
  theme(strip.placement = "outside")

#ggsave(here("experiment", "figures", "growth_MHW_genet.png"), width=12, height=10)
```

## Plot growth rate of each genet during and after MHW - but not faceted by MHW
```{r}
label_plot <- c("0.13", "0.09", "0.071", "0.0522", "0.0761",
                "0.151", "0.0933", "0.0252", "0.116", "0.0413", #extreme during
                "0.166", "0.142", "0.0522", "0.134", "0.0902",
                "-0.0023", "0.00741", "0.026", "0.00843", "-0.0131",
                "0.0102", "0.00652", "0.00881", "0.00421", "-0.0117", #extreme after
                "-0.0104", "-0.000895", "0.0217", "0.0127", "-0.0131")
color_plot <- rep(c("#9370DB", "#C21B78", "#FF9933", "#FF3333", "#662B45"), times = 6)
treatment_plot <- rep(c(rep("cold", 5), rep("extreme", 5), rep("severe", 5)), times = 2)
date_plot <- c(rep("2023-10-01", 15), rep("2024-01-10", 15))

p <- ggplot(all, aes(x=date, y=growth_rate_cont, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +   geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
    geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed") +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
  #stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  #un-comment this to get updated values to add to label_plot
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  p <- p + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 25-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
p

#ggsave(here("experiment", "figures", "growth_MHW_genet_combined.png"), width=12, height=7)
```

## Plot raw population numbers
```{r}
g <- ggplot(all, aes(x=date, y=n_true, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +
  geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed", linewidth = 0.5) +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
 # stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  #coord_cartesian(ylim = c(5, 40), expand=TRUE) +
  labs(x = "Date",
       y = "Number of polyps",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  g <- g + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 40-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
g

#ggsave(here("experiment", "figures", "n_MHW_genet_combined.png"), width=12, height=7)
```

## Plot behavior data
```{r}
# %closed facet grid
ggplot(all, aes(x=genet, y=percent_closed, fill=genet, group=genet)) +
  geom_point(aes(color=genet)) +
  geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
  facet_grid(mhw~treatment)

# %closed facet wrap
ggplot(all, aes(x = genet, y = percent_closed, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9)) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.5, outlier.colour = "transparent") +
  facet_wrap(~treatment) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5)) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17))  # Define shapes for mhw levels

# %fullyOpen facet wrap
ggplot(all, aes(x = genet, y = percent_fully_open, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9), alpha = 0.5) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.3, color = alpha("black", 0.75)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5), labels = c("During MHW", "After MHW")) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17), labels = c("During MHW", "After MHW")) + # Define shapes for mhw levels
  # change color of violin 
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Genet",
       y = "% open polyps",
       color = "Genet",
       shape = "MHW",
       size = "MHW") +
  guides(fill = "none",
         color = "none")

#ggsave(here("experiment", "figures", "open_MHW_genet.png"), width=12, height=7)
```

## Plot total biomass over time
```{r}
ggplot(all, aes(x=date, y=total_biomass_true)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="lm", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Total biomass (g)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "total_biomass_mhw.png"), width=12, height=7)
```

## Plot average body size over time
```{r}
ggplot(all, aes(x=date, y=avg_size)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="loess", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Average body size (mm)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "avg_size_mhw.png"), width=12, height=7)
```

## Plot mortality (lol)
```{r}
ggplot(all, aes(x=date, y=dying_dead, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~treatment, 
             labeller = labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00")) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10), labels = function(x) format(x, nsmall = 0)) +
  labs(x = "Date",
       y = "Mortality (no. polyps)",
       color = "Treatment") +
  guides(color = "none")

#ggsave(here("experiment", "figures", "mortality_MHW.png"), width=12, height=7)
```

############################################################################################################

# Start running stats

## Count data - GLARMA 
```{r}

```

## Body size data - Autoregression
### Log transform data
```{r}
all_size <- all %>%
  filter(!is.na(avg_size)) %>%
  mutate(avg_size_log = log(avg_size))

# scatter plot of transformed data
ggplot(all_size, aes(x=date, y=avg_size_log)) +
  geom_point(aes(color=treatment)) +
  labs(x = "Average body size (mm)",
       y = "Frequency")
# histogram of transformed data
ggplot(all_size, aes(x=avg_size_log)) +
  geom_histogram(binwidth=0.1) +
  labs(x = "log(Average body size (mm))",
       y = "Frequency")
# transformed data density plot
ggplot(all_size, aes(x = avg_size_log)) +
  geom_density() +
  ggtitle("Density Plot of Transformed Data")
# Q-Q plot of log-transformed data
qqnorm(all_size$avg_size_log)
qqline(all_size$avg_size_log, col = "red")
```
It's not perfectly normal but definitely better log-transformed. Looks a bit like exponential decay, so let's make the data stationary by differencing.

### Difference to make the data stationary
```{r}
all_size <- all_size %>%
  group_by(tank, genet) %>%
  mutate(avg_size_diff = avg_size_log - lag(avg_size_log),
         avg_size_diff2 = avg_size_diff - lag(avg_size_diff)) #need second order differencing for slope=0

ggplot(all_size, aes(x=date)) +
  geom_point(aes(y=avg_size_diff), color="black", alpha=0.3) +
  geom_point(aes(y=avg_size_diff2), color="blue", alpha=0.3) +
  geom_smooth(aes(y=avg_size_diff), method="lm", color="black") +
  geom_smooth(aes(y=avg_size_diff2), method="lm", color="blue") +
  labs(x = "Average body size (mm)",
       y = "Frequency")
```

### Plot a slightly more complex vis of the differenced data
```{r}
# Fit linear models and extract coefficients
models <- all_size %>%
  group_by(treatment, genet, mhw) %>%
  do(tidy(lm(avg_size_diff ~ date, data = .))) %>%
  select(treatment, genet, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  mutate(equation = sprintf("y = %.2f + %.2e * x", `(Intercept)`, date)) %>%
  rename(slope = date) %>%
  ungroup() %>%
  group_by(mhw) %>%
  mutate(x_axis = as.POSIXct(ifelse(mhw == "during", "2023-10-15", "2024-01-15")),
         y_axis = rep(seq(from = 0.2, to = 0.5, length.out = 5), length.out = n())) %>%
  ungroup() %>%
  select(-mhw)

# Join regression equations back to the original data
all_size_reg <- all_size %>%
  left_join(models, by = c("treatment", "genet"))

ggplot(all_size_reg) +
  geom_point(aes(x=date, y=avg_size_diff, color=genet), alpha=0.3) +
  geom_smooth(aes(x=date, y=avg_size_diff, color=genet ,linetype = mhw), method="lm") +
  facet_wrap(~treatment) +
  geom_label_repel(data = models, aes(x=x_axis, y=y_axis, label = equation, color = genet), size = 3) +
  labs(x = "Date",
       y = "Differenced average body size (mm)") +
  theme_bw()

#ggsave(here("experiment", "figures", "body_size_regression.png"), width=15, height=7)
```
### Run ADF and KPSS tests on first and second differenced data
```{r}
# List to store results
adf_results <- list()

# Loop through each group
for (group in unique(all_size$treatment)) {
  group_data <- all_size %>%
    filter(treatment == group,
           !is.na(avg_size_diff2)) %>%
    pull(avg_size_diff2)
  # Perform the ADF test
  adf_test <- adf.test(group_data, alternative = "stationary")
  # Store the result
  adf_results[[group]] <- adf_test
}

# Print results
adf_results

# List to store results
kpss_results <- list()
# Loop through each group
for (group in unique(all_size$treatment)) {
  group_data <- all_size %>%
    filter(treatment == group,
           !is.na(avg_size_diff2)) %>%
    pull(avg_size_diff2)
  # Perform the KPSS test
  kpss_test <- ur.kpss(group_data, type = "mu")  # Use "tau" if you expect trend stationarity
  # Store the result
  kpss_results[[group]] <- summary(kpss_test)
}

# Print results
kpss_results
```
Both p-value is less than 0.01 and test statistic is lower than critical values, so second-order differenced data is now almost certainly stationary. First-order differencing only showed strong evidence for stationarity with the ADF test.

### Plot ACF and PACF for the log-transformed body size data
```{r}
Acf(all_size$avg_size_log, main = "ACF of Log-transformed Series")
Pacf(all_size$avg_size_log, main = "PACF of Log-transformed Series")

Acf(all_size$avg_size_diff, main = "ACF of Differenced Series")
Pacf(all_size$avg_size_diff, main = "PACF of Differenced Series")

Acf(all_size$avg_size_diff2, main = "ACF of 2-Differenced Series")
Pacf(all_size$avg_size_diff2, main = "PACF of 2-Differenced Series")
```
There is definitely a lot of autocorrelation with the data, for both 1-diff and 2-diff. Makes sense because body size should be highly correlated over time...

### Remove NA values and convert factors to numeric
```{r}
all_size_na <- all_size %>%
  filter(!is.na(avg_size_diff)) %>%
  mutate(genet = as.numeric(genet),
         treatment = as.numeric(treatment),
         mhw = as.numeric(mhw)) %>%
  select(date, tank, genet, n, treatment, mhw, avg_size, avg_size_log, avg_size_diff, avg_size_diff2)
```

### Auto ARIMA - basic
```{r}
autoArima <- auto.arima(all_size_na$avg_size_diff, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(autoArima)
```
I ran this all again with second-differenced data, and the model doesn't appear significantly better (in some cases like BIC, not better) and so it doesn't justify the added complexity of a second-differenced model. so let's stick with first-differenced model.

### Check model residuals
```{r}
residuals_auto <- residuals(autoArima)
plot(residuals_auto, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto, main = "ACF of Residuals")
Box.test(residuals_auto, type = "Ljung-Box")

adf_test_residuals <- adf.test(residuals_auto, alternative = "stationary") #definitely stationary
qqnorm(residuals_auto)
qqline(residuals_auto)
shapiro.test(residuals_auto)

ggplot() +
  geom_histogram(aes(x = residuals_auto), bins = 20, color = "black") +
  labs(x = "Residuals",
       y = "Frequency")

fitted_values <- fitted(autoArima)
plot(fitted_values, residuals_auto, main="Residuals vs Fitted Values")
abline(h=0, col="red")
```
2-order differencing is not significantly better than 1-order differencing so let's stick with 1-order. 

AR(1) model: This is autoregressive data, and the current value of the time series depends linearly on its immediately preceding value - makes sense because body size shouldn't change that drastically week by week. 
I(1): First-order differencing applied to make the series stationary - which makes sense because preliminary analyses showed slope of zero with second-differenced data. 
MA(2): current value of the time series depends on the last two periods’ forecast errors - means that the current value of the series depends on the errors made in the previous two periods, which doesn't make a ton of intuitive sense except maybe if there are big changes to body size (i.e. reproduction?) that just means we won't see those changes finish until 2 weeks later? <<< CHECK THIS LOGIC.

### Check ts trends/seasonality
```{r}
all_size_ts <- ts(all_size_na$avg_size_diff, frequency = 52) #weekly data as ts object
class(all_size_ts)
summary(all_size_ts)
acf(all_size_ts)
pacf(all_size_ts)

decomposition <- stl(all_size_ts, s.window = "periodic")
plot(decomposition)
```
There's no apparent seasonality and a slight positive trend in the data

### Fit ARIMA with different MA values
```{r}
# Fit ARIMA with different seasonal orders
model1 <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 1))
model2 <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2))
model3 <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 3))

Acf(residuals(model1), main = paste("ACF of Residuals model1"))
Pacf(residuals(model1), main = paste("PACF of Residuals model1"))

Acf(residuals(model2), main = paste("ACF of Residuals model2"))
Pacf(residuals(model2), main = paste("PACF of Residuals model2"))

Acf(residuals(model3), main = paste("ACF of Residuals model3"))
Pacf(residuals(model3), main = paste("PACF of Residuals model3"))

Box.test(residuals(model1), lag = 20, type = "Ljung-Box")
Box.test(residuals(model2), lag = 20, type = "Ljung-Box")
Box.test(residuals(model3), lag = 20, type = "Ljung-Box")
```
MA(2) is definitely the best fit. 

### Consider potential covariates
```{r}
#create matrix of dummy variables incorporating all potential covariates
xreg_All <- model.matrix(~ treatment + genet + mhw - 1, data = all_size_na)
xreg_NoGenet <- model.matrix(~ treatment + mhw - 1, data = all_size_na)
xreg_MHW <- model.matrix(~ mhw - 1, data = all_size_na)
xreg_Treatment <- model.matrix(~ treatment - 1, data = all_size_na)

arima_covariates_all <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_All)
summary(arima_covariates_all)

arima_covariates_noGenet <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_NoGenet)
summary(arima_covariates_noGenet)

arima_covariates_mhw <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_MHW)
summary(arima_covariates_mhw)

arima_covariates_treatment <- Arima(all_size_na$avg_size_diff, order = c(1, 1, 2), xreg = xreg_Treatment)
summary(arima_covariates_treatment)
```
Looks like genet is not a significant covariate, but treatment and during/after MHW are slightly impactful. <<< DISCUSS WITH CHRIS.

#### 
Basic ARIMA(1,1,2): 
Coefficients:
         ar1      ma1     ma2
      0.7682  -1.5296  0.5354
s.e.  0.0518   0.0662  0.0646

sigma^2 = 0.002058:  log likelihood = 2258.51
AIC=-4509.02   AICc=-4508.99   BIC=-4488.19

                      ME       RMSE        MAE MPE MAPE      MASE         ACF1
Training set 0.001281871 0.04529954 0.03479757 Inf  Inf 0.7851294 -0.009148644

####
####
ARIMA(1,1,2) with MHW/Treatment covariates:
Coefficients:
         ar1      ma1     ma2  treatment     mhw
      0.7661  -1.5293  0.5351    -0.0029  0.0012
s.e.  0.0539   0.0682  0.0662     0.0025  0.0155

sigma^2 = 0.002059:  log likelihood = 2259.17
AIC=-4506.35   AICc=-4506.28   BIC=-4475.1

                      ME       RMSE        MAE MPE MAPE      MASE         ACF1
Training set 0.001324029 0.04527702 0.03477904 NaN  Inf 0.7847113 -0.008522662
####

### Check model residuals for MHW/Treatment ARIMA
```{r}
residuals_auto <- residuals(arima_covariates_noGenet)
plot(residuals_auto, main = "Residuals of AUTOSARIMA Model")
Acf(residuals_auto, main = "ACF of Residuals")
Box.test(residuals_auto, type = "Ljung-Box")

adf_test_residuals <- adf.test(residuals_auto, alternative = "stationary") #definitely stationary
qqnorm(residuals_auto)
qqline(residuals_auto)
shapiro.test(residuals_auto)

ggplot() +
  geom_histogram(aes(x = residuals_auto), bins = 20, color = "black") +
  labs(x = "Residuals",
       y = "Frequency")

fitted_values <- fitted(arima_covariates_noGenet)
plot(fitted_values, residuals_auto, main="Residuals vs Fitted Values")
abline(h=0, col="red")
```

```{r}
# To predict future values, use the fitted model on new data
# Example for one combination of treatment and color
example_model <- models$model[1]
future_predictions <- forecast(example_model, h = 10)  # Forecast 10 periods ahead

# Print the forecasted values
print(future_predictions)
```



