---
title: "Corynactis MHWsim experiment data analysis"
author: "Amelia Ritger"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(ggpmisc) #statpolyeq ggplot
```

# Load Tidied Data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
```

# Visualize the data

## Data QA/QC plots
```{r}
ggplot(all, aes(x=date, y=diff_n_cont, color=genet)) +
  geom_point() +
  facet_wrap(~treatment) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal()
```

## Plot basic growth rate across entire experiment
```{r}
ggplot(all, aes(x=date, y=growth_rate_cont, color=genet)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, na.rm=TRUE) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")

#ggsave(here("experiment", "figures", "growth_genet.png"), width=12, height=7)
```

## Plot growth rate of all genets during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~mhw, scales="free", labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00"),
                     labels = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW")) +
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Treatment")

#ggsave(here("experiment", "figures", "growth_MHW.png"), width=12, height=7)
```

## Plot growth rate of each genet during and after MHW
```{r}
ggplot(all, aes(x=date, y=growth_rate, color = genet)) +
  geom_point(alpha=0.1) +
  facet_grid(mhw~treatment, 
             #switch = 'x',
             scales = "free_x",
             labeller = labeller(mhw = c("during" = "During MHW", "after" = "After MHW"),
                                 treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE, label.x.npc = "left") +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  #coord_cartesian(ylim = c(-5, 25), expand=TRUE) +   #set equal y axes coordinates
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet")
  theme(strip.placement = "outside")

#ggsave(here("experiment", "figures", "growth_MHW_genet.png"), width=12, height=10)
```

## Plot growth rate of each genet during and after MHW - but not faceted by MHW
```{r}
label_plot <- c("0.13", "0.09", "0.071", "0.0522", "0.0761",
                "0.151", "0.0933", "0.0252", "0.116", "0.0413", #extreme during
                "0.166", "0.142", "0.0522", "0.134", "0.0902",
                "-0.0023", "0.00741", "0.026", "0.00843", "-0.0131",
                "0.0102", "0.00652", "0.00881", "0.00421", "-0.0117", #extreme after
                "-0.0104", "-0.000895", "0.0217", "0.0127", "-0.0131")
color_plot <- rep(c("#9370DB", "#C21B78", "#FF9933", "#FF3333", "#662B45"), times = 6)
treatment_plot <- rep(c(rep("cold", 5), rep("extreme", 5), rep("severe", 5)), times = 2)
date_plot <- c(rep("2023-10-01", 15), rep("2024-01-10", 15))

p <- ggplot(all, aes(x=date, y=growth_rate_cont, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +   geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
    geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = -5, ymax = -4.5), color = "dimgray", fill = "dimgray", alpha = 0.4) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed") +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
  #stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  #un-comment this to get updated values to add to label_plot
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  coord_cartesian(ylim = c(-5, 25), expand=TRUE) +
  labs(x = "Date",
       y = "Cumulative population growth (no. polyps)",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  p <- p + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 25-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
p

#ggsave(here("experiment", "figures", "growth_MHW_genet_combined.png"), width=12, height=7)
```

## Plot raw population numbers
```{r}
g <- ggplot(all, aes(x=date, y=n_true, color=genet, shape=mhw)) +
  geom_point(alpha=0.15) +
  facet_wrap(~factor(treatment, levels=c("cold", "severe", "extreme"), labels=c("Ambient", "Severe MHW", "Extreme MHW"))) +
  geom_rect(aes(xmin = as.POSIXct("2023-11-30"), xmax = as.POSIXct("2023-12-10"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_rect(aes(xmin = as.POSIXct("2024-02-14"), xmax = as.POSIXct("2024-02-21"), ymin = 5, ymax = 5.5), color = "dimgray", fill = "dimgray", alpha = 0.2) +  # Add horizontal bar
  geom_vline(xintercept = as.POSIXct("2023-12-10"), linetype = "dashed", linewidth = 0.5) +  # Add vertical line delineating during/after MHW
  geom_smooth(method = "lm", se = FALSE) +
 # stat_poly_eq(use_label("eq"), formula = y ~ x, parse = TRUE) +  
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_shape_manual(values = c("during" = 16, "after" = 17)) +  # Define shapes for mhw levels
  theme_minimal() +
  #coord_cartesian(ylim = c(5, 40), expand=TRUE) +
  labs(x = "Date",
       y = "Number of polyps",
       color = "Genet") +
  guides(shape = "none")

loop_count = 0
for (i in 1:length(label_plot)) {
  g <- g + geom_text(data = data.frame(date = as.Date("2023-12-30"), treatment = treatment_plot[i], mhw = "during"), 
              label=paste0("m = ", label_plot[i]), x = as.Date(date_plot[i]), y = 40-loop_count, hjust = 0, vjust = 1, color=color_plot[i])
  loop_count <- loop_count + 1
  if (loop_count > 4) {
    loop_count <- 0  # Reset counter after every 5th iteration
  }
}
g

#ggsave(here("experiment", "figures", "n_MHW_genet_combined.png"), width=12, height=7)
```

## Plot behavior data
```{r}
# %closed facet grid
ggplot(all, aes(x=genet, y=percent_closed, fill=genet, group=genet)) +
  geom_point(aes(color=genet)) +
  geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
  facet_grid(mhw~treatment)

# %closed facet wrap
ggplot(all, aes(x = genet, y = percent_closed, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9)) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.5, outlier.colour = "transparent") +
  facet_wrap(~treatment) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5)) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17))  # Define shapes for mhw levels

# %fullyOpen facet wrap
ggplot(all, aes(x = genet, y = percent_fully_open, fill = genet, group = interaction(genet, mhw), shape = mhw)) +
  geom_point(aes(color = genet), position = position_dodge(width = 0.9), alpha = 0.5) +
  geom_violin(aes(size = mhw), position = position_dodge(width = 0.9), alpha = 0.3, color = alpha("black", 0.75)) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_size_manual(values = c("during" = 1, "after" = 0.5), labels = c("During MHW", "After MHW")) + # Define sizes for mhw levels
  scale_shape_manual(values = c("during" = 16, "after" = 17), labels = c("During MHW", "After MHW")) + # Define shapes for mhw levels
  # change color of violin 
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Genet",
       y = "% open polyps",
       color = "Genet",
       shape = "MHW",
       size = "MHW") +
  guides(fill = "none",
         color = "none")

#ggsave(here("experiment", "figures", "open_MHW_genet.png"), width=12, height=7)
```

## Plot total biomass over time
```{r}
ggplot(all, aes(x=date, y=total_biomass_true)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="lm", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Total biomass (g)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "total_biomass_mhw.png"), width=12, height=7)
```

## Plot average body size over time
```{r}
ggplot(all, aes(x=date, y=avg_size)) +
  geom_point(aes(color=genet), size=0.5) +
  geom_smooth(aes(fill=genet, color=genet), method="loess", se=T) +
  facet_wrap(~treatment,
             labeller =  labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  scale_fill_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  scale_color_manual(values = c("A" = "#9370DB", "B" = "#C21B78", "C" = "#FF9933", "D" ="#FF3333", "E" = "#662B45")) +
  theme_minimal() +
  labs(x = "Date",
       y = "Average body size (mm)",
       color = "Genet",
       fill = "Genet")

#ggsave(here("experiment", "figures", "avg_size_mhw.png"), width=12, height=7)
```

## Plot mortality (lol)
```{r}
ggplot(all, aes(x=date, y=dying_dead, color = treatment)) +
  geom_point(alpha=0.1) +
  facet_wrap(~treatment, 
             labeller = labeller(treatment = c("cold" = "Ambient", "severe" = "Severe MHW", "extreme" = "Extreme MHW"))) +
  geom_smooth(method = "lm", formula = y ~ x, se = F) +
  scale_color_manual(values = c("cold" = "#0072B2", "severe" = "#E69F00", "extreme" = "#D55E00")) +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10), labels = function(x) format(x, nsmall = 0)) +
  labs(x = "Date",
       y = "Mortality (no. polyps)",
       color = "Treatment") +
  guides(color = "none")

#ggsave(here("experiment", "figures", "mortality_MHW.png"), width=12, height=7)
```

############################################################################################################

# Start running stats

## Basic ANOVA
```{r}
sample_anova <-aov(data=all,diff_n~genet+treatment)
summary(sample_anova)
TukeyHSD(sample_anova, which="genet")

#ggsave(here("experiment", "figures", "compare.png"), width=12, height=7)
```

```{r}
# run linear regression of growth rate x date for each genet
model <- all %>%
  filter(mhw == "after",
         treatment == "extreme") %>%
  group_by(genet) %>%
  do(model=lm(growth_rate ~ date, data = .))

#extract equation for each genet
eq <- model %>%
  do(eq = paste0("y = ", format(coef(.$model)[1], digits = 2),
                " + ", format(coef(.$model)[2], digits = 2), "x")) %>%
  bind_cols(genet = model$genet) #add back in genet column from model df

all_filter <- all %>%
  select(date, tank, genet, alive, splitting, diff_n_cont)
```