---
title: "MARSS - body size"
author: "Amelia Ritger"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MARSS)
```

# Load data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
#source("readRPiData.R", local = knitr::knit_global()) # Load temperature data AFTER you have uncommented and run this line
weekly_temp <- read_csv(here("experiment", "data", "weekly_temperature.csv"))
```

# Prep data
```{r}
all_size_marss <- all %>%
  filter(!is.na(avg_size)) %>%
  mutate(week = as.numeric(difftime(date, min(weekly_temp$friday), units = "weeks")) + 1,
         avg_size_log = log(avg_size),
         sd_size_log = log(sd_size),
         tank = as.numeric(tank)) %>%
  select(date, week, tank, mhw, genet, avg_size_log) %>%
  left_join(weekly_temp, by = c("date" = "friday", "tank"))

marss_data_wide <- all_size_marss %>%
  mutate(genet_tank = paste(genet, tank, sep = "_")) %>%
  select(-date, -tank, -genet, -mhw, -treatment, -avg_temp, -min_temp, -max_temp) %>%
  pivot_wider(names_from = genet_tank, values_from = avg_size_log) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()
```

# Make matrices

## Make A matrix - genet ID
```{r}
A_matrix <- matrix(rep(c("A", "B", "C", "D", "E"), times = 15), nrow = 75, ncol = 1)
```

## Make C, c matrices - temperature covariate
```{r}
temp_covariate <- weekly_temp %>%
  filter(friday < ymd("2024-02-16")) %>%
  arrange(friday, tank) %>%
  slice(rep(1:n(), each = 5)) %>%
  mutate(tank_rep = paste0(tank, letters[1:5][rep(1:5, times = n()/5)])) %>%
  select(tank_rep, friday, avg_temp) %>%
  pivot_wider(names_from = tank_rep, values_from = avg_temp) %>%
  mutate(week = row_number()) %>%
  select(-friday) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()

C_matrix <- matrix(factor(rep(1:15, each = 5)), ncol = 1)
```

# Make Z matrix list. Might need to incorporate that # cols = # treatments, because then I need to change C matrix dimensions and resolve underconstrained issues with Z, B matrices
```{r}
Z_models <- list(
  H1 = matrix(factor(rep(1, 75))),
  H2 = matrix(factor(c(rep("amb", 25), rep("sev", 25), rep("ext", 25)))),
  H3 = matrix(factor(rep(c("A", "B", "C", "D", "E"), 15))),
  H4 = matrix(factor(c(rep("amb", 25), rep("mhw", 50)))),
  H5 = matrix(factor(c(rep("amb", 25), rep("sev", 25), rep("amb", 25)))), # post-hoc hypothesis
  H6 = matrix(factor(rep(c("gen1", "gen2", "gen2", "gen2", "gen1"), 15))), # post-hoc hypothesis
  H7 = matrix(factor(rep(c("gen1", "gen2", "gen3", "gen4", "gen1"), 15))), # post-hoc hypothesis
  H8 = matrix(factor(c(rep(c("amb_gen1", "amb_gen2", "amb_gen2", "amb_gen2", "amb_gen1"), 5),
                     rep(c("sev_gen1", "sev_gen2", "sev_gen2", "sev_gen2", "sev_gen1"),5),
                     rep(c("amb_gen1", "amb_gen2", "amb_gen2", "amb_gen2", "amb_gen1"), 5))))) # post-hoc hypothesis
  #H9 = matrix(factor(1:75)) #Do not use! Overfits the data and has by far the worst AIC 

names(Z_models) <- c("all_same", "diff_treat", "diff_genet", "mhw_same", "amb_ext_same", "AE_BCD", "AE_B_C_D", "ambExt_AE_BCD") #, "all_diff")
```

H1: All genets and all treatments are the same
H2: All genets are the same but each treatment is different
H3: All treatments are the same but each genet is different
H4: All genets are the same and both MHW treatments are the same, but different from control treatment
H5: All genets are the same and the extreme MHW is the same as the control treatment (post-hoc hypothesis, based off figures)
H6: All treatments are the same, and genets A/E are the same and genets B/C/D are the same (post-hoc hypothesis, based off figures)
H7: All treatments are the same, and genets A/E are the same (post-hoc hypothesis, alternative to H6)
H8: The exterme MHW is the same as the control treatment, and genets A/E are the same and genets B/C/D are the same (post-hoc hypothesis, alternative to H5:H8)
H9: (Do not use!) All genets and all treatments are different

-------------- Notes from WSN 2024:
C matrix - all numeric matrix, 1 column -> This doesn't work though, and needs to be same dimensions as c matrix
c matrix = temp covariate 
B = Identity, random walk
Z = CUSTOM matrix, the hypotheses
A = A_matrix?
R = diagonal and equal #identity?
U = unequal?
Q = diagonal and equal?

# Set up model list
```{r}
mod_list <- list(
  B = "identity", #pretty sure this is right, random walk diagonal and unequal has worse AIC/BIC
  U = "zero", #unequal has worse AIC/BIC
  Q = "diagonal and equal", #All tanks have same mixing; random walk diagonal and unequal has worse AIC/BIC
  Z = "placeholder",
  A = A_matrix,
  R = "diagonal and equal", #errors are independent and equally distributed
  C = C_matrix,
  c = temp_covariate)
```

# Loop through Z matrix models 
```{r}
out.tab <- NULL
fits <- list()
for (i in 1:length(Z_models)) {
    mod_list$Z <- Z_models[[i]]
    fit <- MARSS(y=marss_data_wide, model = mod_list, silent = TRUE, 
        control = list(maxit = 5000))
    out <- data.frame(H = names(Z_models)[i], logLik = fit$logLik, 
        AICc = fit$AICc, num.param = fit$num.params, m = length(unique(Z_models[[i]])), 
        num.iter = fit$numIter, converged = !fit$convergence)
    out.tab <- rbind(out.tab, out)
    fits <- c(fits, list(fit))
}
```

## Experiment with non 75x1 Z matrix dimensions
```{r}
#Each treatment different
Z_matrix2 <- matrix(factor(c(rep(c(1, 0, 0), 25),
                            rep(c(0, 1, 0), 25),
                            rep(c(0, 0, 1), 25))),
                   nrow = 75, byrow = TRUE)

# Combine the blocks into a single 75x3 matrix
matrix_data <- rbind(matrix(c(rep(1:5, each = 5), rep(0, 25), rep(0, 25)), nrow = 25, ncol = 3, byrow = FALSE),
                     matrix(c(rep(0, 25), rep(6:10, each = 5), rep(0, 25)), nrow = 25, ncol = 3, byrow = FALSE),
                     matrix(c(rep(0, 25), rep(0, 25), rep(10:14, each = 5)), nrow = 25, ncol = 3, byrow = FALSE))

C_matrix2 <- matrix(factor(rep(1:15, each = 5)), ncol = 3)

C_matrix2 <- matrix(0, nrow = 75, ncol = 3)
for (i in 1:3) {
  C_matrix2[((i-1)*5 + 1):(i*5), i] <- 1
}
C_matrix2 <- matrix(factor(C_matrix2), nrow = 75, ncol = 3)

model_list <- list(
  B = "identity",
  U = "zero",
  Q = "diagonal and equal",
  Z = Z_matrix2,
  A = A_matrix,
  R = "diagonal and equal",
  C = matrix_data,
  c = temp_covariate)

# non 75x1 Z matrix Does not run 
#fit1 <- MARSS(y=marss_data_wide, model = model_list,
#              control = list(maxit = 1000))
```

# Generate the z-scored data
```{r}
the.mean <- apply(marss_data_wide, 1, mean)
the.sigma <- sqrt(apply(marss_data_wide, 1, var))
marss_data_wide_z <- (marss_data_wide - the.mean) * (1/the.sigma)

the.mean <- apply(temp_covariate, 1, mean)
the.sigma <- sqrt(apply(temp_covariate, 1, var))
temp_covariate_z <- (temp_covariate - the.mean) * (1/the.sigma)
```

## Model the z-scored data
```{r}
mod_list_z <- list(
  B = "diagonal and equal", #identity is another option, but models have lower logLik and higher AICc; random walk diagonal and unequal gives same model output
  U = "zero", #unequal doesn't converge even after 5k iterations
  Q = "diagonal and equal", #All tanks have same mixing (also, random walk diagonal and unequal is the same AICc and very close Q estimates)
  Z = "placeholder",
  A = "zero", #because data have been z-scored
  R = "diagonal and equal", #observation errors are independent and have equal variances, because there was only ONE sampler (me)
  C = C_matrix,
  c = temp_covariate_z)

out_tab_z <- NULL
fits_z <- list()
for (i in 1:length(Z_models)) {
    mod_list_z$Z <- Z_models[[i]]
    fit <- MARSS(y=marss_data_wide_z, model = mod_list_z, silent = TRUE, 
        control = list(maxit = 5000))
    #print(fit)
    out <- data.frame(H = names(Z_models)[i], logLik = fit$logLik, 
        AICc = fit$AICc, num.param = fit$num.params, m = length(unique(Z_models[[i]])), 
        num.iter = fit$numIter, converged = !fit$convergence)
    out_tab_z <- rbind(out_tab_z, out)
    fits_z <- c(fits_z, list(fit))
}
```

According to AICc (and logLik), definitely don't use "all_diff" and can probably also rule out "diff_genet"

## Visualize z-scored temperatures
```{r}
temp_z <- weekly_temp %>%
  filter(friday < ymd("2024-02-16")) %>%
  arrange(friday, tank) %>%
  select(tank, friday, avg_temp) %>%
  group_by(tank) %>%
  mutate(avg_temp_z = (avg_temp - mean(avg_temp, na.rm = TRUE)) /sd(avg_temp, na.rm = TRUE),
         treatment = ifelse(tank < 6, "ambient",
                            ifelse(tank > 10, "extreme",
                                   "severe")),
         tank = as.factor(tank)) %>%
  ungroup()

ggplot(temp_z, aes(x = friday, y = avg_temp_z, color = tank)) +
  geom_line() +
  labs(title = "Z-scored average temperature by tank",
       x = "Date",
       y = "Z-scored average temperature") +
  theme_minimal() +
  facet_wrap(~treatment)
```

