---
title: "MARSS - body size"
author: "Amelia Ritger"
date: "2024-11-04"
output: html_document
---

*Important note*: all current code only tests the "during MHW" period, and NOT the recovery period. But recovery period should also be tested because it is not just how the animals responded to the HW, but also how they recovered (and even how quickly). MHW exposure ended on Dec 12 2023. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MARSS)
```

# Load data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
#source("readRPiData.R", local = knitr::knit_global()) # Load temperature data AFTER you have uncommented and run this line
weekly_temp <- read_csv(here("experiment", "data", "weekly_temperature.csv"))
daily_temp <- read_csv(here("experiment", "data", "three_day_temperature.csv"))

end_mhw <- ymd("2023-12-12")
U <- c("unequal", "equal", "zero")
```

# Functions
```{r}
# For plotting parameters
plot.pars <- function(fit) {
    df <- broom::tidy(fit)
    df$parameter <- sapply(df$term, function(x) {
        stringr::str_split(x, "[.]")[[1]][1]
    })
    df$parameter[df$parameter %in% c("R", "Q")] <- "R and Q"
    ggplot2::ggplot(df, ggplot2::aes(x = term, y = estimate)) + 
        ggplot2::geom_point() + ggplot2::geom_errorbar(ggplot2::aes(ymin = conf.low, 
        ymax = conf.up), width = 0.2) + geom_hline(yintercept = 0, 
        col = "blue") + facet_wrap(~parameter, scales = "free")
}

# For generating H3/H5 Z matrices - staggered 1 values across specific rows and columns
fill_staggered_ones <- function(Z, rows, cols) { #Fill staggered 1s across specified rows/cols
  rows_to_apply <- intersect(rows, 1:nrow(Z))   # Ensure the row numbers are within the bounds of the matrix
  for (col in cols) { # Loop over the specified columns
    start_row <- col  # Stagger the starting row for each column
    for (i in seq(start_row, max(rows_to_apply), by = 5)) {
      if (i %in% rows_to_apply) {
        Z[i, col] <- 1
      }
    }
  }
  return(Z)
}
```

# Hypotheses
- H1: All genets and all treatments are the same
- H2: All genets are the same but each treatment is different
- H3: All treatments are the same but each genet is different
- H4: All genets are the same and both MHW treatments are the same, but different from control treatment
- H5: All genets are the same and the extreme MHW is the same as the control treatment (post-hoc hypothesis, - based off figures)
- H6: All treatments are the same, and genets A/E are the same and genets B/C/D are the same (post-hoc - hypothesis, based off figures)
- H7: All treatments are the same, and genets A/E are the same (post-hoc hypothesis, alternative to H6)
- H8: The extreme MHW is the same as the control treatment, and genets A/E are the same and genets B/C/D are the same (post-hoc hypothesis, alternative to H5:H7)
- H9: All genets and all treatments are different
- H10 (**not currently tested**): The heatwave period was different from the recovery period

**Q is diagonal and equal, especially because major model convergence issues when it's set to diagonal and unequal**

# Generate the hypotheses matrices
```{r}
zc_matrix_list <- list(
  H1 = list(Z = matrix(1, nrow = 75, ncol = 1), C = matrix(list(1), 1, 15)), #all genets and all treatments are the same
  H2 = list(Z = matrix(0, nrow = 75, ncol = 3), C = matrix(list(0), 3, 15)), #all genets are the same, all treatments are different
  H3 = list(Z = matrix(0, nrow = 75, ncol = 5), C = matrix(list(0), 5, 15)), #all genets are different, all treatments are different
  H4 = list(Z = matrix(0, nrow = 75, ncol = 2), C = matrix(list(0), 2, 15)), #all genets are the same, MHW treatments are different from control
  H9 = list(Z = matrix(0, nrow = 75, ncol = 15), C = matrix(list(0), 15, 15)), #all genets and all treatments are different 
  
  ## post hoc hypotheses:
  H5 = list(Z = matrix(0, nrow = 75, ncol = 2), C = matrix(list(0), 2, 15)), #all genets are the same and the extreme MHW is the same as the control treatment
  H6 = list(Z = matrix(0, nrow = 75, ncol = 2), C = matrix(list(0), 2, 15)), #all treatments are the same, and genets A/E are the same and genets B/C/D are the same 
  H7 = list(Z = matrix(0, nrow = 75, ncol = 4), C = matrix(list(0), 4, 15)), #all treatments are the same, and genets A/E are the same 
  H8 = list(Z = matrix(0, nrow = 75, ncol = 4), C = matrix(list(0), 4, 15)) #the extreme MHW is the same as the control treatment, and genets A/E are the same and genets B/C/D are the same
)

## Hypothesis 1: All genets and all treatments are the same
colnames(zc_matrix_list$H1$Z) <- c("all same")

# H2: All treatments are different -> Z (75x3), C (3x15)
zc_matrix_list$H2$Z[1:25, 1] <- 1  # Ambient
zc_matrix_list$H2$Z[26:50, 2] <- 1  # Severe
zc_matrix_list$H2$Z[51:75, 3] <- 1  # Extreme
colnames(zc_matrix_list$H2$Z) <- c("ambient", "severe", "extreme")
zc_matrix_list$H2$C[1, 1:5] <- "ambient"
zc_matrix_list$H2$C[2, 6:10] <- "severe"
zc_matrix_list$H2$C[3, 11:15] <- "extreme"

# H3: All genets are different -> Z (75x5), C (5x15)
zc_matrix_list$H3$Z <- fill_staggered_ones(zc_matrix_list$H3$Z, 1:75, 1:5)
colnames(zc_matrix_list$H3$Z) <- c("genetA", "genetB", "genetC", "genetD", "genetE")
for (i in 1:nrow(zc_matrix_list$H3$C)) {
  zc_matrix_list$H3$C[i, ] <- rep(letters[i], ncol(zc_matrix_list$H3$C))
}

# H4: MHW treatments are different from control -> Z (75x2), C (2x15)
zc_matrix_list$H4$Z[1:25, 1] <- 1  # Treatment 1 (ambient)
zc_matrix_list$H4$Z[26:75, 2] <- 1  # Treatment 2 (severe and extreme)
colnames(zc_matrix_list$H4$Z) <- c("ambient", "mhw")
zc_matrix_list$H4$C[1, 1:5] <- "ambient"
zc_matrix_list$H4$C[2, 6:15] <- "mhw"

# H9: All genets and all treatments are different -> Z (75x15), C (15x15)
zc_matrix_list$H9$Z <- fill_staggered_ones(zc_matrix_list$H9$Z,1:25, 1:5)
zc_matrix_list$H9$Z <- fill_staggered_ones(zc_matrix_list$H9$Z,26:50, 6:10)
zc_matrix_list$H9$Z <- fill_staggered_ones(zc_matrix_list$H9$Z,51:75, 11:15)
colnames(zc_matrix_list$H9$Z) <- c("amb_A", "amb_B", "amb_C", "amb_D", "amb_E",
                                   "sev_A", "sev_B", "sev_C", "sev_D", "sev_E",
                                   "ext_A", "ext_B", "ext_C", "ext_D", "ext_E")
for (i in 1:nrow(zc_matrix_list$H9$C)) {
  letter_index <- (i - 1) %% 5 + 1  # This ensures we loop through 'a' to 'e'
  letter <- letters[letter_index]  # Get the current genet
    if (i >= 1 & i <= 5) {
    suffix <- "ambient_"
  } else if (i >= 6 & i <= 10) {
    suffix <- "severe_"
  } else if (i >= 11 & i <= 15) {
    suffix <- "extreme_"
  } else {
    suffix <- ""
  }
  zc_matrix_list$H9$C[i, ] <- rep(paste(suffix, letter, sep = ""), 15)
}

# H5: All genets are the same and the extreme MHW is the same as the control treatment (post-hoc hypothesis, based off figures)
colnames(zc_matrix_list$H5$Z) <- c("amb_ext", "severe")
zc_matrix_list$H5$Z[1:25,  1] <- 1  # ambient
zc_matrix_list$H5$Z[26:50, 2] <- 1  # severe
zc_matrix_list$H5$Z[51:75, 1] <- 1  # extreme -> same as ambient
# C: cols 1-5 = ambient, 6-10 = severe, 11-15 = extreme
zc_matrix_list$H5$C[1, c(1:5, 11:15)] <- "amb_ext"  # ambient and extreme cols share label
zc_matrix_list$H5$C[2, 6:10]          <- "severe"

# H6: All treatments are the same, and genets A/E are the same and genets B/C/D are the same (post-hoc hypothesis, based off figures)
colnames(zc_matrix_list$H6$Z) <- c("genetAE", "genetBCD")
# For each treatment, A/E -> state 1 and B/C/D -> state 2
ae_rows  <- c(seq(1, 25, by=5), seq(5, 25, by=5),   # A and E in ambient
              seq(26, 50, by=5), seq(30, 50, by=5),  # A and E in severe
              seq(51, 75, by=5), seq(55, 75, by=5))  # A and E in extreme
bcd_rows <- setdiff(1:75, ae_rows)
zc_matrix_list$H6$Z[ae_rows,  1] <- 1
zc_matrix_list$H6$Z[bcd_rows, 2] <- 1
# C: treatments are all the same
zc_matrix_list$H6$C[1, ] <- "genetAE"
zc_matrix_list$H6$C[2, ] <- "genetBCD"

# H7: All treatments are the same, and genets A/E are the same (post-hoc hypothesis, alternative to H6)
colnames(zc_matrix_list$H7$Z) <- c("genetAE", "genetB", "genetC", "genetD")
ae_rows <- c(seq(1, 75, by=5), seq(5, 75, by=5))
b_rows  <- seq(2, 75, by=5)
c_rows  <- seq(3, 75, by=5)
d_rows  <- seq(4, 75, by=5)
zc_matrix_list$H7$Z[ae_rows, 1] <- 1
zc_matrix_list$H7$Z[b_rows,  2] <- 1
zc_matrix_list$H7$Z[c_rows,  3] <- 1
zc_matrix_list$H7$Z[d_rows,  4] <- 1
zc_matrix_list$H7$C[1, ] <- "genetAE"
zc_matrix_list$H7$C[2, ] <- "genetB"
zc_matrix_list$H7$C[3, ] <- "genetC"
zc_matrix_list$H7$C[4, ] <- "genetD"

# H8: The extreme MHW is the same as the control treatment, and genets A/E are the same and genets B/C/D are the same
colnames(zc_matrix_list$H8$Z) <- c("ambext_AE", "ambext_BCD", "sev_AE", "sev_BCD")

ae_amb  <- c(seq(1, 25, by=5), seq(5, 25, by=5))   # A and E in ambient (rows 1-25)
ae_sev  <- c(seq(26, 50, by=5), seq(30, 50, by=5)) # A and E in severe (rows 26-50)
ae_ext  <- c(seq(51, 75, by=5), seq(55, 75, by=5)) # A and E in extreme (rows 51-75)
amb_ext_ae  <- c(ae_amb, ae_ext)
amb_ext_bcd <- setdiff(c(1:25, 51:75), amb_ext_ae)
sev_ae      <- ae_sev
sev_bcd     <- setdiff(26:50, sev_ae)

zc_matrix_list$H8$Z[amb_ext_ae, 1] <- 1
zc_matrix_list$H8$Z[amb_ext_bcd, 2] <- 1
zc_matrix_list$H8$Z[sev_ae, 3] <- 1
zc_matrix_list$H8$Z[sev_bcd, 4] <- 1

# C cols 1-5 = ambient, 6-10 = severe, 11-15 = extreme
zc_matrix_list$H8$C[1, c(1:5, 11:15)] <- "ambext_AE"
zc_matrix_list$H8$C[2, c(1:5, 11:15)] <- "ambext_BCD"
zc_matrix_list$H8$C[3, 6:10]          <- "sev_AE"
zc_matrix_list$H8$C[4, 6:10]          <- "sev_BCD"
```

# Body size 

## Generate temp covariate matrix - ONLY for MHW times
```{r}
temp_covariate <- weekly_temp %>%
  filter(friday < ymd("2024-02-16"),
         friday <= end_mhw) %>%
  arrange(friday, tank) %>%
  #slice(rep(1:n(), each = 5)) %>%
  #mutate(tank_rep = paste0(tank, letters[1:5][rep(1:5, times = n()/5)])) %>%
  select(tank, friday, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(week = row_number()) %>%
  select(-friday) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()
```

## Prep body size data
```{r}
all_size <- all %>%
  filter(!is.na(avg_size),
         date <= end_mhw) %>%
  mutate(week = as.numeric(difftime(date, min(weekly_temp$friday), units = "weeks")) + 1,
         avg_size_log = log(avg_size),
         sd_size_log = log(sd_size),
         tank = as.numeric(tank)) %>%
  select(date, week, tank, mhw, genet, avg_size_log) %>%
  left_join(weekly_temp, by = c("date" = "friday", "tank"))

all_size_marss <- all_size %>%
  mutate(genet_tank = paste(genet, tank, sep = "_")) %>%
  select(-date, -tank, -genet, -mhw, -treatment, -avg_temp, -min_temp, -max_temp) %>%
  pivot_wider(names_from = genet_tank, values_from = avg_size_log) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()

marss_data <- all_size_marss
```

## Hypothesis 2: All genets are the same but each treatment is different 
```{r, eval=FALSE}
Z <- matrix(0, nrow = 75, ncol = 3) # Fill in the matrix # First 25 rows get a 1 in column 1 (treatment 1) 
Z[1:25, 1] <- 1 # Next 25 rows (26-50) get a 1 in column 2 (treatment 2) 
Z[26:50, 2] <- 1 # Last 25 rows (51-75) get a 1 in column 3 (treatment 3) 
Z[51:75, 3] <- 1 # Optional: Name the columns for clarity 
colnames(Z) <- c("treatment 1", "treatment 2", "treatment 3")

C = matrix(list(0), 3, 15)
C[1, 1:5] <- "treatment1"
C[2, 6:10] <- "treatment2"
C[3, 11:15] <- "treatment3"

#resids <- MARSSresiduals(fit_indv_alt, type = "tt1")
#for (j in 1:75) {
#    plot(resids$model.residuals[j, ], ylab = "model residuals", 
#        xlab = "")
#    abline(h = 0)
#    title(paste(rownames(marss_data)[j]))
#    }


####################### for loop running through options of U:

U <- c("unequal", "equal", "zero")

# Create an empty data frame to store the results
results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  U = character(),  # Add model parameter labels
  stringsAsFactors = FALSE
)

# Loop over each combination of parameters
for (i in 1:length(U)) {
  # Extract the current combination of parameters
  current_u <- U[i]
  
  # Create the mod_list with current parameter values
  mod_list_loop <- list(
    B = "identity",
    U = current_u,
    Q = "diagonal and equal",
    Z = Z,
    A = "zero",  # DO NOT CHANGE!
    R = "diagonal and equal",  # DO NOT CHANGE!
    C = C,
    c = temp_covariate,
    tinitx = 0)

    # Fit the MARSS model
  fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
  
  # Extract relevant output
  out_indv <- data.frame(
    U = current_u,
    logLik = fit_marss$logLik, 
    AICc = fit_marss$AICc, 
    num.param = fit_marss$num.params, 
    num.iter = fit_marss$numIter, 
    converged = !fit_marss$convergence
  )
  
  #MARSSparamCIs(fit_marss)
  #plot.pars(fit_marss)
  
  # Append the results to the results data frame
  results <- rbind(results, out_indv)
}

results <- results %>%
  mutate(hypothesis = "H2")

# save results as .csv
#write_csv(results, here("experiment", "data", "marss_output", "H2.csv"))
```

Unequal U is best fit.

## For loop to test each U with each hypothesis (Z/C matrix)
```{r}
marss_results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  U = character(),  # Add model parameter labels
  Z_group = character(), C_group = character(),  # To track Z and C combinations
  stringsAsFactors = FALSE
)

# Loop over each combination of Z and C
for (hypothesis_number in names(zc_matrix_list)) {
  # Extract the current Z and C combination
  Z <- zc_matrix_list[[hypothesis_number]]$Z
  C <- zc_matrix_list[[hypothesis_number]]$C
  
  # Loop over each combination of U
  for (i in 1:length(U)) {
    # Extract the current combination of parameters
    current_u <- U[i]
    
    # Create the mod_list with current parameter values
    mod_list_loop <- list(
      B = "identity",
      U = current_u,
      Q = "diagonal and equal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0
    )
    
    # Fit the MARSS model
    fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
    
    # Extract relevant output
    out_indv <- data.frame(
      Hypothesis = hypothesis_number, U = current_u,
      logLik = fit_marss$logLik, 
      AICc = fit_marss$AICc, 
      num.param = fit_marss$num.params, 
      num.iter = fit_marss$numIter, 
      converged = !fit_marss$convergence
    )
    
    # Append the results to the results data frame
    marss_results <- rbind(marss_results, out_indv)
  }
}

marss_results %>%
  arrange(AICc)

marss_results_size <- marss_results
```

H9 U=equal is best fit. 

## The following model diagnostics, I'm not sure what to look for

### Check out the best performing model for H9 (U=equal)
```{r, eval=FALSE}
Z <- matrix(0, nrow = 75, ncol = 15)
C <- matrix(list(0), 15, 15)
  
Z <- fill_staggered_ones(Z,1:25, 1:5)
Z <- fill_staggered_ones(Z,26:50, 6:10)
Z <- fill_staggered_ones(Z,51:75, 11:15)
colnames(zc_matrix_list$H9$Z) <- c("amb_A", "amb_B", "amb_C", "amb_D", "amb_E",
                                   "sev_A", "sev_B", "sev_C", "sev_D", "sev_E",
                                   "ext_A", "ext_B", "ext_C", "ext_D", "ext_E")

for (i in 1:nrow(C)) {
  letter_index <- (i - 1) %% 5 + 1  # This ensures we loop through 'a' to 'e'
  letter <- letters[letter_index]  # Get the current genet
    if (i >= 1 & i <= 5) {
    suffix <- "ambient_"
  } else if (i >= 6 & i <= 10) {
    suffix <- "severe_"
  } else if (i >= 11 & i <= 15) {
    suffix <- "extreme_"
  } else {
    suffix <- ""
  }
  C[i, ] <- rep(paste(suffix, letter, sep = ""), 15)
}

model_list <- list(
      B = "identity",
      U = "equal",
      Q = "diagonal and equal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0)

fit_marss_indv <- MARSS(y = marss_data, model = model_list, silent = TRUE, control = list(maxit = 5000))
fit_marss_indv
MARSSparamCIs(fit_marss_indv)
plot.pars(fit_marss_indv)

Q.est <- coef(fit_marss_indv, type = "matrix")$Q
corrplot::corrplot(cov2cor(Q.est))

resids <- MARSSresiduals(fit_marss_indv, type = "tt1")
for (j in 1:75) {
    plot(resids$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(rownames(marss_data)[j]))
}

boot_results <- MARSSboot(fit_marss_indv, nboot = 5, silent = TRUE)
```

```{r, eval=FALSE}
# Convert boot.params to matrix if not already
boot_mat <- as.matrix(boot_results$boot.params)

# Compute summary stats
param_names <- rownames(boot_mat)
point_estimates <- apply(boot_mat, 1, mean)
lower_ci <- apply(boot_mat, 1, quantile, probs = 0.025)
upper_ci <- apply(boot_mat, 1, quantile, probs = 0.975)

# Combine into a dataframe for plotting
plot_df <- data.frame(
  Parameter = param_names,
  Estimate = point_estimates,
  CI_lower = lower_ci,
  CI_upper = upper_ci
)

# Optional: group into categories like environmental, density-dependence, etc.
plot_df$Group <- ifelse(grepl("ambient|severe|extreme", plot_df$Parameter), "Environmental",
                 ifelse(grepl("U\\.1", plot_df$Parameter), "Density", "Other"))

ggplot(plot_df, aes(x = reorder(Parameter, Estimate), y = Estimate, color = Group)) +
  geom_point() +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  theme_minimal() +
  labs(x = "Parameter", y = "Estimate ± 95% CI", title = "Bootstrap Estimates with 95% CIs") +
  facet_wrap(~Group, scales = "free_y")
```

# Population growth

## Generate temp covariate matrix - ONLY for MHW times
```{r}
temp_covariate <- daily_temp %>%
  filter(date >= ymd("2023-10-06"),
         date < ymd("2024-02-16"),
         date <= end_mhw) %>%
  arrange(date, tank) %>%
  mutate(avg_temp = avg_temp_last_3_days) %>%
  select(tank, date, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(day = row_number()) %>%
  select(-date) %>%
  column_to_rownames(var = "day") %>%
  as.matrix() %>%
  t()
```

## Prep reproduction data
**changed n_true to n_rel (aka relative to starting number) because that's what is being used for SEM and other analyses** I z-score those data for SEM, do I need to do that here as well?
```{r}
all_repro <- all %>%
  filter(date >= ymd("2023-10-06"),
                    date <= end_mhw) %>%
  mutate(#week = as.numeric(difftime(date, min(weekly_temp$friday), units = "weeks")) + 1,
         #avg_size_log = log(avg_size),
         #sd_size_log = log(sd_size),
          tank = as.numeric(tank),
          tank_genet = paste(tank, genet, sep = "_"),
          date_num = as.numeric(date - as.Date("2023-10-01"))) %>%
  arrange(date_num) %>%
  group_by(tank_genet) %>%
  mutate(n_rel = n_true / first(n_true)) %>% # relative to starting number
  ungroup() %>%
  select(date, tank, mhw, genet, n_rel) %>%
  left_join(daily_temp, by = c("date", "tank"))

all_repro_marss <- all_repro %>%
  mutate(genet_tank = paste(genet, tank, sep = "_")) %>%
  select(date, genet_tank, n_rel) %>%
  #select(-date, -tank, -genet, -mhw, -treatment, -avg_temp, -min_temp, -max_temp) %>%
  pivot_wider(names_from = genet_tank, values_from = n_rel) %>%
  column_to_rownames(var = "date") %>%
  as.matrix() %>%
  t()

marss_data <- all_repro_marss
```

## Run the for loop to test each U with each hypothesis (Z/C matrix)
```{r}
marss_results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  U = character(),  # Add model parameter labels
  Z_group = character(), C_group = character(),  # To track Z and C combinations
  stringsAsFactors = FALSE
)

# Loop over each combination of Z and C
for (hypothesis_number in names(zc_matrix_list)) {
  # Extract the current Z and C combination
  Z <- zc_matrix_list[[hypothesis_number]]$Z
  C <- zc_matrix_list[[hypothesis_number]]$C
  
  # Loop over each combination of U
  for (i in 1:length(U)) {
    # Extract the current combination of parameters
    current_u <- U[i]
    
    # Create the mod_list with current parameter values
    mod_list_loop <- list(
      B = "identity",
      U = current_u,
      Q = "diagonal and equal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0
    )
    
    # Fit the MARSS model
    fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
    
    # Extract relevant output
    out_indv <- data.frame(
      Hypothesis = hypothesis_number, U = current_u,
      logLik = fit_marss$logLik, 
      AICc = fit_marss$AICc, 
      num.param = fit_marss$num.params, 
      num.iter = fit_marss$numIter, 
      converged = !fit_marss$convergence
    )
    
    # Append the results to the results data frame
    marss_results <- rbind(marss_results, out_indv)
  }
}

marss_results %>%
  arrange(AICc)

marss_results_reproduction <- marss_results
```

H9 U=equal is best model, and all converged.

# Behavior 

## Generate temp covariate matrix - ONLY for MHW times
```{r}
temp_covariate <- daily_temp %>%
  filter(date >= ymd("2023-10-06"),
         date < ymd("2024-02-16"),
         date <= end_mhw) %>%
  arrange(date, tank) %>%
  mutate(avg_temp = avg_temp_last_3_days) %>%
  select(tank, date, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(day = row_number()) %>%
  select(-date) %>%
  column_to_rownames(var = "day") %>%
  as.matrix() %>%
  t()
```

## Prep behavior data
**Check comment below regarding disparity in how I'm handling behavioral data between all analyses. For SEM I did scale(percent_fully_open)[, 1] and for GLMM I did ifelse(df$percent_fully_open == 1, df$percent_fully_open - 0.001, df$percent_fully_open).**

```{r}
all_behavior <- all %>%
  filter(date >= ymd("2023-10-06"),
                    date <= end_mhw) %>%
  mutate(tank = as.numeric(tank)) %>%
  select(date, tank, mhw, genet, percent_fully_open) %>%
  left_join(daily_temp, by = c("date", "tank"))

all_behavior_marss <- all_behavior %>%
  mutate(genet_tank = paste(genet, tank, sep = "_"),
         number_open = percent_fully_open*100,
         # Everything below is not exactly what I've done for other analyses.
         data_squished = pmin(pmax(percent_fully_open, 0.0001), 1 - 0.0001),
         transform_percent = log(data_squished / (1 - data_squished)),
         proportion_open_adjusted = case_when(
           percent_fully_open == 0 ~ (0.5 / n()), # n() is the number of observations
           percent_fully_open == 1 ~ (1 - (0.5 / n())),
           TRUE ~ percent_fully_open),
         logit_percent_open = log(proportion_open_adjusted / (1 - proportion_open_adjusted)),
         arcsin_sqrt_percent_open = asin(sqrt(percent_fully_open))) %>%
  select(date, genet_tank, arcsin_sqrt_percent_open) %>%
  pivot_wider(names_from = genet_tank, values_from = arcsin_sqrt_percent_open) %>%
  column_to_rownames(var = "date") %>%
  as.matrix() %>%
  t()

marss_data <- all_behavior_marss
```

### Z-score the data (maybe? but currently not used in the next chunk.)
```{r, eval=FALSE}
the.mean <- apply(marss_data, 1, mean)
the.sigma <- sqrt(apply(marss_data, 1, var))
marss_data_z <- (marss_data - the.mean) * (1/the.sigma)

the.mean <- apply(temp_covariate, 1, mean)
the.sigma <- sqrt(apply(temp_covariate, 1, var))
temp_covariate_z <- (temp_covariate - the.mean) * (1/the.sigma)
```

## Run the for loop to test each U with each hypothesis (Z/C matrix)
```{r}
marss_results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  U = character(),  # Add model parameter labels
  Z_group = character(), C_group = character(),  # To track Z and C combinations
  stringsAsFactors = FALSE
)

# Loop over each combination of Z and C
for (hypothesis_number in names(zc_matrix_list)) {
  # Extract the current Z and C combination
  Z <- zc_matrix_list[[hypothesis_number]]$Z
  C <- zc_matrix_list[[hypothesis_number]]$C
  
  # Loop over each combination of U
  for (i in 1:length(U)) {
    # Extract the current combination of parameters
    current_u <- U[i]
    
    # Create the mod_list with current parameter values
    mod_list_loop <- list(
      B = "identity",
      U = current_u,
      Q = "diagonal and equal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0
    )
    
    # Fit the MARSS model
    fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
    
    # Extract relevant output
    out_indv <- data.frame(
      Hypothesis = hypothesis_number, U = current_u,
      logLik = fit_marss$logLik, 
      AICc = fit_marss$AICc, 
      num.param = fit_marss$num.params, 
      num.iter = fit_marss$numIter, 
      converged = !fit_marss$convergence
    )
    
    # Append the results to the results data frame
    marss_results <- rbind(marss_results, out_indv)
  }
}

marss_results %>%
  arrange(AICc)

marss_results_behavior <- marss_results
```

H9 U=equal is best fit model.

--------------------------------------------------------------------------------

# Everything below this is "old" (but reference it if needed)

# Make matrices

## Make C, c matrices - temperature covariate (15 tanks, 5 genets in each tank)
```{r, eval=FALSE}
temp_covariate <- weekly_temp %>%
  filter(friday < ymd("2024-02-16")) %>%
  arrange(friday, tank) %>%
  #slice(rep(1:n(), each = 5)) %>%
  #mutate(tank_rep = paste0(tank, letters[1:5][rep(1:5, times = n()/5)])) %>%
  select(tank, friday, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(week = row_number()) %>%
  select(-friday) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()

C_matrix <- matrix(factor(paste("tank", rep(1:15, each = 1), sep = "")), ncol = 1)
#3x15 matrix
```

## Make Z matrix list. Might need to incorporate that # cols = # treatments, because then I need to change C matrix dimensions and resolve underconstrained issues with Z, B matrices
```{r, eval=FALSE}
Z_models <- list(
  H1 = matrix(factor(rep(1, 75))),
  H2 = matrix(factor(c(rep("amb", 25), rep("sev", 25), rep("ext", 25)))),
  H3 = matrix(factor(rep(c("A", "B", "C", "D", "E"), 15))),
  H4 = matrix(factor(c(rep("amb", 25), rep("mhw", 50)))),
  H5 = matrix(factor(c(rep("amb", 25), rep("sev", 25), rep("amb", 25)))), # post-hoc hypothesis
  H6 = matrix(factor(rep(c("gen1", "gen2", "gen2", "gen2", "gen1"), 15))), # post-hoc hypothesis
  H7 = matrix(factor(rep(c("gen1", "gen2", "gen3", "gen4", "gen1"), 15))), # post-hoc hypothesis
  H8 = matrix(factor(c(rep(c("amb_gen1", "amb_gen2", "amb_gen2", "amb_gen2", "amb_gen1"), 5),
                     rep(c("sev_gen1", "sev_gen2", "sev_gen2", "sev_gen2", "sev_gen1"),5),
                     rep(c("amb_gen1", "amb_gen2", "amb_gen2", "amb_gen2", "amb_gen1"), 5)))))

names(Z_models) <- c("all_same", "diff_treat", "diff_genet", "mhw_same", "amb_ext_same", "AE_BCD", "AE_B_C_D", "ambExt_AE_BCD") #, "heatwave_recovery") #, "all_diff")
```

#### -------------- Notes from WSN 2024:
- C matrix - all numeric matrix, 1 column -> This doesn't work though, and needs to be same dimensions as c matrix
- c matrix = temp covariate 
- B = Identity, random walk
- Z = CUSTOM matrix, the hypotheses
- A = A_matrix?
- R = diagonal and equal #identity?
- U = unequal?
- Q = diagonal and equal?

#### -------------- My old OG MARSS model with notes:
- B = "identity", #pretty sure this is right. Could also be diagonal and unequal or diagonal and equal though.
- U = "zero", #unequal has worse AIC/BIC -  [[ONLY WORKS IF Z SCORE DATA]]
- Q = "diagonal and equal", #All tanks have same mixing (but are not correlated, hence diagonal); random walk diagonal and unequal has worse AIC/BIC (and that would mean diff mixing between tanks - but check this with "new" model)
- Z = "placeholder",
- A = "zero", # SET TO ZERO! (according to Robert)
- R = "diagonal and equal", #observation errors are independent and equally distributed!

## Old model list
```{r, eval=FALSE}
mod_list <- list(
  B = "identity", #MAYBE diagonal and equal or diagonal and unequal because body growth over time is autocorrelated, however both these options don't converge
  U = "unequal",
  Q = "diagonal and equal", #MAYBE diagonal and unequal if water mixing in each tank is different (but unlikely)
  Z = "placeholder",
  A = "zero",
  R = "diagonal and equal",
  C = C_matrix,
  c = temp_covariate)
```
