---
title: "MARSS - body size"
author: "Amelia Ritger"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MARSS)
```

# Load data
```{r}
source("mergeAllData.R", local = knitr::knit_global())
#source("readRPiData.R", local = knitr::knit_global()) # Load temperature data AFTER you have uncommented and run this line
weekly_temp <- read_csv(here("experiment", "data", "weekly_temperature.csv"))
daily_temp <- read_csv(here("experiment", "data", "three_day_temperature.csv"))

end_mhw <- ymd("2023-12-12")
```

# Functions
```{r}
# For plotting parameters
plot.pars <- function(fit) {
    df <- broom::tidy(fit)
    df$parameter <- sapply(df$term, function(x) {
        stringr::str_split(x, "[.]")[[1]][1]
    })
    df$parameter[df$parameter %in% c("R", "Q")] <- "R and Q"
    ggplot2::ggplot(df, ggplot2::aes(x = term, y = estimate)) + 
        ggplot2::geom_point() + ggplot2::geom_errorbar(ggplot2::aes(ymin = conf.low, 
        ymax = conf.up), width = 0.2) + geom_hline(yintercept = 0, 
        col = "blue") + facet_wrap(~parameter, scales = "free")
}

# For generating H3/H5 Z matrices - staggered 1 values across specific rows and columns
fill_staggered_ones <- function(Z, rows, cols) { #Fill staggered 1s across specified rows/cols
  rows_to_apply <- intersect(rows, 1:nrow(Z))   # Ensure the row numbers are within the bounds of the matrix
  for (col in cols) { # Loop over the specified columns
    start_row <- col  # Stagger the starting row for each column
    for (i in seq(start_row, max(rows_to_apply), by = 5)) {
      if (i %in% rows_to_apply) {
        Z[i, col] <- 1
      }
    }
  }
  return(Z)
}
```

#### Hypotheses
- H1: All genets and all treatments are the same
- H2: All genets are the same but each treatment is different
- H3: All treatments are the same but each genet is different
- H4: All genets are the same and both MHW treatments are the same, but different from control treatment
- H5: All genets are the same and the extreme MHW is the same as the control treatment (post-hoc hypothesis, - based off figures)
- H6: All treatments are the same, and genets A/E are the same and genets B/C/D are the same (post-hoc - hypothesis, based off figures)
- H7: All treatments are the same, and genets A/E are the same (post-hoc hypothesis, alternative to H6)
- H8: The exterme MHW is the same as the control treatment, and genets A/E are the same and genets B/C/D are the same (post-hoc hypothesis, alternative to H5:H8)
- H9: All genets and all treatments are different

# Body size 

## Generate temp covariate matrix - ONLY for MHW times
```{r}
temp_covariate <- weekly_temp %>%
  filter(friday < ymd("2024-02-16"),
         friday <= end_mhw) %>%
  arrange(friday, tank) %>%
  #slice(rep(1:n(), each = 5)) %>%
  #mutate(tank_rep = paste0(tank, letters[1:5][rep(1:5, times = n()/5)])) %>%
  select(tank, friday, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(week = row_number()) %>%
  select(-friday) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()
```

## Prep  body size data
```{r}
all_size <- all %>%
  filter(!is.na(avg_size),
         date <= end_mhw) %>%
  mutate(week = as.numeric(difftime(date, min(weekly_temp$friday), units = "weeks")) + 1,
         avg_size_log = log(avg_size),
         sd_size_log = log(sd_size),
         tank = as.numeric(tank)) %>%
  select(date, week, tank, mhw, genet, avg_size_log) %>%
  left_join(weekly_temp, by = c("date" = "friday", "tank"))

all_size_marss <- all_size %>%
  mutate(genet_tank = paste(genet, tank, sep = "_")) %>%
  select(-date, -tank, -genet, -mhw, -treatment, -avg_temp, -min_temp, -max_temp) %>%
  pivot_wider(names_from = genet_tank, values_from = avg_size_log) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()

marss_data <- all_size_marss
```

## Hypothesis 2: All genets are the same but each treatment is different 
```{r, eval=FALSE}
Z <- matrix(0, nrow = 75, ncol = 3) # Fill in the matrix # First 25 rows get a 1 in column 1 (treatment 1) 
Z[1:25, 1] <- 1 # Next 25 rows (26-50) get a 1 in column 2 (treatment 2) 
Z[26:50, 2] <- 1 # Last 25 rows (51-75) get a 1 in column 3 (treatment 3) 
Z[51:75, 3] <- 1 # Optional: Name the columns for clarity 
colnames(Z) <- c("treatment 1", "treatment 2", "treatment 3")

C = matrix(list(0), 3, 15)
C[1, 1:5] <- "treatment1"
C[2, 6:10] <- "treatment2"
C[3, 11:15] <- "treatment3"

mod_list <- list(
  B = "identity",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z,
  A = "zero",
  R = "diagonal and equal",
  C = C,
  c = temp_covariate,
  tinitx=0)

fit_marss <- MARSS(y=marss_data, model = mod_list, silent=TRUE, control = list(maxit = 5000))

out_indv <- data.frame(logLik = fit_marss$logLik, 
                  AICc = fit_marss$AICc, num.param = fit_marss$num.params, 
                  num.iter = fit_marss$numIter, converged = !fit_marss$convergence)
out_indv
MARSSparamCIs(fit_marss)
plot.pars(fit_marss)

#resids <- MARSSresiduals(fit_indv_alt, type = "tt1")
#for (j in 1:75) {
#    plot(resids$model.residuals[j, ], ylab = "model residuals", 
#        xlab = "")
#    abline(h = 0)
#    title(paste(rownames(marss_data)[j]))
#    }
```

### Let's create a "quick" for loop running through options of Q and U with this:
```{r, eval=FALSE}
parameter_combinations <- expand.grid(
  #R = c("identity", "unequal", "zero", "diagonal and unequal", "diagonal and equal"),
  Q = c("diagonal and unequal", "diagonal and equal"),
  U = c("unequal", "equal", "zero"),
  stringsAsFactors = FALSE
)

# Create an empty data frame to store the results
results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  Q = character(), U = character(),  # Add model parameter labels
  stringsAsFactors = FALSE
)

# Loop over each combination of parameters
for (i in 1:nrow(parameter_combinations)) {
  # Extract the current combination of parameters
  current_params <- parameter_combinations[i, ]
  
  # Create the mod_list with current parameter values
  mod_list_loop <- list(
    B = "identity",
    U = current_params$U,
    Q = current_params$Q,
    Z = Z,
    A = "zero",  # DO NOT CHANGE!
    R = "diagonal and equal",  # DO NOT CHANGE!
    C = C,
    c = temp_covariate,
    tinitx = 0)

    # Fit the MARSS model
  fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
  
  # Extract relevant output
  out_indv <- data.frame(
    Q = current_params$Q, U = current_params$U,
    logLik = fit_marss$logLik, 
    AICc = fit_marss$AICc, 
    num.param = fit_marss$num.params, 
    num.iter = fit_marss$numIter, 
    converged = !fit_marss$convergence
  )
  
  # Append the results to the results data frame
  results <- rbind(results, out_indv)
}

results <- results %>%
  mutate(hypothesis = "H2")

# save results as .csv
#write_csv(results, here("experiment", "data", "marss_output", "H2.csv"))
```

## Hypothesis 1: All genets and all treatments are the same
```{r, eval=FALSE}
Z <- matrix(1, nrow = 75, ncol = 1)
colnames(Z) <- c("all same")

C = matrix(list(1), 1, 15)

mod_list <- list(
  B = "identity",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z,
  A = "zero",
  R = "diagonal and equal",
  C = C,
  c = temp_covariate,
  tinitx=0)

fit_marss <- MARSS(y=marss_data, model = mod_list, silent=TRUE, control = list(maxit = 5000))

out_indv <- data.frame(logLik = fit_indv_alt$logLik, 
                  AICc = fit_indv_alt$AICc, num.param = fit_indv_alt$num.params, 
                  num.iter = fit_indv_alt$numIter, converged = !fit_indv_alt$convergence)
out_indv
MARSSparamCIs(fit_indv_alt)
plot.pars(fit_indv_alt)
```

### For loop
```{r, eval=FALSE}
parameter_combinations <- expand.grid(
  Q = c("diagonal and unequal", "diagonal and equal"),
  U = c("unequal", "equal", "zero"),
  stringsAsFactors = FALSE
)

# Create an empty data frame to store the results
results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  Q = character(), U = character(),  # Add model parameter labels
  stringsAsFactors = FALSE
)

# Loop over each combination of parameters
for (i in 1:nrow(parameter_combinations)) {
  # Extract the current combination of parameters
  current_params <- parameter_combinations[i, ]
  
  # Create the mod_list with current parameter values
  mod_list_loop <- list(
    B = "identity",
    U = current_params$U,
    Q = current_params$Q,
    Z = Z,
    A = "zero",  # DO NOT CHANGE!
    R = "diagonal and equal",  # DO NOT CHANGE!
    C = C,
    c = temp_covariate,
    tinitx = 0)

    # Fit the MARSS model
  fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
  
  # Extract relevant output
  out_indv <- data.frame(
    Q = current_params$Q, U = current_params$U,
    logLik = fit_marss$logLik, 
    AICc = fit_marss$AICc, 
    num.param = fit_marss$num.params, 
    num.iter = fit_marss$numIter, 
    converged = !fit_marss$convergence
  )
  
  # Append the results to the results data frame
  results <- rbind(results, out_indv)
}

results <- results %>%
  mutate(hypothesis = "H1")

# save results as .csv
#write_csv(results, here("experiment", "data", "marss_output", "H1.csv"))
```


## Let's try one big for loop where we test different combos of Z/C, Q and U...

# Generate the Q and U comparisons and the list of hypotheses
```{r}
parameter_combinations <- expand.grid(
  Q = c("diagonal and unequal", "diagonal and equal"),
  U = c("unequal", "equal"), #U should not be zero because our data are not de-meaned
  stringsAsFactors = FALSE
)

zc_matrix_list <- list(
  H1 = list(Z = matrix(1, nrow = 75, ncol = 1), C = matrix(list(1), 1, 15)), #all genets and all treatments are the same
  H2 = list(Z = matrix(0, nrow = 75, ncol = 3), C = matrix(list(0), 3, 15)), #all genets are the same, all treatments are different
  H3 = list(Z = matrix(0, nrow = 75, ncol = 5), C = matrix(list(0), 5, 15)), #all genets are different, all treatments are different
  H4 = list(Z = matrix(0, nrow = 75, ncol = 2), C = matrix(list(0), 2, 15)), #all genets are the same, MHW treatments are different from control
  H5 = list(Z = matrix(0, nrow = 75, ncol = 15), C = matrix(list(0), 15, 15)) #all genets and all treatments are different 
)

# H2: All treatments are different -> Z (75x3), C (3x15)
zc_matrix_list$H2$Z[1:25, 1] <- 1  # Treatment 1
zc_matrix_list$H2$Z[26:50, 2] <- 1  # Treatment 2
zc_matrix_list$H2$Z[51:75, 3] <- 1  # Treatment 3
colnames(zc_matrix_list$H2$Z) <- c("treatment 1", "treatment 2", "treatment 3")
zc_matrix_list$H2$C[1, 1:5] <- "treatment1"
zc_matrix_list$H2$C[2, 6:10] <- "treatment2"
zc_matrix_list$H2$C[3, 11:15] <- "treatment3"

# H3: All genets are different -> Z (75x5), C (5x15)
# #fill_staggered_ones <- function(Z, cols) {
#   for (col in cols) {
#     start_row <- col  # Stagger the starting row for each column
#     for (i in seq(start_row, nrow(Z), by = 5)) {
#       Z[i, col] <- 1
#     }
#   }
#   return(Z)
# }
zc_matrix_list$H3$Z <- fill_staggered_ones(zc_matrix_list$H3$Z, 1:75, 1:5)
colnames(zc_matrix_list$H3$Z) <- c("genetA", "genetB", "genetC", "genetD", "genetE")
for (i in 1:nrow(zc_matrix_list$H3$C)) {
  zc_matrix_list$H3$C[i, ] <- rep(letters[i], ncol(zc_matrix_list$H3$C))
}

# H4: MHW treatments are different from control -> Z (75x2), C (2x15)
zc_matrix_list$H4$Z[1:25, 1] <- 1  # Treatment 1
zc_matrix_list$H4$Z[26:75, 2] <- 1  # Treatment 2
colnames(zc_matrix_list$H4$Z) <- c("ambient", "mhw")
zc_matrix_list$H4$C[1, 1:5] <- "ambient"
zc_matrix_list$H4$C[2, 6:15] <- "mhw"

# H5: All genets and all treatments are different -> Z (75x15), C (15x15)
zc_matrix_list$H5$Z <- fill_staggered_ones(zc_matrix_list$H5$Z,1:25, 1:5)
zc_matrix_list$H5$Z <- fill_staggered_ones(zc_matrix_list$H5$Z,26:50, 6:10)
zc_matrix_list$H5$Z <- fill_staggered_ones(zc_matrix_list$H5$Z,51:75, 11:15)
colnames(zc_matrix_list$H5$Z) <- c("amb_A", "amb_B", "amb_C", "amb_D", "amb_E",
                                   "sev_A", "sev_B", "sev_C", "sev_D", "sev_E",
                                   "ext_A", "ext_B", "ext_C", "ext_D", "ext_E")
for (i in 1:nrow(zc_matrix_list$H5$C)) {
  letter_index <- (i - 1) %% 5 + 1  # This ensures we loop through 'a' to 'e'
  letter <- letters[letter_index]  # Get the current genet
    if (i >= 1 & i <= 5) {
    suffix <- "ambient_"
  } else if (i >= 6 & i <= 10) {
    suffix <- "severe_"
  } else if (i >= 11 & i <= 15) {
    suffix <- "extreme_"
  } else {
    suffix <- ""
  }
  zc_matrix_list$H5$C[i, ] <- rep(paste(suffix, letter, sep = ""), 15)
}
```

### Run the for loop to test each Q and U with each hypothesis (Z/C matrix)
```{r, eval=FALSE}
marss_results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  Q = character(), U = character(),  # Add model parameter labels
  Z_group = character(), C_group = character(),  # To track Z and C combinations
  stringsAsFactors = FALSE
)

# Loop over each combination of Z and C
for (hypothesis_number in names(zc_matrix_list)) {
  # Extract the current Z and C combination
  Z <- zc_matrix_list[[hypothesis_number]]$Z
  C <- zc_matrix_list[[hypothesis_number]]$C
  
  # Loop over each combination of parameters (Q and U)
  for (i in 1:nrow(parameter_combinations)) {
    # Extract the current combination of parameters for Q and U
    current_params <- parameter_combinations[i, ]
    
    # Create the mod_list with current parameter values
    mod_list_loop <- list(
      B = "identity",
      U = current_params$U,
      Q = current_params$Q,
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0
    )
    
    # Fit the MARSS model
    fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
    
    # Extract relevant output
    out_indv <- data.frame(
      Hypothesis = hypothesis_number, Q = current_params$Q, U = current_params$U,
      logLik = fit_marss$logLik, 
      AICc = fit_marss$AICc, 
      num.param = fit_marss$num.params, 
      num.iter = fit_marss$numIter, 
      converged = !fit_marss$convergence
    )
    
    # Append the results to the results data frame
    marss_results <- rbind(marss_results, out_indv)
  }
}

marss_results %>%
  arrange(AICc)
```

### Check out the best performing model for H5 (Q=diagonal and equal, U=equal)
```{r, eval=FALSE}
Z <- matrix(0, nrow = 75, ncol = 15)
C <- matrix(list(0), 15, 15)
  
Z <- fill_staggered_ones(Z,1:25, 1:5)
Z <- fill_staggered_ones(Z,26:50, 6:10)
Z <- fill_staggered_ones(Z,51:75, 11:15)
colnames(zc_matrix_list$H5$Z) <- c("amb_A", "amb_B", "amb_C", "amb_D", "amb_E",
                                   "sev_A", "sev_B", "sev_C", "sev_D", "sev_E",
                                   "ext_A", "ext_B", "ext_C", "ext_D", "ext_E")

for (i in 1:nrow(C)) {
  letter_index <- (i - 1) %% 5 + 1  # This ensures we loop through 'a' to 'e'
  letter <- letters[letter_index]  # Get the current genet
    if (i >= 1 & i <= 5) {
    suffix <- "ambient_"
  } else if (i >= 6 & i <= 10) {
    suffix <- "severe_"
  } else if (i >= 11 & i <= 15) {
    suffix <- "extreme_"
  } else {
    suffix <- ""
  }
  C[i, ] <- rep(paste(suffix, letter, sep = ""), 15)
}

model_list <- list(
      B = "identity",
      U = "equal",
      Q = "diagonal and equal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0)

fit_marss_indv <- MARSS(y = marss_data, model = model_list, silent = TRUE, control = list(maxit = 5000))
fit_marss_indv
MARSSparamCIs(fit_marss_indv)
plot.pars(fit_marss_indv)

Q.est <- coef(fit_marss_indv, type = "matrix")$Q
corrplot::corrplot(cov2cor(Q.est))

resids <- MARSSresiduals(fit_marss_indv, type = "tt1")
for (j in 1:75) {
    plot(resids$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(rownames(marss_data)[j]))
    }
```

Red flags with the correlation matrix, so let's try hypothesis 3

### Check out the best performing model for H3 (Q=diagonal and equal, U=equal)
```{r, eval=FALSE}
Z = matrix(0, nrow = 75, ncol = 5)
C = matrix(list(0), 5, 15)

Z <- fill_staggered_ones(Z, 1:75, 1:5)
colnames(Z) <- c("genetA", "genetB", "genetC", "genetD", "genetE")
for (i in 1:nrow(C)) {
  C[i, ] <- rep(letters[i], ncol(C))
}

model_list <- list(
      B = "identity",
      U = "equal",
      Q = "diagonal and equal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0)

fit_marss_indv <- MARSS(y = marss_data, model = model_list, silent = TRUE, control = list(maxit = 5000))
fit_marss_indv

MARSSparamCIs(fit_marss_indv)
plot.pars(fit_marss_indv)
resids <- MARSSresiduals(fit_marss_indv, type = "tt1")
for (j in 1:75) {
    plot(resids$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(rownames(marss_data)[j]))
    }
```

# Population growth

## Generate temp covariate matrix - ONLY for MHW times
```{r}
temp_covariate <- daily_temp %>%
  filter(date >= ymd("2023-10-06"),
         date < ymd("2024-02-16"),
         date <= end_mhw) %>%
  arrange(date, tank) %>%
  mutate(avg_temp = avg_temp_last_3_days) %>%
  select(tank, date, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(day = row_number()) %>%
  select(-date) %>%
  column_to_rownames(var = "day") %>%
  as.matrix() %>%
  t()
```

## Prep reproduction data
```{r}
all_repro <- all %>%
  filter(date >= ymd("2023-10-06"),
                    date <= end_mhw) %>%
  mutate(#week = as.numeric(difftime(date, min(weekly_temp$friday), units = "weeks")) + 1,
         #avg_size_log = log(avg_size),
         #sd_size_log = log(sd_size),
         tank = as.numeric(tank)) %>%
  select(date, tank, mhw, genet, n_true) %>%
  left_join(daily_temp, by = c("date", "tank"))

all_repro_marss <- all_repro %>%
  mutate(genet_tank = paste(genet, tank, sep = "_")) %>%
  select(date, genet_tank, n_true) %>%
  #select(-date, -tank, -genet, -mhw, -treatment, -avg_temp, -min_temp, -max_temp) %>%
  pivot_wider(names_from = genet_tank, values_from = n_true) %>%
  column_to_rownames(var = "date") %>%
  as.matrix() %>%
  t()

marss_data <- all_repro_marss
```

## Run the for loop to test each Q and U with each hypothesis (Z/C matrix)
```{r, eval=FALSE}
marss_results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  Q = character(), U = character(),  # Add model parameter labels
  Z_group = character(), C_group = character(),  # To track Z and C combinations
  stringsAsFactors = FALSE
)

# Loop over each combination of Z and C
for (hypothesis_number in names(zc_matrix_list)) {
  # Extract the current Z and C combination
  Z <- zc_matrix_list[[hypothesis_number]]$Z
  C <- zc_matrix_list[[hypothesis_number]]$C
  
  # Loop over each combination of parameters (Q and U)
  for (i in 1:nrow(parameter_combinations)) {
    # Extract the current combination of parameters for Q and U
    current_params <- parameter_combinations[i, ]
    
    # Create the mod_list with current parameter values
    mod_list_loop <- list(
      B = "identity",
      U = current_params$U,
      Q = current_params$Q,
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0
    )
    
    # Fit the MARSS model
    fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
    
    # Extract relevant output
    out_indv <- data.frame(
      Hypothesis = hypothesis_number, Q = current_params$Q, U = current_params$U,
      logLik = fit_marss$logLik, 
      AICc = fit_marss$AICc, 
      num.param = fit_marss$num.params, 
      num.iter = fit_marss$numIter, 
      converged = !fit_marss$convergence
    )
    
    # Append the results to the results data frame
    marss_results <- rbind(marss_results, out_indv)
  }
}

marss_results %>%
  arrange(AICc)
```

Did not converge: Check why?
- H2, Q = diagonal and unequal
- H3, Q = diagonal and unequal
- H4, Q = diagonal and unequal, unequal (but equal converged in 38 iterations!)
- H5, Q = diagonal and unequal

### Check Q = diagonal and unequal convergence issue using H3 as an example
```{r}
Z = matrix(0, nrow = 75, ncol = 5)
C = matrix(list(0), 5, 15)

Z <- fill_staggered_ones(Z, 1:75, 1:5)
colnames(Z) <- c("genetA", "genetB", "genetC", "genetD", "genetE")
for (i in 1:nrow(C)) {
  C[i, ] <- rep(letters[i], ncol(C))
}

model_list <- list(
      B = "identity",
      U = "equal",
      Q = "diagonal and unequal",
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0)

fit_marss_indv <- MARSS(y = marss_data, model = model_list, silent = TRUE, control = list(maxit = 10000))
fit_marss_indv

MARSSparamCIs(fit_marss_indv)
plot.pars(fit_marss_indv)
resids <- MARSSresiduals(fit_marss_indv, type = "tt1")
for (j in 1:75) {
    plot(resids$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(rownames(marss_data)[j]))
    }
```

Model still doesn't converge even after 10k iterations...

# Behavior 

## Generate temp covariate matrix - ONLY for MHW times
```{r}
temp_covariate <- daily_temp %>%
  filter(date >= ymd("2023-10-06"),
         date < ymd("2024-02-16"),
         date <= end_mhw) %>%
  arrange(date, tank) %>%
  mutate(avg_temp = avg_temp_last_3_days) %>%
  select(tank, date, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(day = row_number()) %>%
  select(-date) %>%
  column_to_rownames(var = "day") %>%
  as.matrix() %>%
  t()
```

## Prep behavior data
```{r}
all_behavior <- all %>%
  filter(date >= ymd("2023-10-06"),
                    date <= end_mhw) %>%
  mutate(tank = as.numeric(tank)) %>%
  select(date, tank, mhw, genet, percent_open) %>%
  left_join(daily_temp, by = c("date", "tank"))

all_behavior_marss <- all_behavior %>%
  mutate(genet_tank = paste(genet, tank, sep = "_"),
         number_open = percent_open*100) %>% #only need this if we aren't z-scoring I think
  select(date, genet_tank, number_open) %>%
  pivot_wider(names_from = genet_tank, values_from = number_open) %>%
  column_to_rownames(var = "date") %>%
  as.matrix() %>%
  t()

marss_data <- all_behavior_marss
```

### Z-score the data (BECAUSE)
```{r, eval=FALSE}
the.mean <- apply(marss_data, 1, mean)
the.sigma <- sqrt(apply(marss_data, 1, var))
marss_data_z <- (marss_data - the.mean) * (1/the.sigma)

the.mean <- apply(temp_covariate, 1, mean)
the.sigma <- sqrt(apply(temp_covariate, 1, var))
temp_covariate_z <- (temp_covariate - the.mean) * (1/the.sigma)
```

## Run the for loop to test each Q with each hypothesis (Z/C matrix)
```{r, eval=FALSE}
marss_results <- data.frame(
  logLik = numeric(), 
  AICc = numeric(), 
  num.param = numeric(),
  num.iter = numeric(), 
  converged = logical(),
  Q = character(), U = character(),  # Add model parameter labels
  Z_group = character(), C_group = character(),  # To track Z and C combinations
  stringsAsFactors = FALSE
)

# Loop over each combination of Z and C
for (hypothesis_number in names(zc_matrix_list)) {
  # Extract the current Z and C combination
  Z <- zc_matrix_list[[hypothesis_number]]$Z
  C <- zc_matrix_list[[hypothesis_number]]$C
  
  # Loop over each combination of parameters (Q and U)
  for (i in 1:nrow(parameter_combinations)) {
    # Extract the current combination of parameters for Q and U
    current_params <- parameter_combinations[i, ]
    
    # Create the mod_list with current parameter values
    mod_list_loop <- list(
      B = "identity",
      U = current_params$U,
      Q = current_params$Q,
      Z = Z,
      A = "zero",  # DO NOT CHANGE!
      R = "diagonal and equal",  # DO NOT CHANGE!
      C = C,
      c = temp_covariate,
      tinitx = 0
    )
    
    # Fit the MARSS model
    fit_marss <- MARSS(y = marss_data, model = mod_list_loop, silent = TRUE, control = list(maxit = 5000))
    
    # Extract relevant output
    out_indv <- data.frame(
      Hypothesis = hypothesis_number, Q = current_params$Q, U = current_params$U,
      logLik = fit_marss$logLik, 
      AICc = fit_marss$AICc, 
      num.param = fit_marss$num.params, 
      num.iter = fit_marss$numIter, 
      converged = !fit_marss$convergence
    )
    
    # Append the results to the results data frame
    marss_results <- rbind(marss_results, out_indv)
  }
}

marss_results %>%
  arrange(AICc)
```

H5 diagonal and unequal did not converge for either U - even with 10k iterations. Why? (TBD, let's roll with it though)

                  
--------------------------------------------------------------------------------

# Everything below this is "old" (but reference it if needed)

# Make matrices

## Make C, c matrices - temperature covariate (15 tanks, 5 genets in each tank)
```{r}
temp_covariate <- weekly_temp %>%
  filter(friday < ymd("2024-02-16")) %>%
  arrange(friday, tank) %>%
  #slice(rep(1:n(), each = 5)) %>%
  #mutate(tank_rep = paste0(tank, letters[1:5][rep(1:5, times = n()/5)])) %>%
  select(tank, friday, avg_temp) %>%
  pivot_wider(names_from = tank, values_from = avg_temp) %>%
  mutate(week = row_number()) %>%
  select(-friday) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()

C_matrix <- matrix(factor(paste("tank", rep(1:15, each = 1), sep = "")), ncol = 1)
#3x15 matrix
```

## Make Z matrix list. Might need to incorporate that # cols = # treatments, because then I need to change C matrix dimensions and resolve underconstrained issues with Z, B matrices
```{r}
Z_models <- list(
  H1 = matrix(factor(rep(1, 75))),
  H2 = matrix(factor(c(rep("amb", 25), rep("sev", 25), rep("ext", 25)))),
  H3 = matrix(factor(rep(c("A", "B", "C", "D", "E"), 15))),
  H4 = matrix(factor(c(rep("amb", 25), rep("mhw", 50)))),
  H5 = matrix(factor(c(rep("amb", 25), rep("sev", 25), rep("amb", 25)))), # post-hoc hypothesis
  H6 = matrix(factor(rep(c("gen1", "gen2", "gen2", "gen2", "gen1"), 15))), # post-hoc hypothesis
  H7 = matrix(factor(rep(c("gen1", "gen2", "gen3", "gen4", "gen1"), 15))), # post-hoc hypothesis
  H8 = matrix(factor(c(rep(c("amb_gen1", "amb_gen2", "amb_gen2", "amb_gen2", "amb_gen1"), 5),
                     rep(c("sev_gen1", "sev_gen2", "sev_gen2", "sev_gen2", "sev_gen1"),5),
                     rep(c("amb_gen1", "amb_gen2", "amb_gen2", "amb_gen2", "amb_gen1"), 5)))))

names(Z_models) <- c("all_same", "diff_treat", "diff_genet", "mhw_same", "amb_ext_same", "AE_BCD", "AE_B_C_D", "ambExt_AE_BCD") #, "heatwave_recovery") #, "all_diff")
```

#### Hypotheses
- H1: All genets and all treatments are the same
- H2: All genets are the same but each treatment is different
- H3: All treatments are the same but each genet is different
- H4: All genets are the same and both MHW treatments are the same, but different from control treatment
- H5: All genets are the same and the extreme MHW is the same as the control treatment (post-hoc hypothesis, - based off figures)
- H6: All treatments are the same, and genets A/E are the same and genets B/C/D are the same (post-hoc - hypothesis, based off figures)
- H7: All treatments are the same, and genets A/E are the same (post-hoc hypothesis, alternative to H6)
- H8: The exterme MHW is the same as the control treatment, and genets A/E are the same and genets B/C/D are the same (post-hoc hypothesis, alternative to H5:H8)
- H9: (Do not use!) All genets and all treatments are different
- H10: The heatwave period was different from the recovery period 

#### -------------- Notes from WSN 2024:
- C matrix - all numeric matrix, 1 column -> This doesn't work though, and needs to be same dimensions as c matrix
- c matrix = temp covariate 
- B = Identity, random walk
- Z = CUSTOM matrix, the hypotheses
- A = A_matrix?
- R = diagonal and equal #identity?
- U = unequal?
- Q = diagonal and equal?

#### -------------- My old OG MARSS model with notes:
- B = "identity", #pretty sure this is right. Could also be diagonal and unequal or diagonal and equal though.
- U = "zero", #unequal has worse AIC/BIC -  [[ONLY WORKS IF Z SCORE DATA]]
- Q = "diagonal and equal", #All tanks have same mixing (but are not correlated, hence diagonal); random walk diagonal and unequal has worse AIC/BIC (and that would mean diff mixing between tanks - but check this with "new" model)
- Z = "placeholder",
- A = "zero", # SET TO ZERO! (according to Robert)
- R = "diagonal and equal", #observation errors are independent and equally distributed!

## Set up model list
```{r}
mod_list <- list(
  B = "identity", #MAYBE diagonal and equal or diagonal and unequal because body growth over time is autocorrelated, however both these options don't converge
  U = "unequal",
  Q = "diagonal and equal", #MAYBE diagonal and unequal if water mixing in each tank is different (but unlikely)
  Z = "placeholder",
  A = "zero",
  R = "diagonal and equal",
  C = C_matrix,
  c = temp_covariate)
```

# Body size 

## Prep  data
```{r}
all_size <- all %>%
  filter(!is.na(avg_size)) %>%
  mutate(week = as.numeric(difftime(date, min(weekly_temp$friday), units = "weeks")) + 1,
         avg_size_log = log(avg_size),
         sd_size_log = log(sd_size),
         tank = as.numeric(tank)) %>%
  select(date, week, tank, mhw, genet, avg_size_log) %>%
  left_join(weekly_temp, by = c("date" = "friday", "tank"))

all_size_marss <- all_size %>%
  mutate(genet_tank = paste(genet, tank, sep = "_")) %>%
  select(-date, -tank, -genet, -mhw, -treatment, -avg_temp, -min_temp, -max_temp) %>%
  pivot_wider(names_from = genet_tank, values_from = avg_size_log) %>%
  column_to_rownames(var = "week") %>%
  as.matrix() %>%
  t()

marss_data <- all_size_marss
```

#Set up function for plotting parameters
```{r}
plot.pars <- function(fit) {
    df <- broom::tidy(fit)
    df$parameter <- sapply(df$term, function(x) {
        stringr::str_split(x, "[.]")[[1]][1]
    })
    df$parameter[df$parameter %in% c("R", "Q")] <- "R and Q"
    ggplot2::ggplot(df, ggplot2::aes(x = term, y = estimate)) + 
        ggplot2::geom_point() + ggplot2::geom_errorbar(ggplot2::aes(ymin = conf.low, 
        ymax = conf.up), width = 0.2) + geom_hline(yintercept = 0, 
        col = "blue") + facet_wrap(~parameter, scales = "free")
}
```

# Loop through Z matrix models 
```{r, eval=FALSE}
out.tab <- NULL
fits <- list()

for (i in 1:length(Z_models)) {
    mod_list$Z <- Z_models[[i]]
    fit <- MARSS(y=marss_data, model = mod_list, silent = TRUE, 
        control = list(maxit = 5000))
    out <- data.frame(H = names(Z_models)[i], logLik = fit$logLik, 
        AICc = fit$AICc, num.param = fit$num.params, m = length(unique(Z_models[[i]])), 
        num.iter = fit$numIter, converged = !fit$convergence)
    out.tab <- rbind(out.tab, out)
    fits <- c(fits, list(fit))
}

print(out.tab) #looking for: higher log likelihood, lower AIC

for (i in 1:length(fits)) {
  par(mfrow = c(2, 3))
  resids <- MARSSresiduals(fits[[i]], type = "tt1")
  for (j in 1:15) {
    plot(resids$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(names(Z_models[i]), ":", rownames(marss_data)[j]))
    }
}

# Plots 1 (xtT) & 2 (fitted.ytT): Do fitted values seem reasonable?
# Plot 3 (model.resids.ytt1): Do resids have temporal patterns? Do 95% of resids fall withing the CIs?
# Plot 4 (std.model.resids.ytT): Do resids have temporal patterns? Do 95% of resids fall withing the CIs?
# Plot 5 (std.state.resids.xtT): Any outliers?
# Plots 6 & 7 (qqplot.std.model.resids.ytt1: Are resids normal (straight line)?
# Plot 8 (acf.std.model.resids.ytt1): Do resids have temporal autocorrelation?
```

#### The residual plots are not looking great... and they're essentially the same across all models(??) 

## Alternative Z/C matrices
```{r}
Z <- matrix(0, nrow = 75, ncol = 3) # Fill in the matrix # First 25 rows get a 1 in column 1 (treatment 1) 
Z[1:25, 1] <- 1 # Next 25 rows (26-50) get a 1 in column 2 (treatment 2) 
Z[26:50, 2] <- 1 # Last 25 rows (51-75) get a 1 in column 3 (treatment 3) 
Z[51:75, 3] <- 1 # Optional: Name the columns for clarity 
colnames(Z) <- c("treatment 1", "treatment 2", "treatment 3")

C = matrix(list(0), 3, 15)
C[1, 1:5] <- "treatment1" # Next 25 rows (26-50) get a 1 in column 2 (treatment 2) 
C[2, 5:10] <- "treatment2" # Last 25 rows (51-75) get a 1 in column 3 (treatment 3) 
C[3, 11:15] <- "treatment3" # Optional: Name the columns for clarity 

mod_list_alt <- list(
  B = "identity",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z,
  A = "zero",
  R = "diagonal and equal",
  C = C,
  c = temp_covariate,
  tinitx=0)

fit_indv_alt <- MARSS(y=marss_data, model = mod_list_alt, silent=TRUE, control = list(maxit = 5000))
fit_indv_alt

out_indv <- data.frame(logLik = fit_indv_alt$logLik, 
                  AICc = fit_indv_alt$AICc, num.param = fit_indv_alt$num.params, 
                  num.iter = fit_indv_alt$numIter, converged = !fit_indv_alt$convergence)

#broom::tidy(fit_indv_alt)
MARSSparamCIs(fit_indv_alt)
plot.pars(fit_indv_alt)

resids <- MARSSresiduals(fit_indv_alt, type = "tt1")
for (j in 1:75) {
    plot(resids$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(rownames(marss_data)[j]))
    }
```

# DON'T RUN BELOW, WE DON'T WANT TO ZSCORE THE DATA
# Generate the z-scored data
```{r}
the.mean <- apply(marss_data, 1, mean)
the.sigma <- sqrt(apply(marss_data, 1, var))
marss_data_z <- (marss_data - the.mean) * (1/the.sigma)

the.mean <- apply(temp_covariate, 1, mean)
the.sigma <- sqrt(apply(temp_covariate, 1, var))
temp_covariate_z <- (temp_covariate - the.mean) * (1/the.sigma)
```

## Model the z-scored data
```{r}
mod_list_z <- list(
  B = "diagonal and equal", #identity is another option, but models have lower logLik and higher AICc; random walk diagonal and unequal gives same model output
  U = "zero", #note: unequal doesn't converge even after 5k iterations
  Q = "diagonal and equal", #All tanks have same mixing (also, random walk diagonal and unequal is the same AICc and very close Q estimates)
  Z = "placeholder",
  A = "zero", #because data have been z-scored
  R = "diagonal and equal", #observation errors are independent and have equal variances, because there was only ONE sampler (me)
  C = C_matrix,
  c = temp_covariate_z)

out_tab_z <- NULL
fits_z <- list()
for (i in 1:length(Z_models)) {
    mod_list_z$Z <- Z_models[[i]]
    fit <- MARSS(y=marss_data_z, model = mod_list_z, silent = TRUE, 
        control = list(maxit = 5000))
    #print(fit)
    out <- data.frame(H = names(Z_models)[i], logLik = fit$logLik, 
        AICc = fit$AICc, num.param = fit$num.params, m = length(unique(Z_models[[i]])), 
        num.iter = fit$numIter, converged = !fit$convergence)
    out_tab_z <- rbind(out_tab_z, out)
    fits_z <- c(fits_z, list(fit))
}

print(out_tab_z) #looking for: higher log likelihood, lower AIC

for (i in 1:length(fits_z)) {
  par(mfrow = c(2, 3))
  resids_z <- MARSSresiduals(fits_z[[i]], type = "tt1")
  for (j in 1:15) {
    plot(resids_z$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(names(Z_models[i]), ":", rownames(marss_data_z)[j]))
    }
}
```

#### The residual plots are looking better... but they're still essentially the same across all models(??) 

#### According to AICc (and logLik), definitely don't use "all_diff" and can probably also rule out "diff_genet"

### Try z-scored data with different matrices?
```{r}
mod_list_z <- list(
  B = "diagonal and unequal", #identity is another option, but models have lower logLik and higher AICc; random walk diagonal and unequal gives same model output
  U = "zero", #note: unequal doesn't converge even after 5k iterations
  Q = "diagonal and equal", #All tanks have same mixing (also, random walk diagonal and unequal is the same AICc and very close Q estimates)
  Z = "placeholder",
  A = "zero", #because data have been z-scored
  R = "diagonal and equal", #observation errors are independent and have equal variances, because there was only ONE sampler (me)
  C = C_matrix,
  c = temp_covariate_z)

out_tab_z <- NULL
fits_z <- list()
for (i in 1:length(Z_models)) {
    mod_list_z$Z <- Z_models[[i]]
    fit <- MARSS(y=marss_data_z, model = mod_list_z, silent = TRUE, 
        control = list(maxit = 5000))
    #print(fit)
    out <- data.frame(H = names(Z_models)[i], logLik = fit$logLik, 
        AICc = fit$AICc, num.param = fit$num.params, m = length(unique(Z_models[[i]])), 
        num.iter = fit$numIter, converged = !fit$convergence)
    out_tab_z <- rbind(out_tab_z, out)
    fits_z <- c(fits_z, list(fit))
}

print(out_tab_z) #looking for: higher log likelihood, lower AIC

for (i in 1:length(fits_z)) {
  par(mfrow = c(2, 3))
  resids_z <- MARSSresiduals(fits_z[[i]], type = "tt1")
  for (j in 1:15) {
    plot(resids_z$model.residuals[j, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(paste(names(Z_models[i]), ":", rownames(marss_data_z)[j]))
    }
}
```

## Visualize z-scored temperatures
```{r}
temp_z <- weekly_temp %>%
  filter(friday < ymd("2024-02-16")) %>%
  arrange(friday, tank) %>%
  select(tank, friday, avg_temp) %>%
  group_by(tank) %>%
  mutate(avg_temp_z = (avg_temp - mean(avg_temp, na.rm = TRUE)) /sd(avg_temp, na.rm = TRUE),
         treatment = ifelse(tank < 6, "ambient",
                            ifelse(tank > 10, "extreme",
                                   "severe")),
         tank = as.factor(tank)) %>%
  ungroup()

ggplot(temp_z, aes(x = friday, y = avg_temp_z, color = tank)) +
  geom_line() +
  labs(title = "Z-scored average temperature by tank",
       x = "Date",
       y = "Z-scored average temperature") +
  theme_minimal() +
  facet_wrap(~treatment)
```

# Do not run.... 

## Experiment with non 75x1 Z matrix dimensions
```{r, eval=FALSE}
#Each treatment different
Z_matrix2 <- matrix(factor(c(rep(c(1, 0, 0), 25),
                            rep(c(0, 1, 0), 25),
                            rep(c(0, 0, 1), 25))),
                   nrow = 75, byrow = TRUE)

# Combine the blocks into a single 75x3 matrix
matrix_data <- rbind(matrix(c(rep(1:5, each = 5), rep(0, 25), rep(0, 25)), nrow = 25, ncol = 3, byrow = FALSE),
                     matrix(c(rep(0, 25), rep(6:10, each = 5), rep(0, 25)), nrow = 25, ncol = 3, byrow = FALSE),
                     matrix(c(rep(0, 25), rep(0, 25), rep(10:14, each = 5)), nrow = 25, ncol = 3, byrow = FALSE))

C_matrix2 <- matrix(factor(rep(1:15, each = 5)), ncol = 3)

C_matrix2 <- matrix(0, nrow = 75, ncol = 3)
for (i in 1:3) {
  C_matrix2[((i-1)*5 + 1):(i*5), i] <- 1
}
C_matrix2 <- matrix(factor(C_matrix2), nrow = 75, ncol = 3)

model_list <- list(
  B = "identity",
  U = "zero",
  Q = "diagonal and equal",
  Z = Z_matrix2,
  A = A_matrix,
  R = "diagonal and equal",
  C = matrix_data,
  c = temp_covariate)

# non 75x1 Z matrix Does not run 
#fit1 <- MARSS(y=marss_data, model = model_list,
#              control = list(maxit = 1000))
```
