---
title: "RPi ALL data visualization - For METHODS paper"
author: "Amelia Ritger"
date: "2023-08-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(janitor)
```

# Load data and merge all temp files
```{r, message=FALSE, warning=FALSE}
rm(list = ls())
#temp_files <- list.files(here("experiment", "data", "rpi"), pattern=".*csv")
temp_files <- list.files(here("experiment", "code", "python", "methodsPub", "external"), pattern=".*csv")
temp_all <- data.frame()

for (i in 1:length(temp_files)){
  print(temp_files[i])
  temp_data <- read_csv(here("experiment", "code", "python", "methodsPub", "external", temp_files[i])) #each file will be read in, specify which columns you need read in to avoid any error
  temp_all <- rbind(temp_all, temp_data) #for each iteration, bind the new data to the building dataset
}

temp_all <- temp_all %>%
  clean_names() %>%
  select(1:8)
```

# 2015 Blob data simulation
- FOR 2015 Blob data, ONLY KEEP DATA SINCE 03/13 11:00:00 (that's when I fixed the PID loop)
- I did a small PID loop adjustment at 13:00:00 but it shouldn't significantly affect the plot

## Plot tank temps
```{r}
temp_clean <- temp_all %>%
  mutate(timestamp = ymd_hms(timestamp)) %>%
  select(!last_col(offset=1):last_col(), #%>% #remove hot/cold tank temps
         -heater_status) %>%
  pivot_longer(cols=3:5, names_to = "sensor", values_to = "temperature") %>%
  filter(timestamp > ymd_hms("2024-03-13 11:00:00"),
         timestamp < ymd_hms("2024-03-16 18:01:00"))

ggplot(temp_clean, aes(x=timestamp)) +
  geom_line(aes(y=temperature, color=sensor), size=1) +
  geom_line(aes(y=temp_set), color="black")
```

## Plot average tank temp for every 15 sec reading and for every 1 minute 

```{r}
temp_avg <- temp_all %>%
  mutate(timestamp = ymd_hms(timestamp)) %>%
  select(!last_col(offset=1):last_col()) %>%
  pivot_longer(cols=4:6, names_to = "sensor", values_to = "temperature") %>%
  group_by(timestamp, temp_set, heater_status) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ungroup() %>%
  filter(timestamp > ymd_hms("2024-03-13 11:00:00"),
         timestamp < ymd_hms("2024-03-16 18:01:00"))

temp_five <- temp_all %>%
  mutate(timestamp = ymd_hms(timestamp),
         timestamp = floor_date(timestamp, "1 minute")) %>%
  select(!last_col(offset=1):last_col()) %>%
  pivot_longer(cols=4:6, names_to = "sensor", values_to = "temperature") %>%
  group_by(timestamp, temp_set) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ungroup() %>%
  filter(timestamp > ymd_hms("2024-03-13 11:00:00"),
         timestamp < ymd_hms("2024-03-16 18:01:00"))
  
image <- ggplot(temp_five, aes(x=timestamp)) +
  geom_point(aes(y=avg_temp), color="#D55E00", size=1) +
  geom_line(aes(y=temp_set), color="#000000", size=1.5, alpha=0.7) +
  scale_x_datetime(date_labels = "%b %d - %H:%M",
                   breaks = scales::date_breaks("12 hours"),
                   expand=c(0,0),
                   limits = c(min(temp_five$timestamp),
                              max(temp_five$timestamp))) +
  scale_y_continuous(limits = c(14.8, 15.8), breaks = seq(14.9, 15.8, by = 0.3)) +
  labs(y = expression(paste("Temperature (", degree, "C)")),
       x = "Date and time") +
       #title = "B") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=18),
        axis.text.y=element_text(size=18),
        axis.title.x=element_text(size=20),
        axis.title.y=element_text(size=20),
        legend.position="top")
        #plot.title = element_text(size=40))

#ggsave(here("experiment", "figures", "methodsPub", "2015blob.png"), width=12, height=6)
ggsave(file = here("experiment", "figures", "methodsPub","Fig3B.svg"), plot = image, width=12, height=6)
```

## Plot difference between "expected" and "observed"
```{r}
temp_diff <- temp_five %>%
  mutate(diff = round(temp_set - avg_temp, 3))

temp_diff_avg <- mean(temp_diff$diff)
temp_diff_min <- min(temp_diff$diff)
temp_diff_max <- max(temp_diff$diff)
temp_diff_sd <- sd(temp_diff$diff)
temp_diff_range <- temp_diff_max-temp_diff_min


ggplot(temp_diff, aes(x=timestamp)) +
  geom_point(aes(y=diff), color="#000000", size=1) +
  scale_x_datetime(date_labels = "%b %d - %H:%M",
                   breaks = scales::date_breaks("12 hours"),
                   expand=c(0,0),
                   limits = c(min(temp_five$timestamp),
                              max(temp_five$timestamp))) +
  scale_y_continuous(limits = c(-0.15, 0.07), breaks = seq(-0.15, 0.07, by = 0.1)) +
  labs(y = expression(paste("Temperature difference (", degree, "C)")),
       x = "Date and time") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=14),
        axis.text.y=element_text(size=14),
        axis.title.x=element_text(size=18),
        axis.title.y=element_text(size=18),
        legend.position="top")

#ggsave(here("experiment", "figures", "methodsPub", "blobDiff.png"), width=12, height=6)
```

# Conventional MHW data

## Plot data
```{r}
temp_simple <- temp_all %>%
  mutate(timestamp = ymd_hms(timestamp),
         timestamp = floor_date(timestamp, "1 minute")) %>%
  select(!last_col(offset=1):last_col()) %>%
  pivot_longer(cols=4:6, names_to = "sensor", values_to = "temperature") %>%
  group_by(timestamp, temp_set) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ungroup() %>%
  #remove datapoints between 2024-03-17 16:32:00 and 2024-03-17 16:57:00
  filter(timestamp != ymd_hms("2024-03-17 16:32:00"),
         timestamp > ymd_hms("2024-03-16 23:00:00"), #remove Blob data
         timestamp > ymd_hms("2024-03-17 08:00:00"), #DELETE THIS WHEN DONE
         timestamp < ymd_hms("2024-03-17 16:38:00") |
           timestamp > ymd_hms("2024-03-17 17:02:00")) %>%
  mutate(timestamp_posix = as.POSIXct(timestamp),
         timestamp_posix_plus = timestamp_posix + (25 * 60),
         timestamp = if_else(timestamp < ymd_hms("2024-03-17 17:03:00"), timestamp_posix_plus, timestamp_posix)) %>%
  #remove the last 2 columns
  select(!last_col(offset=1):last_col())

#get average temperature for entire dataset
temp_avg <- mean(temp_simple$avg_temp) #17.9999
temp_sd <- sd(temp_simple$avg_temp) #0.006 

image <- ggplot(temp_simple, aes(x=timestamp)) +
  #geom_line(aes(y=avg_temp, color=heater_status), size=1) +
  geom_point(aes(y=avg_temp), color="#D55E00", size=1) +
  geom_line(aes(y=temp_set), color="#000000", size=1.5, alpha=0.7) +
  scale_x_datetime(date_labels = "%b %d - %H:%M",
                   breaks = scales::date_breaks("6 hours"),
                   expand=c(0,0),
                   limits = c(min(temp_simple$timestamp),
                              max(temp_simple$timestamp))) +
  scale_y_continuous(limits = c(17.8, 20.2), breaks = seq(18, 20, by = 0.5)) +
  labs(y = expression(paste("Temperature (", degree, "C)")),
       x = "Date and time") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=14),
        axis.text.y=element_text(size=14),
        axis.title.x=element_text(size=18),
        axis.title.y=element_text(size=18),
        legend.position="top")

#ggsave(here("experiment", "figures", "methodsPub", "simple.png"), width=12, height=6)
#ggsave(file = here("experiment", "figures", "methodsPub","simple.svg"), plot = image, width=12, height=6)
```

## Check response time/stabilization time of system
```{r}

# Define stabilization criteria
tolerance <- 0.1  # Define your tolerance range (±0.5 degrees)
stabilization_times <- data.frame()

for (i in 2:nrow(temp_simple)) {
  if (temp_simple$temp_set[i] != temp_simple$temp_set[i - 1]) {
    change_time <- temp_simple$timestamp[i]
    new_set_point <- temp_simple$temp_set[i]
    
    # Filter for the rows after the set point change
    stabilization_period <- temp_simple %>%
      filter(timestamp > change_time) %>%
      mutate(stabilized = abs(avg_temp - new_set_point) <= tolerance)
    
    # Check if stabilization occurs
    stabilization_row <- stabilization_period %>%
      filter(stabilized) %>%
      slice(1)  # Get the first row where stabilization occurs
    
    if (nrow(stabilization_row) > 0) {
      stabilization_time <- stabilization_row$timestamp - change_time
      stabilization_times <- rbind(stabilization_times, data.frame(
        timestamp = change_time,
        temp_set = new_set_point,
        stabilization_time = as.numeric(stabilization_time, units = "secs")
      ))
    } else {
      stabilization_times <- rbind(stabilization_times, data.frame(
        timestamp = change_time,
        temp_set = new_set_point,
        stabilization_time = NA
      ))
    }
  }
}

# View results
print(stabilization_times)

temp_check <- temp_simple %>%
  filter(temp_set == 20) %>%
  slice(1)

temp_check2 <- temp_simple %>%
  filter(avg_temp >= 19.9) %>%
  slice(1)

2024-03-18 07:52:00
2024-03-18 08:30:00

2024-03-18 10:02:00



```


## Here is the "real" simple MHW data! (adjusting for differences in temp profile start times)
```{r}
temp_down <- temp_all %>%
  mutate(timestamp = ymd_hms(timestamp)) %>%
  filter(timestamp > ymd_hms("2024-03-17 13:00:00"),
         timestamp < ymd_hms("2024-03-18 15:54:00") | timestamp > ymd_hms("2024-03-18 15:57:00")) %>%
  mutate(timestamp_posix = as.POSIXct(timestamp),
         timestamp_posix_minus = timestamp_posix - ((3*60)) + 16,
         timestamp = if_else(timestamp > ymd_hms("2024-03-18 15:53:44"), timestamp_posix_minus, timestamp_posix),
         timestamp = floor_date(timestamp, "1 minute")) %>%
  select(!last_col(offset=1):last_col()) %>%
  pivot_longer(cols=4:6, names_to = "sensor", values_to = "temperature") %>%
  group_by(timestamp, temp_set) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ungroup() %>%
  #remove datapoints between 2024-03-17 16:32:00 and 2024-03-17 16:57:00
  filter(timestamp != ymd_hms("2024-03-17 16:32:00"),
         timestamp > ymd_hms("2024-03-16 23:00:00"), #remove Blob data
         timestamp > ymd_hms("2024-03-17 08:00:00"), #DELETE THIS WHEN DONE
         timestamp < ymd_hms("2024-03-17 16:38:00") |
           timestamp > ymd_hms("2024-03-17 17:02:00")) %>%
  mutate(timestamp_posix = as.POSIXct(timestamp),
         timestamp_posix_plus = timestamp_posix + (25 * 60),
         timestamp = if_else(timestamp < ymd_hms("2024-03-17 17:03:00"), timestamp_posix_plus, timestamp_posix)) %>%
  #remove the last 2 columns
  select(!last_col(offset=1):last_col())

image <- ggplot(temp_down, aes(x=timestamp)) +
  #geom_line(aes(y=avg_temp, color=heater_status), size=1) +
  geom_point(aes(y=avg_temp), color="#D55E00", size=1) +
  geom_line(aes(y=temp_set), color="#000000", size=1.5, alpha=0.7) +
  scale_x_datetime(date_labels = "%b %d - %H:%M",
                   breaks = scales::date_breaks("6 hours"),
                   expand=c(0,0),
                   limits = c(min(temp_down$timestamp),
                              max(temp_down$timestamp))) +
  scale_y_continuous(limits = c(17.8, 20.2), 
                     breaks = seq(18, 20, by = 0.5)) +
  labs(y = expression(paste("Temperature (", degree, "C)")),
       x = "Date and time") +
       #title = "A") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=18),
        axis.text.y=element_text(size=18),
        axis.title.x=element_text(size=20),
        axis.title.y=element_text(size=20),
        legend.position="top")
        #plot.title = element_text(size=40))

#ggsave(here("experiment", "figures", "methodsPub", "simple.png"), width=12, height=6)
ggsave(file = here("experiment", "figures", "methodsPub","Fig3A.svg"), plot = image, width=12, height=6)
```

## Plot difference between "expected" and "observed"
```{r}
temp_diff_simple <- temp_down %>%
  mutate(diff = round(temp_set - avg_temp, 3))

mean(temp_diff_simple$diff)
sd(temp_diff_simple$diff)
max(temp_diff_simple$diff)
min(temp_diff_simple$diff)
max(temp_diff_simple$diff)+abs(min(temp_diff_simple$diff))

ggplot(temp_diff_simple, aes(x=timestamp)) +
  geom_point(aes(y=diff), color="#000000", size=1) +
  scale_x_datetime(date_labels = "%b %d - %H:%M",
                   breaks = scales::date_breaks("6 hours"),
                   expand=c(0,0),
                   limits = c(min(temp_diff_simple$timestamp),
                              max(temp_diff_simple$timestamp))) +
  #scale_y_continuous(limits = c(-0.05, 0.05), breaks = seq(-0.05, 0.05, by = 0.02)) +
  labs(y = expression(paste("Temperature difference (", degree, "C)")),
       x = "Date and time") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=14),
        axis.text.y=element_text(size=14),
        axis.title.x=element_text(size=18),
        axis.title.y=element_text(size=18),
        legend.position="top")

#ggsave(here("experiment", "figures", "methodsPub", "simpleDiff.png"), width=12, height=6)
```

# Import HOBO data
```{r}
hobo_data <- read_csv(here("experiment", "data", "newAmeliaHobo 2024-03-19 10_04_08 PDT (Data PDT).csv")) %>% #each file will be read in, specify which columns you need read in to avoid any error
  select(2:3) %>%
  rename(timestamp = 1,
         temperature = 2) %>%
  mutate(timestamp = mdy_hms(timestamp),
         timestamp = floor_date(timestamp, "1 minute")) %>%
  group_by(timestamp) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ungroup() %>%
  mutate(avg_temp_sub = avg_temp - 0.22,
         avg_temp = (avg_temp - min(avg_temp)) / (max(avg_temp) - min(avg_temp)) * (20 - 18) + 18) # calibrate the data so the maximum temp should be 20 and the minimum temp should be 18

hobo_data_clean <- hobo_data %>%
  rename(avg_temp_hobo = 2)

#merge hobo_data_clean with temp_down
temp_compare <- temp_down %>%
  left_join(hobo_data_clean, by="timestamp") %>%
  #remove NA rows
  filter(!is.na(avg_temp_hobo))

#get the difference between the two data
temp_compare <- temp_compare %>%
  mutate(diff = round(avg_temp - avg_temp_sub, 3))

mean(temp_compare$diff)
sd(temp_compare$diff)

#calculate the number of hours between start and end of temp_compare
hours <- as.numeric(difftime(max(temp_compare$timestamp), min(temp_compare$timestamp), units="hours"))

#plot the data
image <- ggplot(temp_compare, aes(x=timestamp)) +
  geom_point(aes(y=diff), color="black", size=1) +
  #geom_point(aes(y=avg_temp_hobo), color="red", size=1) +
  #geom_point(aes(y=avg_temp_sub), color="green", size=1) +
  scale_x_datetime(date_labels = "%b %d - %H:%M",
                   breaks = scales::date_breaks("6 hours"),
                   expand=c(0,0),
                   limits = c(min(hobo_data$timestamp),
                              max(hobo_data$timestamp))) +
  scale_y_continuous(limits = c(-0.045, 0.02), breaks = seq(-0.04, 0.02, by = 0.02)) +
  labs(y = expression(paste("Temperature difference (", degree, "C)")),
       x = "Date and time") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=14),
        axis.text.y=element_text(size=14),
        axis.title.x=element_text(size=18),
        axis.title.y=element_text(size=18),
        legend.position="top")
  
#ggsave(here("experiment", "figures", "methodsPub", "hoboTempDiff.png"), width=12, height=6)
#ggsave(file = here("experiment", "figures", "methodsPub","hoboTempDiff.svg"), plot = image, width=12, height=6)

```

