---
title: "Looking at MHW trends in the SBC LTER to inform the MHWsim experiment"
author: "Amelia Ritger"
date: "2023-07-31"
output: html_document
---

#Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(scales) #date_format for x axis plot
library(heatwaveR)
library(ggpubr) #combine multiple ggplots
library(chron) #convert decimal time
library(EDIutils) #to download LTER data
library(xml2) #to open LTER xlsx data

source("read_sbc_data.r")
dt <- read.csv(here("experiment", "data", "vw_timeseries_entities.csv") ,stringsAsFactors = FALSE)
dsname <- "Moored CTD and ADCP: NAP - Moored CTD and ADCP at Naples Reef NAP"
```

Moored CTD and ADCP: ARQ - Moored CTD and ADCP at Arroyo Quemado Reef ARQ
Moored CTD and ADCP: CAR - Moored CTD and ADCP at Carpinteria Reef CAR
Moored CTD and ADCP: MKO - Moored CTD and ADCP at Mohawk Reef MKO
Moored CTD and ADCP: NAP - Moored CTD and ADCP at Naples Reef NAP

#Load data
```{r}
raw_temp <-  read_sbc_data(dsname) %>%
  select(year, month, day,  decimal_time, Temp_adcp, Temp_top, Temp_mid, Temp_bot,  ADCP_depth) %>%
  clean_names()
```

#Clean up the data
```{r}
temp_date <- raw_temp %>%
  unite("date", "year", "month", "day", sep=" ", remove=FALSE) %>%
  mutate(date=ymd(date)) %>% #apply lubridate to date
  arrange(date, decimal_time)# %>%
  #mutate(time=dminutes(decimal_time))

temp_clean <- temp_date %>%
  mutate(source = "adcp",
         source = ifelse(temp_adcp>999,"bottom",source),
         temp=ifelse(temp_adcp>999,temp_bot,temp_adcp),
         source = ifelse(temp>999,"middle",source),
         temp=ifelse(temp>999,temp_mid,temp),
         source = ifelse(temp>999,"top",source),
         temp=ifelse(temp>99,temp_top,temp)) %>%
  filter(temp<999)
```

## Get monthly average temperature for each month across years
```{r}
decade_time <- "2013-08-01"

mhw_sum <- temp_clean %>%
  group_by(year, month) %>%
  summarize(temp=mean(temp)) %>%
  filter(month > 7 | month < 3) %>%
  mutate(year=as.factor(year),
         month=as.factor(month),
         month=fct_relevel(month,c("8","9","10","11","12","1","2")))

ggplot(mhw_sum, aes(x=month, y=temp, group=year)) +
  geom_point(aes(color=year)) +
  geom_line(aes(color=year)) +
  labs(title="Average temperature for each month")

#ggsave(here("figures", "avg_temp_mko.png"))
```

## Look at extremes - highs and lows - over time
```{r}
max_temp <- max(temp_clean$temp)
min_temp <- min(temp_clean$temp)

temp_top <- temp_clean %>%
  filter(temp>20)

temp_high <- temp_clean %>%
  group_by(year, month) %>%
  summarize(temp=max(temp)) %>%
  filter(month > 7 | month < 3) %>%
  mutate(year=as.factor(year),
         month=as.factor(month),
         month=fct_relevel(month,c("8","9","10","11","12","1","2")))

ggplot(temp_high, aes(x=month, y=temp, group=year)) +
  geom_point(aes(color=year)) +
  geom_line(aes(color=year)) +
  labs(title="Hottest temperature for each month")

#temp_clean <- nap_clean
#ggsave(here("figures", "max_temp_mko.png"))
```

## Assess MHW
```{r}
# Reformat to run heatwaveR
temp_heatwaver <- temp_clean %>%
  group_by(date) %>%
  summarize(temp=mean(temp)) %>%
  mutate(t=as.Date(date)) %>%
  select(t, temp)

#heatwave_list <- list(arq_heatwaver, mko_heatwaver, carp_heatwaver, nap_heatwaver)
#mhw_daily <- heatwave_list %>% #make sure you re-run code above naming mhw_event appropriately
#  reduce(full_join) %>%
#  group_by(t) %>%
#  summarize(temp=mean(temp))

#identify start and end dates of df
start_date <- as.character((head(temp_heatwaver$t,1)))
end_date <- as.character((tail(temp_heatwaver$t,1)))

# Detect the events in a time series
ts <- heatwaveR::ts2clm(temp_heatwaver, x=t, y=temp, climatologyPeriod = c(start_date, end_date))
mhw <- heatwaveR::detect_event(ts)
```

## Identify MHW events
```{r}
mhw_event <- mhw$event %>%
   dplyr::ungroup() %>%
   dplyr::select(event_no, duration, date_start, date_peak, date_end, intensity_max, intensity_max_relThresh, intensity_max_abs, intensity_mean, intensity_cumulative, rate_onset, rate_decline) %>%
  filter(date_start > ymd(decade_time)) %>% #Only look at MHWs in the last decade
  mutate(month_start=month(date_start),
         month_end=month(date_end)) %>%
  filter(between(month_start,1,2) | between(month_start,8,12)) %>% #CHECK MANUALLY BEFORE FILTERING, make sure there aren't long duration MHWs that started in June/July and extended well into August and beyond
  dplyr::arrange(-duration) %>%
  select(1:12) %>%
  filter(date_start < ymd("2017-01-01") & date_end > ymd("2014-06-01"))

nap_event <- mhw_event
#write_csv(mhw_event, here("tables", "mohawk_mhw_all.csv"))
```

Combine NAP and MKO to summarize MHW patterns
```{r}
event_list <- list(arq_event, car_event, mko_event, nap_event)
mhw_event <- event_list %>% #make sure you re-run code above naming mhw_event appropriately
  reduce(full_join)

ggplot(data = mhw_event, aes(x=duration)) +
  geom_histogram()

duration_95 <- unname(quantile(mhw_event$duration, c(.95)))
duration_90 <- unname(quantile(mhw_event$duration, c(.90)))
duration_mean <- mean(mhw_event$duration)

duration_sd <- sd(mhw_event$duration)
duration_se <-  sd(mhw_event$duration)/sqrt(length((mhw_event$duration)))
duration_min <- min(mhw_event$duration)
duration_max <- max(mhw_event$duration)
duration_quantile <- quantile(mhw_event$duration, prob=c(.25,.5,.75), type=1)

intensity_95 <- unname(quantile(mhw_event$intensity_max, c(.95)))
intensity_mean <- mean(mhw_event$intensity_max)

intensity2_95 <- unname(quantile(mhw_event$intensity_mean, c(.95)))
intensity2_mean <- mean(mhw_event$intensity_mean)

onset_95 <- unname(quantile(mhw_event$rate_onset, c(.95)))
onset_mean <- mean(mhw_event$rate_onset)

decline_95 <- unname(quantile(mhw_event$rate_decline, c(.95)))
decline_mean <- mean(mhw_event$rate_decline)

mhw_summary <- data.frame(Metric=c("Duration", "Max intensity", "Mean intensity", "Onset rate", "Decline rate"),
                          Percentile_95=c(duration_95,intensity_95,intensity2_95,onset_95,decline_95),
                          Mean=c(duration_mean,intensity_mean,intensity2_mean,onset_mean,decline_mean))

#write_csv(mhw_summary, here("tables", "arq-carp-mko-nap_mhw_summary.csv"))
```

Plot histogram(s)
```{r}
ggplot(mhw_event, aes(x=duration)) +
  geom_histogram(bins=20) +
  labs(title="MHW duration since 2013")

ggplot(mhw_event, aes(x=intensity_max)) +
  geom_histogram(bins=15) +
  labs(title="MHW maximum intensity since 2013")

#ggsave(here("figures", "mhw_hist_int_mko.png"), width = 20, height = 7)
```

Get 95% quantile and average values of duration, intensity
```{r}
duration_95 <- unname(quantile(mhw_event$duration, c(.95)))
duration_mean <- mean(mhw_event$duration)

intensity_95 <- unname(quantile(mhw_event$intensity_max, c(.95)))
intensity_mean <- mean(mhw_event$intensity_max)

intensity2_95 <- unname(quantile(mhw_event$intensity_mean, c(.95)))
intensity2_mean <- mean(mhw_event$intensity_mean)

mhw_summary <- data.frame(Metric=c("Duration", "Max intensity", "Mean intensity"),
                          Percentile_95=c(duration_95,intensity_95,intensity2_95),
                          Mean=c(duration_mean,intensity_mean,intensity2_mean))

#write_csv(mhw_summary, here("tables", "mohawk_mhw_summary.csv"))
```

# Plot it all up
```{r}
#identify dates of interest
climatology <- mhw$climatology %>%
  arrange(t)

mhw_start <- which(climatology$t == decade_time)[1]
mhw_end <- which(climatology$t == end_date)[1]

#subset for dates of interest
mhw_subset <- mhw$climatology %>%
  slice(mhw_start:mhw_end) %>%
  mutate(month=as.factor(month(t))) %>%
  mutate(thresh2=seas+5,
         thresh3=seas+8,
         extreme=temp+4,
         chill=temp-4) %>%
  filter(t>=ymd("2015-09-01") & t<ymd("2015-10-01"))
 
#mhw_profile_mko <- mhw$climatology %>%
# select(t, temp) %>%
# filter(t > ymd("2020-01-01") & t < ymd("2021-01-01"))

#mhw_profile_mko <- mhw_profile_mko %>%
# rename(temp_mko=temp)

#mhw_profile_all <- full_join(mhw_profile_mko, mhw_profile_nap) %>%
#  mutate(temp_final = rowMeans(select(.,starts_with("temp")), na.rm = TRUE)) %>%
#  select(t, temp_final) %>%
#  rename(temp=temp_final)

#write_csv(mhw_profile_all, here("data", "mhw_profile.csv"))

#Create a df of dates to shade the time period of interest in ggplot
dateRanges <- data.frame(
  start = seq(ymd("1900-08-01"), ymd("2100-08-01"), "1 year"),
  end = seq(ymd("1901-02-01"), ymd("2101-02-01"), "1 year")
)

#plot clean ribbon
ggplot(mhw_subset, aes(x = t, ymax = pmax(thresh, temp), ymin = thresh)) +
  geom_rect(data = dateRanges, aes(xmin = start , xmax = end, ymin = -Inf, ymax = Inf),
            inherit.aes=FALSE, alpha = 0.4, fill = c("lightgrey"))+
  geom_line(aes(y = seas, colour = "seas"), size = 0.8) +
  geom_ribbon(fill = "red") +
  geom_line(aes(y = thresh, colour = "thresh"), size = 0.8) +
  #geom_line(aes(y = chill, colour = "thresh2"), size = 0.8) +
  #geom_line(aes(y = extreme, colour = "thresh3"), size = 0.8) +
  geom_line(aes(y = temp, colour = "temp"), size=0.8, alpha=0.3) +
  #geom_line(aes(y=25, colour="extreme"), linetype="dashed", size=1, alpha=0.4) +
  scale_colour_manual(name = "Line Colour",
                      values = c("thresh" =  "chartreuse4",
                                 "thresh2" =  "orange2",
                                 "thresh3" =  "red4"),
                      labels= c("Climatology","Temperature","Threshold")) +
  scale_x_date(date_labels = "%b %d",
               breaks = scales::date_breaks("1 month"),
               expand=c(0,0),
               limits = c(min(mhw_subset$t), 
                         max = max(mhw_subset$t))) +
  scale_y_continuous(limits = c(14, 22), breaks = seq(14, 22, by = 2)) +
  guides(colour = guide_legend(override.aes = list(fill = NA))) +
  labs(y = expression(paste("Temperature (", degree, "C)")),
       x = "Date") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=22),
        axis.title.y=element_text(size=22),
        legend.text=element_text(size=20),
        legend.position="top",
        legend.title = element_blank())

#ggsave(here("figures", "MHWsim_simple.png"), width = 20, height = 7)
```

## Plot diurnal cycle changes!
```{r}
library(chron)

mhw_subset_diurnal <- temp_clean %>%
  filter(date>=ymd("2015-09-01") & date<ymd("2015-10-01")) %>%
  mutate(t = date,
         time=format(times(decimal_time))) %>%
  unite(date_time, c(t, time), sep = " ", remove = TRUE) %>% # combine the date and time columns
  mutate(date_time=ymd_hms(date_time)) %>%
  select(date_time, temp) %>%
  mutate(extreme=temp+4,
         chill=temp-4)

ggplot(mhw_subset_diurnal, aes(x = date_time)) +
  geom_line(aes(y = temp, colour = "temp"), size = 0.8) +
  geom_line(aes(y = extreme, colour = "extreme"), size = 0.8) +
  geom_line(aes(y = chill, colour = "chill"), size = 0.8) +
  geom_line(aes(y=25, colour="black"), linetype="dashed", size=1) +
  scale_colour_manual(name = "Line Colour",
                      values = c("chill" =  "dodgerblue4",
                                 "temp" = "red", 
                                 "extreme" =  "darkorange1"),
                      labels= c("Historical climate","Historical HW", "Extreme HW")) +
  scale_x_datetime(date_labels = "%b %d",
               breaks = scales::date_breaks("1 week"),
               expand=c(0,0),
               limits = c(min(mhw_subset_diurnal$date_time), 
                         max = max(mhw_subset_diurnal$date_time))) +
  scale_y_continuous(limits = c(13, 27), breaks = seq(13, 27, by = 2)) +
  guides(colour = guide_legend(override.aes = list(fill = NA))) +
  labs(y = expression(paste("Temperature (", degree, "C)")),
       x = "Date") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=18),
        axis.title.y=element_text(size=18),
        legend.text=element_text(size=20),
        legend.position="top",
        legend.title = element_blank())

#ggsave(here("figures", "MHWsim_daily.png"), width = 20, height = 7)
```

## Combine the two plots to visualize
```{r}
mhw_subset_avg <- mhw_subset_diurnal %>%
  mutate(t=date(date_time)) #%>%
  #group_by(date) %>%
  #mutate(avg_temp = mean(temp),
  #       avg_xtr = mean(extreme),
  #       avg_chill=mean(chill)) %>%
  #distinct(date, .keep_all=TRUE) %>%
  #mutate(t=date)

mhw_subset_comb <- left_join(mhw_subset_avg, mhw_subset, by="t") %>%
  group_by(t) %>%
  mutate_at(vars(temp.y,chill.y,extreme.y), funs(replace(., duplicated(.), NA))) #don't fill with values, but with NAs (for ggplot)

ggplot(mhw_subset_comb, aes(x = date_time)) +
      geom_line(data=mhw_subset_comb[!is.na(mhw_subset_comb$seas),], aes(y = seas), colour="black", size = 0.8, linetype="dotted") +
  geom_line(aes(y = temp.x, colour = "temp"), size = 0.8) +
  geom_line(aes(y = extreme.x, colour = "extreme"), size = 0.8) +
  geom_line(aes(y = chill.x, colour = "chill"), size = 0.8) +
  geom_line(data=mhw_subset_comb[!is.na(mhw_subset_comb$temp.y),], aes(y = temp.y, colour = "temp2"), size = 0.8, alpha=0.7, linetype="dashed") +
  geom_line(data=mhw_subset_comb[!is.na(mhw_subset_comb$chill.y),], aes(y = chill.y, colour = "chill2"), size = 0.8, alpha=0.7, linetype="dashed") +
  geom_line(data=mhw_subset_comb[!is.na(mhw_subset_comb$extreme.y),], aes(y = extreme.y, colour = "extreme2"), size = 0.8, alpha=0.7, linetype="dashed") +
  scale_colour_manual(name = "Line Colour",
                      values = c("chill2" =  "blue",
                                 "temp2" = "red", 
                                 "extreme2" =  "purple",
                                 "chill" =  "dodgerblue4",
                                 "temp" = "darkred", 
                                 "extreme" =  "darkorchid4"),
                      labels= c("Historical climate", "Historical HW trend", "Extreme HW", "Extreme HW trend", "Historical HW", "Historical HW trend")) +
  scale_x_datetime(date_labels = "%b %d",
               breaks = scales::date_breaks("1 week"),
               expand=c(0,0),
               limits = c(min(mhw_subset_avg$date_time), 
                         max = max(mhw_subset_avg$date_time))) +
  scale_y_continuous(limits = c(13, 27), breaks = seq(13, 27, by = 2)) +
  guides(colour = guide_legend(override.aes = list(fill = NA))) +
  labs(y = expression(paste("Temperature (", degree, "C)")),
       x = "Date") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, vjust = 1, hjust=1, size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=22),
        axis.title.y=element_text(size=22),
        legend.text=element_text(size=20),
        legend.position="top",
        legend.title = element_blank())

#ggsave(here("figures", "MHWsim_all.png"), width = 20, height = 7)
```

#Decide what value of C for "chilled" treatment
```{r}
mhw_comp <- mhw_subset_comb %>%
  select(date_time, t, temp.x, extreme.x, chill.x, temp.y, extreme.y, chill.y, seas) %>%
  group_by(t) %>%
  summarize(mean_day_b = mean(temp.x),
            mhw_b = mean(temp.y, na.rm = TRUE),
            mean_day_x = mean(extreme.x),
            mhw_x = mean(extreme.y, na.rm = TRUE),
            mean_day_c = mean(chill.x),
            mhw_c = mean(chill.y, na.rm = TRUE),
            seas = mean(seas)) %>%
  mutate(diff = mhw_c-seas)# %>%
  #summarize(mean_diff = mean(diff)) #comment this out, used to compare 4 vs 4.5C with climatology

#plot difference between climatology and -XC
ggplot(mhw_comp, aes(x=t)) +
  geom_line(aes(y = diff), colour = "blue", size = 0.8)

#Compare -XC with Climatology
ggplot(mhw_comp, aes(x=t)) +
  geom_line(aes(y = mean_day_c), colour = "blue", size = 0.8) +
  geom_smooth(aes(y=mhw_c), method = "lm", se = F, colour = "red") +
  geom_line(aes(y = mhw_c), colour = "red", size = 0.8) +
  geom_line(aes(y = seas), colour = "orange", size = 0.8)
  
#ggsave(here("figures", "MHWdiff_4.5.png"), width = 20, height = 7)
```

#Create dfs for MHW temperatures and climatology values
```{r}
all_clean <- list(arq_clean, carp_clean, mko_clean, nap_clean)
all_data <- all_clean %>% #make sure you re-run code above naming mhw_event appropriately
  reduce(full_join)

mhw_profile <- temp_clean %>% #or use "all_data"
  filter(date>=ymd("2015-03-06") & date<ymd("2015-03-15")) %>% #date>=ymd("2015-08-01") & date<ymd("2016-04-01")
  mutate(t = date,
         time=format(times(decimal_time))) %>%
  unite(date_time, c(t, time), sep = " ", remove = TRUE) %>% # combine the date and time columns
  mutate(date_time=ymd_hms(date_time)) %>%
  group_by(date_time) %>%
  summarize(temp=mean(temp)) %>%
  select(date_time, temp) %>%
  mutate(t=date(date_time))

# save mhw_profile as .csv file using read_csv
write_csv(mhw_profile, here("experiment", "data", "2015mhw_profile.csv"))

#mhw_profile <- nap_clean %>%
#    filter(date>=ymd("2015-08-15") & date<ymd("2016-03-01")) %>%
#    mutate(t = date,
#           time=format(times(decimal_time))) %>%
#    unite(date_time, c(t, time), sep = " ", remove = TRUE) %>% # combine the date and time columns
#    mutate(date_time=ymd_hms(date_time))
  
ggplot(mhw_profile, aes(x=date_time)) +
  geom_point(aes(y=temp), color="black")
  #geom_point(aes(y=chill), color="blue")
```

# Extract the seasonal climatology data
```{r}
mhw_seas <- mhw$climatology %>%
  slice(mhw_start:mhw_end) %>%
  mutate(month=as.factor(month(t))) %>%
  filter(t>=ymd("2015-08-01") & t<ymd("2016-04-01")) %>%
  select(t, seas)
```

# Transform temperature data based on climatology data, plot and generate .csv file
```{r}
joined_df <- left_join(mhw_profile, mhw_seas, by = "t") %>%
  mutate(seas_thresh = seas+4,
         seas_extreme = seas+8) %>%
  group_by(t) %>%
  mutate(mean_temp = mean(temp), #get the mean temp for the day (because seasonal climatology value is daily)
         adjusted_temp=temp-(mean_temp-seas), #get difference between seasonal value for day and measured average value for day
         adjusted_thresh=temp-(mean_temp-seas_thresh), #get difference between seasonal value+4 and measured value
         adjusted_extreme=temp-(mean_temp-seas_extreme)) %>% #get difference between seasonal value+8 and measured value
  ungroup()

mhw_daily <- mhw_profile %>%
  group_by(t) %>%
  mutate(mean_daily = mean(temp),
         sd_daily = sd(temp)) %>%
  ungroup() %>%
  select(t, mean_daily, sd_daily) %>%
  distinct()

daily_new <- left_join(mhw_profile, mhw_daily, by = "t")

new_df <- left_join(daily_new, mhw_seas, by = "t") %>%
  mutate(seas_thresh = seas+4,
         seas_extreme = seas+8) %>%
  mutate(zscore = (temp - mean_daily)/sd_daily,
         adjusted_temp = zscore * sd_daily + seas,
         adjusted_thresh = adjusted_temp+4,
         adjusted_extreme = adjusted_temp+8)

#Plot it all, compare transformed "Blob" data with real Blob data
ggplot(new_df, aes(x = date_time)) +
  geom_line(aes(y = temp, colour="temp"), size = 0.8) +
  geom_line(aes(y = adjusted_thresh, colour="adjusted_thresh"),size = 0.8) +
  geom_line(aes(y = adjusted_extreme, colour="adjusted_extreme"),size = 0.8) +
  #geom_smooth(aes(y=temp), method = "lm", se = F, colour = "blue4") +
  #geom_smooth(aes(y=adjusted_thresh), method = "lm", se = F, colour = "purple4") +
  #geom_smooth(aes(y=adjusted_extreme), method = "lm", se = F, colour = "blue") +
  geom_line(aes(y = adjusted_temp, colour="adjusted_temp"), size = 0.8) +
  geom_line(data = new_df[!is.na(new_df$seas), ],aes(y = seas, colour="seas"),size = 0.8) +
  scale_colour_manual(name = "Line Colour",
                      values = c("temp" =  "lightblue",
                                 "adjusted_thresh" = "red", 
                                 "adjusted_extreme" =  "darkred",
                                 "adjusted_temp" =  "purple",
                                 "seas" = "dodgerblue4"),
                      labels= c("Transformed extreme", "Transformed chill", "Transformed severe", "Climatology", "The Blob")) +
  scale_x_datetime(date_labels = "%b %d",
                   breaks = scales::date_breaks("1 week"),
                   expand = c(0, 0),
                   limits = c(min(new_df$date_time),max = max(new_df$date_time))) +
  scale_y_continuous(limits = c(13, 27), breaks = seq(13, 27, by = 2)) +
  guides(colour = guide_legend(override.aes = list(fill = NA))) +
  labs(y = expression(paste("Temperature (", degree, "C)")),x = "Date") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1,hjust = 1,size = 20),
        axis.text.y = element_text(size = 20),
        axis.title.x = element_text(size = 22),
        axis.title.y = element_text(size = 22),
        legend.text = element_text(size = 20),
        legend.position = "top",
        legend.title = element_blank())

#ggsave(here("figures", "mhw_profile.png"), width = 20, height = 7)

mhw_profile_final <- new_df %>%
  select(1,"temp"=11,"extreme"=12,"chill"=10,3) %>% #reformat for use with RPi
   mutate_at(2:4, round, 2)

#write_csv(mhw_profile_final, here("data", "mhw_profile_final.csv"))
```

